# Spaces

For spatial data, the location of observations are characterised
by coordinates, and coordinates are defined in a certain space.
Different spaces are used for this, and the most important
difference is whether coordinates are defined over a 2-dimensional
or 3-dimensional space referenced to orthogonal axes (Cartesian
coordinates), or using distance and direction angles (polar
coordinates, spherical coordinates).

```{r polar, fig.cap="Two-dimensional polar (red) and Cartesian (blue) coordinates", eval=TRUE, fig=TRUE, echo=FALSE}
par(mar = rep(0,4))
plot(3, 4, xlim = c(-6,6), ylim = c(-6,6), asp = 1)
axis(1, pos = 0, at = 0:6)
axis(2, pos = 0, at = -6:6)
xd = seq(-5, 5, by = .1)
lines(xd, sqrt(25 - xd^2), col = 'grey')
lines(xd, -sqrt(25 - xd^2), col = 'grey')
arrows(0, 0, 3, 4, col = 'red', length = .15, angle = 20)
text(1.5, 2.7, label = "r", col = 'red')
xd = seq(3/5, 1, by = .1)
lines(xd, sqrt(1 - xd^2), col = 'red')
text(1.2, 0.5, label = parse(text = "phi"), col = 'red')
lines(c(3,3), c(0,4), lty = 2, col = 'blue')
lines(c(0,3), c(4,4), lty = 2, col = 'blue')
text(3.3, 0.3, label = "x", col = 'blue')
text(0.3, 4.3, label = "y", col = 'blue')
```

## Cartesian and polar coordinates
Figure \@ref(fig:polar) shows both polar and Cartesian coordinates
for a standard two-dimensional situation. In Cartesian coordinates,
the point shown is $(x,y) = (3,4)$, for polar coordinates it is
$(r,\phi) = (5, \mbox{arctan}(4/3))$, where $\mbox{arctan}(4/3)$ is
approximately $0.93$ radians, or $53^{\circ}$. Note that $x$, $y$
and $r$ all have length units, where $\phi$ is an angle (a unitless
length/length ratio). Converting back and forth between Cartesian
and polar coordinates is trivial, 
$$x = r~\mbox{cos} \phi$$
$$y = r~\mbox{sin} \phi$$
$$\phi = \mbox{arctan}(y/x)$$
$$r = \sqrt{x^2 + y^2}$$
but requires care with using the
right quadrant for $\mbox{arctan}$; the `atan2` function is helpful
here.

## Spherical and ellipsoidal coordinates

In three dimensions, where Cartesian coordinates are expressed as
$(x,y,z)$, spherical coordinates are the three-dimensional equivalent
of polar coordinates and can be expressed as $(r,\lambda,\phi)$, where

* $r$ is the radius of the sphere,
* $\lambda$ is the longitude, measured in the $(x,y)$ plane counter-clockwise from positive $x$, and
* $\phi$ is the latitude, the angle between the vector and the $(x,y)$ plane.

$\lambda$ typically varies between $-180^{\circ}$ and $180^{\circ}$
(or alternatively from $0^{\circ}$ to $360^{\circ}$), $\phi$ from
$-90^{\circ}$ to $90^{\circ}$.  When we are only interested in points
_on_ a sphere with given radius, we can drop $r$ and $(\lambda,\phi)$
suffice to identify any point.

It should be noted that this is just _a_ definition, one could for
instance also choose to measure polar angle, i.e.~the angle between
the vector and $z$, instead of latitude. There is also a long
tradition of specifying points as $(\phi,\lambda)$ but throughout
this book we will stick to longitude-latitude, $(\lambda,\phi)$.

For points on an ellipse, there are two ways in which angle can be
expressed (figure \@ref(fig:ellipse)): measured from the center of
the ellipse ($\psi$), or measured perpendicular to the tangent on
the ellipse at the target point ($\phi$).

```{r ellipse, fig.cap="Angles on an ellipse: geodetic (blue) and geocentric (red) latitude", eval=TRUE, fig=TRUE, echo=FALSE}
par(mar = rep(0,4))
x = 4
y = 5/8 * sqrt(48)
plot(x, y, xlim = c(-6,6), ylim = c(-8,8), asp = 1)
axis(1, pos = 0, at = 0:9)
axis(2, pos = 0, at = -5:5)
xd = seq(-8, 8, by = .1)
lines(xd, 5/8 * sqrt(64 - xd^2), col = 'grey')
lines(xd, 5/8 * -sqrt(64 - xd^2), col = 'grey')
arrows(0, 0, x, y, col = 'red', length = .15, angle = 20)
b = (x * 25) / (-y * 64)
a = y - x * b
abline(a, b, col = 'grey')
b = -1/b
x0 = x - y / b
arrows(x0, 0, x, y, col = 'blue', length = .15, angle = 20)
text(1.2, 0.5, label = parse(text = "psi"), col = 'red')
text(3, 0.5, label = parse(text = "phi"), col = 'blue')
```

The most common parametric model for the Earth is a spheroid, also
called an ellipsoid of revolution, which is an ellipsoid with two
equal semi-axes. In effect, this is a flattened sphere: the distance
between the poles is (slightly: about 0.33%) smaller than the distance between
two opposite points on the equator. Under this model, longitude is
always measured along a circle, but latitude along an ellispe. If
we think of figure \@ref(fig:ellipse) as a cross section of the
Earth passing through the poles, the latitude measure $\phi$ is
the one used when no further specification is given; it is also
called _geodetic latitude_. The latitude measure $\psi$ is called
the _geocentric latitude_.

In addition to longitude and latitude we can add _altitude_ to define
points that are not on the spheroid, and obtain a three dimensional
space again. When defining altitude, we need to choose where zero
altitude is (on the ellipsoid, or on the mean sea level?), as well
as which direction is positive, and which direction is "straight
up": perpendicular to the spheroid surface, or in the direction
opposite to gravity? All these choices may matter, depending on
the application area.

## Distances

Distances between two points $p_i$ and $p_j$ in Cartesian coordinates are computed
as Euclidian distances, in two dimensions by
$$d_{ij} = \sqrt{(x_i-x_j)^2+(y_i-y_j)^2}$$
with $p_i = (x_i,y_i)$
and in three dimensions by
$$d_{ij} = \sqrt{(x_i-x_j)^2+(y_i-y_j)^2+(z_i-z_j)^2}$$
with $p_i = (x_i,y_i,z_i).$
These distances represent the length of a _straight_ line between
two points $i$ and $j$.

For two points on a circle, the length of the arc of two points $c_1 = (r,{\phi}_i)$ and
$c_2 = (r, \phi_2)$ is
$$s_{ij}=r~|\phi_1-\phi_2| = r ~\theta$$
with $\theta$ the angle between $\phi_1$ and $\phi_2$ in radians.
For very small values of $\theta$, we will have $s_{ij} \approx d_{ij}$.

For two points $p_1 = (\lambda_1,\phi_1)$ and $p_2 =
(\lambda_2,\phi_2)$ on a sphere with radius $r'$, the _great circle
distance_ is the arc length between $p_1$ and $p_2$ on the circle
that passes through $p_1$ and $p_2$ and has the center of the sphere as its center, and
is given by $s_{12} = r ~ \theta_{12}$ with
$$\theta_{12} = \arccos(\sin \phi_1 \cdot \sin \phi_2 + \cos \phi_1 \cdot \cos \phi_2 \cdot \cos(|\lambda_1-\lambda_2|))$$
the angle in radians.

Arc distances between two points on a spheroid are more complicated
to compute.  Details about the computation method used in package
`lwgeom` (which, in turn, is used by package `sf`) are given in
[@karney2013algorithms].

## Bounded spaces

All the above assumes pure geometric computations in spaces where
there are no unexpected obstructions. When we move through space
on a daily basis, we typically find constraints e.g.

* when we walk through a building
* when we cycle to work
* or when we drive to a shop.

Distances, in such case, can be made up of a sequence of shortest
(straight line or great circle) distances, but follow a more
complex path than a straight line. Typical constraints come from
road networks, or from the requirement for a certain habitat (water
for fish, forest for certain bird species).

## Time

When we describe over which "space" a certain phenomenon is
characterised, time forms an integral component: nothing exists
forever. It is tempting to think that time, compared to geographical
"space", is one-dimensional and that is often the way we address
time, e.g. in R where `Date` is represented by the number of days
since 1970-01-01
```{r}
(d = as.Date("1970-02-01"))
as.numeric(d)
```
and `POSIXt` time by the number of seconds since that date
```{r}
(t = as.POSIXct("1970-01-02 00:00", tz = "UTC"))
as.numeric(t)
```
In practical cases however, we may be interested in how certain
phenomena vary over the day, or over the year, in which case it may
more convenient to represent time as the tuple (year, day-of-year),
or (day, time-of-day). When we study for instance traffic patterns,
day of week plays a role, in which we may end up with (year,
week-of-year, day-of-week, time-of-day) in order to quantify hour,
day, week, and year signals.

## Exercises

* convert the $(x,y)$ points $(10,2)$, $(-10,-2)$, $(10,-2)$ and $(0,10)$ to polar coordinates
* convert the polar $(r,\phi)$ points $(10,45^{\circ})$, $(0,100^{\circ})$ and $(5,359^{\circ})$ to Cartesian coordinates
* assuming the Earth is a sphere with a radius of 6371 km, compute for $(\lambda,\phi)$ points the great circle distance between $(10,10)$ and $(11,10)$, between $(10,80)$ and $(11,80)$, between $(10,10)$ and $(10,11)$ and between $(10,80)$ and $(10,81)$ (units: degree). What are the distance units?
