# Coordinates {#cs}

"_Data are not just numbers, they are numbers with a context_";
"_In data analysis, context provides meaning_"  [@cobbmoore]

Before we can try to understand geometries like points, lines,
polygons, coverage and grids, it is useful to review coordinate
systems so that we have an idea what exactly coordinates of a
point reflect.  For spatial data, the location of observations
are characterized by coordinates, and coordinates are defined in
a coordinate system.  Different coordinate systems can be used for
this, and the most important difference is whether coordinates are
defined over a 2-dimensional or 3-dimensional space referenced
to orthogonal axes (Cartesian coordinates), or using distance
and directions (polar coordinates, spherical and ellipsoidal
coordinates). Besides a location of observation, all observations
are associated with time of observation, and so time coordinate
systems are also briefly discussed. First we will briefly review
_quantities_, to learn what units and datum are.

## Quantities, units, datum {#units}

The VIM (International vocabulary of metrology, @vim)
defines a _quantity_ as a "property of a phenomenon, body, or
substance, where the property has a magnitude that can be expressed
as a number and a reference", where "[a] reference can be a measurement
unit, a measurement procedure, a reference material, or a combination
of such."

Although one could argue whether all data is constituted of
quantities, there is no need to argue that proper data handling
requires that numbers (or symbols) are accompanied by information
on what they mean, in particular what they refer to.

A measurement system consist of _base units_ for base quantities, and
_derived units_ for derived quantities. For instance, the SI system
of units [@SI] consist of the seven base units length (metre,
m), mass (kilogram, kg), time (second, s), electric current
(ampere, A), thermodynamic temperature (Kelvin, K), amount of
substance (mole, mol), and luminous intensity (candela, cd).
Derived units are composed of products of integer powers of base
units; examples are speed ($\mbox{m}~\mbox{s}^{-1}$), density
($\mbox{kg}~\mbox{m}^{-3}$) and area ($\mbox{m}^2$).

The special case of unitless measures can refer to either cases
where units cancel out (e.g. mass fraction: kg/kg, or angle measured
in rad: m/m) or to cases where objects or events were counted
(e.g. 5 apples).  Adding an angle to a count of apples would not
make sense; adding 5 apples to 3 oranges may make sense if the
result is reinterpreted as a superclass, e.g. as _pieces of fruit_.
Many data variables have units that are not expressible as SI base
units or derived units. @hand discusses many such measurement scales,
e.g. those used to measure intelligence in social sciences, in the
context of measurement units.

For many quantities, the natural origin of values is zero. This
works for amounts, where differences between amounts result in
meaningful negative values. For locations and times, differences
have a natural zero interpretation: distance and duration.  Absolute
location (position) and time need a fixed origin, from which we
can meaningfully measure other absolute space-time points: we
call this **a datum**.
For space, a datum involves more than one dimension. The combination
of a datum and a measurement unit (scale) is a _reference system_.

We will now elaborate how spatial locations can be expressed as
either ellipsoidal or Cartesian coordinates.  The next sections will
deal with temporal and spatial reference systems, and how they are
handled in R.

## Ellipsoidal coordinates

```{r polar, out.width = '50%', fig.cap="Two-dimensional polar (red) and Cartesian (blue) coordinates", Eva=TRUE, fig=TRUE, echo=!knitr::is_latex_output() }
par(mar = rep(0,4))
plot(3, 4, xlim = c(-6,6), ylim = c(-6,6), asp = 1)
axis(1, pos = 0, at = 0:6)
axis(2, pos = 0, at = -6:6)
xd = seq(-5, 5, by = .1)
lines(xd, sqrt(25 - xd^2), col = 'grey')
lines(xd, -sqrt(25 - xd^2), col = 'grey')
arrows(0, 0, 3, 4, col = 'red', length = .15, angle = 20)
text(1.5, 2.7, label = "r", col = 'red')
xd = seq(3/5, 1, by = .1)
lines(xd, sqrt(1 - xd^2), col = 'red')
text(1.2, 0.5, label = parse(text = "phi"), col = 'red')
lines(c(3,3), c(0,4), lty = 2, col = 'blue')
lines(c(0,3), c(4,4), lty = 2, col = 'blue')
text(3.3, 0.3, label = "x", col = 'blue')
text(0.3, 4.3, label = "y", col = 'blue')
```

Figure \@ref(fig:polar) shows both polar and Cartesian coordinates
for a two-dimensional situation. In Cartesian coordinates,
the point shown is $(x,y) = (3,4)$, for polar coordinates it is
$(r,\phi) = (5, \mbox{arctan}(4/3))$, where $\mbox{arctan}(4/3)$ is
approximately $0.93$ radians, or $53^{\circ}$. Note that $x$, $y$
and $r$ all have length units, where $\phi$ is an angle (a unitless
length/length ratio). Converting back and forth between Cartesian
and polar coordinates is trivial, as
$$x = r~\mbox{cos} \phi,$$
$$y = r~\mbox{sin} \phi,$$
$$r = \sqrt{x^2 + y^2}, \ \mbox{and}$$
$$\phi = \mbox{atan2}(y, x)$$
where $\mbox{atan2}$ is used in favor of $\mbox{atan}(y/x)$ to take care
of the right quadrant.

### Ellipsoidal coordinates

In three dimensions, where Cartesian coordinates are expressed as
$(x,y,z)$, spherical coordinates are the three-dimensional equivalent
of polar coordinates and can be expressed as $(r,\lambda,\phi)$, where

* $r$ is the radius of the sphere,
* $\lambda$ is the longitude, measured in the $(x,y)$ plane counter-clockwise from positive $x$, and
* $\phi$ is the latitude, the angle between the vector and the $(x,y)$ plane.

Figure \@ref(fig:sphere) illustrates Cartesian geocentric and
ellipsoidal coordinates.

```{r sphere,echo=TRUE,fig.cap="Cartesian geocentric coordinates (left) measure three distances, ellipsoidal coordinates (right) measure two angles, and possibly an ellipsoidal height", echo=!knitr::is_latex_output()}
suppressPackageStartupMessages(library(sf))
e = cbind(-90:90,0) # equator
f1 = rbind(cbind(0, -90:90)) # 0/antimerid.
f2 = rbind(cbind(90, -90:90), cbind(270, 90:-90))# +/- 90
eq = st_sfc(st_linestring(e), st_linestring(f1), st_linestring(f2), crs=4326)

geoc = st_transform(eq, "+proj=geocent")
cc = rbind(geoc[[1]], NA, geoc[[2]], NA, geoc[[3]])
from3d = function(x, offset, maxz, minz) {
	x = x[,c(2,3,1)] + offset # move to y right, x up, z backw
	x[,2] = x[,2] - maxz      # shift y to left
	d = maxz
	z = x[,3] - minz + offset
	x[,1] = x[,1] * (d/z)
	x[,2] = x[,2] * (d/z)
	x[,1:2]
}
maxz = max(cc[,3], na.rm = TRUE)
minz = min(cc[,3], na.rm = TRUE)
offset = 3e7
circ = from3d(cc, offset, maxz, minz)
mx = max(cc, na.rm = TRUE) * 1.1
x = rbind(c(0, 0, 0), c(mx, 0, 0))
y = rbind(c(0, 0, 0), c(0, mx, 0))
z = rbind(c(0, 0, 0), c(0, 0, mx))
ll = rbind(x, NA, y, NA, z)
l0 =  from3d(ll, offset, maxz, minz)
mx = max(cc, na.rm = TRUE) * 1.2
x = rbind(c(0, 0, 0), c(mx, 0, 0))
y = rbind(c(0, 0, 0), c(0, mx, 0))
z = rbind(c(0, 0, 0), c(0, 0, mx))
ll = rbind(x, NA, y, NA, z)
l =  from3d(ll, offset, maxz, minz)

par(mfrow = c(1, 2))
par(mar=rep(0,4))
plot.new()
plot.window(xlim = c(min(circ[,1],na.rm = TRUE), 3607103*1.02), 
						ylim = c(min(circ[,2],na.rm = TRUE), 2873898*1.1), asp = 1)
lines(circ)
lines(l0)
text(l[c(2,5,8),], labels = c("x", "y", "z"), col = 'red')
# add POINT(60 47)
p = st_as_sfc("POINT(60 47)", crs = 4326) %>% st_transform("+proj=geocent")
p = p[[1]]
pts = rbind(c(0,0,0), c(p[1],0,0), c(p[1],p[2],0), c(p[1],p[2],p[2]))
ptsl =  from3d(pts, offset, maxz, minz)
lines(ptsl, col = 'blue', lty = 2, lwd = 2)
points(ptsl[4,1], ptsl[4,2], col = 'blue', cex = 1, pch = 16)

plot.new()
plot.window(xlim = c(min(circ[,1],na.rm = TRUE), 3607103*1.02), 
						ylim = c(min(circ[,2],na.rm = TRUE), 2873898*1.1), asp = 1)
lines(circ)

p = st_as_sfc("POINT(60 47)", crs = 4326) %>% st_transform("+proj=geocent")
p = p[[1]]
pts = rbind(c(0,0,0), c(p[1],p[2],p[3]))
pt =  from3d(pts, offset, maxz, minz)
lines(pt)
points(pt[2,1], pt[2,2], col = 'blue', cex = 1, pch = 16)

p0 = st_as_sfc("POINT(60 0)", crs = 4326) %>% st_transform("+proj=geocent")
p0 = p0[[1]]
pts = rbind(c(0,0,0), c(p0[1],p0[2],p0[3]))
pt =  from3d(pts, offset, maxz, minz)
lines(pt)

p0 = st_as_sfc("POINT(0 0)", crs = 4326) %>% st_transform("+proj=geocent")
p0 = p0[[1]]
pts = rbind(c(0,0,0), c(p0[1],p0[2],p0[3]))
pt =  from3d(pts, offset, maxz, minz)
lines(pt)

p0 = st_as_sfc("POINT(0 90)", crs = 4326) %>% st_transform("+proj=geocent")
p0 = p0[[1]]
pts = rbind(c(0,0,0), c(p0[1],p0[2],p0[3]))
pt =  from3d(pts, offset, maxz, minz)
lines(pt, lty = 2)

p0 = st_as_sfc("POINT(90 0)", crs = 4326) %>% st_transform("+proj=geocent")
p0 = p0[[1]]
pts = rbind(c(0,0,0), c(p0[1],p0[2],p0[3]))
pt =  from3d(pts, offset, maxz, minz)
lines(pt, lty = 2)

f1 = rbind(cbind(0:60, 0))
arc = st_sfc(st_linestring(f1), crs=4326)
geoc = st_transform(arc, "+proj=geocent")
cc = rbind(geoc[[1]])
circ = from3d(cc, offset, maxz, minz)
lines(circ, col = 'red', lwd = 2, lty = 2)

f1 = rbind(cbind(60, 0:47))
arc = st_sfc(st_linestring(f1), crs=4326)
geoc = st_transform(arc, "+proj=geocent")
cc = rbind(geoc[[1]])
circ = from3d(cc, offset, maxz, minz)
lines(circ, col = 'blue', lwd = 2, lty = 2)

text(pt[1,1]+100000, pt[1,2]+50000, labels = expression(phi), col = 'blue') # lat
text(pt[1,1]+20000, pt[1,2]-50000, labels = expression(lambda), col = 'red') # lng
```

$\lambda$ typically varies between $-180^{\circ}$ and $180^{\circ}$
(or alternatively from $0^{\circ}$ to $360^{\circ}$), $\phi$ from
$-90^{\circ}$ to $90^{\circ}$.  When we are only interested in points
_on_ a sphere with given radius, we can drop $r$: $(\lambda,\phi)$
now suffice to identify any point.

It should be noted that this is just _a_ definition, one could for
instance also choose to measure polar angle, the angle between
the vector and $z$, instead of latitude. There is also a long
tradition of specifying points as $(\phi,\lambda)$ but throughout
this book we will stick to longitude-latitude, $(\lambda,\phi)$.
The point denoted in figure \@ref(fig:sphere) has $(\lambda,\phi)$
or ellipsoidal coordinates with values
```{r echo=!knitr::is_latex_output()}
p = st_as_sfc("POINT(60 47)", crs = 4326)
p[[1]]
```

with angles measured in degrees, and geocentric coordinates
```{r echo=!knitr::is_latex_output()}
p = st_as_sfc("POINT(60 47)", crs = 4326) %>% st_transform("+proj=geocent")
p[[1]]
```
with unit metres.

For points on an ellipse, there are two ways in which angle can be
expressed (figure \@ref(fig:ellipse)): measured from the center of
the ellipse ($\psi$), or measured perpendicular to the tangent on
the ellipse at the target point ($\phi$).

```{r ellipse, out.width = '50%', fig.cap="Angles on an ellipse: geodetic (blue) and geocentric (red) latitude", eval=TRUE, fig=TRUE, echo=!knitr::is_latex_output() }
par(mar = rep(0,4))
x = 4
y = 5/8 * sqrt(48)
plot(x, y, xlim = c(-6,6), ylim = c(-8,8), asp = 1)
axis(1, pos = 0, at = 0:9)
axis(2, pos = 0, at = -5:5)
xd = seq(-8, 8, by = .1)
lines(xd, 5/8 * sqrt(64 - xd^2), col = 'grey')
lines(xd, 5/8 * -sqrt(64 - xd^2), col = 'grey')
arrows(0, 0, x, y, col = 'red', length = .15, angle = 20)
b = (x * 25) / (-y * 64)
a = y - x * b
abline(a, b, col = 'grey')
b = -1/b
x0 = x - y / b
arrows(x0, 0, x, y, col = 'blue', length = .15, angle = 20)
text(1.2, 0.5, label = parse(text = "psi"), col = 'red')
text(3, 0.5, label = parse(text = "phi"), col = 'blue')
```

The most commonly used parametric model for the Earth is _an
ellipsoid of revolution_, an ellipsoid with two equal semi-axes
[@iliffelott]. In effect, this is a flattened sphere (or spheroid):
the distance between the poles is (slightly: about 0.33%) smaller
than the distance between two opposite points on the equator. Under
this model, longitude is always measured along a circle (as in
figure \@ref(fig:sphere)), and latitude along an ellipse (as in
figure \@ref(fig:ellipse)). If we
think of figure \@ref(fig:ellipse) as a cross section of the Earth
passing through the poles, the _geodetic_ latitude measure $\phi$
is the one used when no further specification is given. The latitude
measure $\psi$ is called the _geocentric latitude_.

In addition to longitude and latitude we can add _altitude_
or elevation to define points that are not on the ellipsoid, and
obtain a three dimensional space again. When defining altitude,
we need to choose

* where zero altitude is: on the ellipsoid, or relative to the surface approximating mean sea level (the geoid)?
* which direction is positive, and 
* which direction is "straight up": perpendicular to the ellipsoid surface,
or in the direction perpendicular to the surface of the geoid?

All these choices may matter, depending on the application area
and required measurement accuracies.

The shape of the Earth is not a perfect ellipsoid. As a consequence,
several ellipsoids with different shape parameters and bound to
the Earth in different ways are being used. Such ellipsoids are called
_datums_, and are briefly discussed in section \@ref(crs), along
with _coordinate reference systems_.

### Projected coordinates, distances {#projections}

Because paper maps and computer screens are much more abundant
and practical than globes, most of the time we look at spatial
data we see it _projected_: drawn on a flat, two-dimensional
surface. Computing the locations in a two-dimensional space means
that we work with _projected_ coordinates. Projecting ellipsoidal
coordinates means that shapes, directions, areas, or even all three,
are distorted [@iliffelott].

Distances between two points $p_i$ and $p_j$ in Cartesian coordinates are computed
as Euclidean distances, in two dimensions by
$$d_{ij} = \sqrt{(x_i-x_j)^2+(y_i-y_j)^2}$$
with $p_i = (x_i,y_i)$
and in three dimensions by
$$d_{ij} = \sqrt{(x_i-x_j)^2+(y_i-y_j)^2+(z_i-z_j)^2}$$
with $p_i = (x_i,y_i,z_i).$
These distances represent the length of a _straight_ line between
two points $i$ and $j$.

For two points on a circle, the length of the arc of two points $c_1 = (r,{\phi}_i)$ and
$c_2 = (r, \phi_2)$ is
$$s_{ij}=r~|\phi_1-\phi_2| = r ~\theta$$
with $\theta$ the angle between $\phi_1$ and $\phi_2$ in radians.
For very small values of $\theta$, we will have $s_{ij} \approx d_{ij}$,
because a small arc segment is nearly straight.

For two points $p_1 = (\lambda_1,\phi_1)$ and $p_2 =
(\lambda_2,\phi_2)$ on a sphere with radius $r'$, the _great circle
distance_ is the arc length between $p_1$ and $p_2$ on the circle
that passes through $p_1$ and $p_2$ and has the center of the sphere as its center, and
is given by $s_{12} = r ~ \theta_{12}$ with
$$\theta_{12} = \arccos(\sin \phi_1 \cdot \sin \phi_2 + \cos \phi_1 \cdot \cos \phi_2 \cdot \cos(|\lambda_1-\lambda_2|))$$
the angle between $p_1$ and $p_2$, in radians.

Arc distances between two points on a spheroid are more complicated
to compute; a good discussion on the topic and an explanation of
the method implemented in GeographicLib (part of PROJ) is given
in @karney2013algorithms.

To show that these distance measures actually give different
values, we computed them for the distance Berlin - Paris. Here,
`gc_` refers to ellipsoidal and spherical great circle distances,
`straight_` refers to straight line, Euclidean distances between
Cartesian geocentric coordinates associated on the WGS84 ellipse
and sphere:
```{r echo=!knitr::is_latex_output()}
pts = st_sfc(st_point(c(13.4050, 52.5200)), st_point(c(2.3522, 48.8566)), crs = 'EPSG:4326')
s2_orig = sf_use_s2(FALSE)
d1 = c(gc_ellipse = st_distance(pts)[1,2])
sf_use_s2(TRUE)
# or, without using s2, use st_distance(st_transform(pts, "+proj=longlat +ellps=sphere"))
d2 = c(gc_sphere = st_distance(pts)[1,2])
p = st_transform(pts, "+proj=geocent")
d3 = c(straight_ellipse = units::set_units(sqrt(sum(apply(do.call(cbind, p), 1, diff)^2)), m))
p2 = st_transform(pts, "+proj=longlat +ellps=sphere") %>% st_transform("+proj=geocent")
d4 = c(straight_sphere = units::set_units(sqrt(sum(apply(do.call(cbind, p2), 1, diff)^2)), m))
res = c(d1,d3,d2,d4)
# print as km, re-add names:
sf_use_s2(s2_orig) # as it was
res %>% units::set_units(km) %>% setNames(names(res)) %>% print(digits = 5)
```

### Bounded and unbounded spaces {#bounded}

Two-dimensional and three-dimensional Euclidean spaces ($R^2$ and
$R^3$) are unbounded: every line in this space has infinite length,
distances, areas or volumes are unbounded. In contrast, spaces
defined on a circle ($S^1$) or sphere ($S^2$) define a bounded set:
there may be infinitely many points but the length and area of the
circle and the radius, area and volume of a sphere are bound.

This may sound trivial, but leads to some interesting findings
when handling spatial data. A polygon on $R^2$ has unambiguously an
inside and an outside. On a sphere, $S^2$, any polygon divides
the sphere in two parts, and which of these two is to be considered
inside and which outside is ambiguous and needs to be defined
e.g. by the traversal direction. Chapter \@ref(spherical) will
further discuss consequences when working with geometries on $S^2$.

## Coordinate Reference Systems {#crs}

We follow @lott2015 when defining the following concepts (italics indicate literal quoting):

* a **coordinate system** is a _set of mathematical rules for specifying how coordinates are to be assigned to points_
* a **datum** is a _parameter or set of parameters that define the position of the origin, the scale, and the orientation of a coordinate system_, 
* a **geodetic datum** is a _datum describing the relationship of a two- or three-dimensional coordinate system to the Earth_, and
* a **coordinate reference system** is a _coordinate system that is related to an object by a datum; for geodetic and vertical datums, the object will be the Earth._

A readable text that further explains these concepts is @iliffelott.

The Earth does not follow a regular shape. The topography of the
Earth is of course known to vary strongly, but also the surface
formed by constant gravity at mean sea level, the geoid, is
irregular. A commonly used model that is fit to the geoid is an
ellipsoid of revolution, which is an ellipse with two identical
minor axes. Fitting such an ellipsoid to the Earth gives a datum.
However, fitting it to different areas, or based on different sets of
reference points gives different fits, and hence different datums:
a datum can for instance be fixed to a particular tectonic plate
(like ETRS89), others can be globally fit (like WGS84). More local
fits lead to smaller approximation errors.

The definitions above imply that coordinates in degrees longitude
and latitude only have a meaning, i.e. can only be interpreted
unambiguously as Earth coordinates, when the datum they are
associated with is given.

Note that for projected data, the data that _were_ projected
are associated with a reference ellipsoid (datum). Going from
one projection to another _without_ changing datum is called
_coordinate conversion_, and passes through the ellipsoidal
coordinates associated with the datum involved. This process is
lossless and invertible: the parameters and equations associated
with a _conversion_ are not empirical. Recomputing coordinates in a
new datum is called _coordinate transformation_, and is approximate:
because datums are a result of model fitting, transformations
between datums are models too that have been fit; the equations
involved are empirical, and multiple transformation paths, based
on different model fits and associated with different accuracies,
are possible.

Plate tectonics imply that within a global datum, fixed objects may
have coordinates that change over time, and that transformations
from one datum to another may be time-dependent. Earthquakes are a
cause of more local and sudden changes in coordinates.


## PROJ and mapping accuracy {#proj}

Very few living people active in open source geospatial software
can remember the time before PROJ. PROJ [@evenden:90] started in the
1970s as a Fortran project, and was released in 1985 as a C library
for cartographic projections. It came with command line tools for
direct and inverse projections, and could be linked to software
to let it support (re)projection directly. Originally, datums were
considered implicit, and no datum transformations were allowed.

In the early 2000s, PROJ was known as PROJ.4, after its never changing
major version number.  Amongst others motivated by the rise of GPS,
the need for datum transformations increased and PROJ.4 was extended
with rudimentary datum support. PROJ definitions for coordinate
reference systems would look like this:
```
+proj=utm +zone=33 +datum=WGS84 +units=m +no_defs
```
where _key_=_value_ pairs are preceded by a `+` and separated by
a space. This form came to be known as "PROJ.4 string", since the
PROJ project stayed at version 4.x for several decades. Other datums
would come with fields like
```
+ellps=bessel +towgs84=565.4,50.3,465.6,-0.399,0.344,-1.877,4.072
```
indicating another ellipse, as well as the seven (or three)
parameters for transforming from this ellipse to WGS84 (the "World
Geodetic System 1984" global datum once popularized by GPS),
effectively defining the datum in terms of a transformation to WGS84.

Along with PROJ.4 came a set of databases with known (registered)
projections, from which the best known is the EPSG registry.
National mapping agencies would provide (and update over time)
their best guesses of `+towgs84=` parameters for national coordinate
reference systems, and distribute it through the EPSG registry,
which was part of PROJ distributions.
For some transformations, _datum grids_ were available and
distributed as part of PROJ.4: such grids are raster maps that provide
for every location pre-computed values for the shift in longitude
and latitude, or elevation, for a particular datum transformation.

```{r horizontalgrid, fig.cap="UK horizontal datum grid, from datum OSGB 1936 (EPSG:4277) to datum ETRS89 (EPSG:4258); units arc-seconds", echo=!knitr::is_latex_output() }
library(stars)
library(rnaturalearth)
countries110 = st_as_sf(countries110)
uk = countries110[countries110$admin %in% c("United Kingdom"),] %>%
		st_geometry()
r = read_stars("data/uk_os_OSTN15_NTv2_OSGBtoETRS.tif")
# r = read_stars("/vsicurl/https://cdn.proj.org/uk_os_OSTN15_NTv2_OSGBtoETRS.tif")
hook = function() {
		plot(uk, border = "orange", col = NA, add = TRUE)
}
plot(r[,,,1:2], axes = TRUE, hook = hook, key.pos = 4)
```

```{r verticalgrid, fig.cap="UK vertical datum grid, from ETRS89 (EPSG:4937) to ODN height (EPSG:5701), units m", echo=!knitr::is_latex_output() }
h = read_stars("data/uk_os_OSGM15_GB.tif")
# h = read_stars("/vsicurl/https://cdn.proj.org/uk_os_OSGM15_GB.tif")
plot(h, axes = TRUE, reset = FALSE)
plot(uk, border = "orange", col = NA, add = TRUE)
```

In PROJ.4, every coordinate transformation had to go through a
conversion to and from WGS84; even reprojecting data associated with
a datum different from WGS84 had to go through a transformation to
and from WGS84. The associated errors of up to 100 m were acceptable
for mapping purposes for not too small areas, but applications that
need high accuracy transformations, e.g. precision agriculture,
planning flights of UAV's, or object tracking are often more
demanding in terms of accuracy.

In 2018, after a successful "GDAL Coordinate System Barn Raising"
initiative, a number of companies profiting from the open source
geospatial software stack supported the development of a more modern,
mature coordinate transformation system in PROJ. Over a few years,
PROJ.4 evolved through versions 5, 6, 7 and 8 and was hence renamed
into PROJ (or PR$\phi$J).

The most notable changes include:

* although PROJ.4 strings can still be used to initialize certain
coordinate reference systems, they are no longer sufficient to
represent all of them; a new format, WKT2 (described in next section)
replaces it.
* WGS84 as a hub datum is dropped: coordinate transformation no longer
need to go through a particular datum
* Multiple conversion or transformation paths (so-called pipelines)
to go from CRS A to CRS B are possible, and can be reported along
with the associated accuracy; PROJ will by default use the most accurate
one but user control is possible
* transformation pipelines can chain an arbitrary number of
elementary transformation operations, including swapping of axes
and unit transformations 
* datum grids, of which there are now _many_ more, are no longer distributed
with the library but are accessible from a CDN; PROJ allows to enabling and
disabling network access to access these grids, and only downloads the section 
of the grid actually needed, storing it in a cache on the user's machine for 
future use 
* coordinate transformations receive support for epochs, time-dependent
transformations (and hence: four-dimensional coordinates, including the
source and target time)
* the set of files with registered coordinate reference systems is handled
in an SQLite database.
* instead of always handling axis order (longitude, latitude),
when the authority defines differently this is now obeyed (with the
most notable example: EPSG:4326 defines axis order to be latitude,
longitude.)

All these points sound like massive improvements, and accuracies
of transformation can be below 1 metre. An interesting point
is the last: Where we could safely assume for many decades that
spatial data with ellipsoidal coordinates would have axis order
(longitude, latitude), this is no longer the case. We will see in
section \@ref(axisorder) how to deal with this.

Examples of a horizontal datum grids, downloaded from cdn.proj.org,
are shown in figure \@ref(fig:horizontalgrid) and for a vertical
datum grid in figure \@ref(fig:verticalgrid). Datum grids may carry
per-pixel accuracy values.

## WKT-2 {#wkt2}

@lott2015 describes a standard for encoding coordinate reference
systems, as well as transformations between them using _well known
text_; the standard (and format) is referred to informally as
WKT-2. As mentioned above, GDAL and PROJ fully support this encoding.
An example of WKT2 for CRS `OGC:CRS84` is:

```
GEOGCRS["WGS 84",
    DATUM["World Geodetic System 1984",
        ELLIPSOID["WGS 84",6378137,298.257223563,
            LENGTHUNIT["metre",1]],
        ID["EPSG",6326]],
    PRIMEM["Greenwich",0,
        ANGLEUNIT["degree",0.0174532925199433],
        ID["EPSG",8901]],
    CS[ellipsoidal,2],
        AXIS["longitude",east,
            ORDER[1],
            ANGLEUNIT["degree",0.0174532925199433,
                ID["EPSG",9122]]],
        AXIS["latitude",north,
            ORDER[2],
            ANGLEUNIT["degree",0.0174532925199433,
                ID["EPSG",9122]]]]
```
This shows a WGS84 ellipsoid, and a coordinate system with the axis
order (longitude, latitude) that can be used to replace `EPSG:4326`
when one wants unambiguously "traditional" (GIS) axis order.

A longer introduction on the history and recent changes in PROJ is
given in [@rogerCRS], building upon the work of [@knudsen+evers17;
@evers+knudsen17]

## Exercises

* list three geographic measures that do not have a natural zero origin
<!-- latitude, longitude, elevation -->
* convert the $(x,y)$ points $(10,2)$, $(-10,-2)$, $(10,-2)$ and $(0,10)$ to polar coordinates
* convert the polar $(r,\phi)$ points $(10,45^{\circ})$, $(0,100^{\circ})$ and $(5,359^{\circ})$ to Cartesian coordinates
* assuming the Earth is a sphere with a radius of 6371 km, compute for $(\lambda,\phi)$ points the great circle distance between $(10,10)$ and $(11,10)$, between $(10,80)$ and $(11,80)$, between $(10,10)$ and $(10,11)$ and between $(10,80)$ and $(10,81)$ (units: degree). What are the distance units?
<!-- relevant? Too difficult? -->
