# Geometries {#geometries}

Having learned how we describe spaces, we can define how geometries
can be described in these space. This chapter will mostly explain the
geometries for _simple features_, and introduce the three classes
`sfg`, `sfc` and `sf` for single geometries, geometry sets, and
geometry sets with associated attributes.

## Simple feature geometry types

Simple feature geometries are a way to describe the geometries of
_features_.  By _features_ we mean _things_ that have a geometry,
some time properties, and other attributes.  The main application of
simple feature geometries is to describe two-dimensional geometries
by points, lines, or polygons. The "simple" adjective refers to
the fact that the line or polygon geometries are represented by
sequences of points connected with straight lines.

_Simple features access_ is a standard [@sfa, @sfa2, @iso]
for describing simple feature geometries that includes

* a class hierarchy
* a set of operations
* binary and text encodings

We will now discuss the seven most common simple feature geometry
types. Although in practice we will most often import spatial data
from external sources (files, databases, web services), we will
create them here from scratch using simple constructor functions.

### The big seven

The most commonly used simple features geometries, used to represent a _single_ feature are:

| type | description                                                               |
|------|---------------------------------------------------------------------------|
| `POINT` | single point geometry |
| `MULTIPOINT` | set of points |
| `LINESTRING` | single linestring (two or more points connected by straight lines) |
| `MULTILINESTRING` | set of linestrings |
| `POLYGON` | exterior ring with zero or more inner rings, denoting holes |
| `MULTIPOLYGON` | set of polygons |
| `GEOMETRYCOLLECTION` | set of the geometries above  |

Points in a geometry contain at least two coordinates: x and y, in that order.

### Valid geometries

Valid geometries obey the following properties:

* linestrings shall not self-intersect
* polygon rings shall be closed (the last point equals the first)
* polygon holes (inner rings) shall be inside their exterior ring
* polygon inner rings shall maximally touch the exterior ring in single points, not over a line
* a polygon ring shall not repeat its own path

If this is not the case, the geometry concerned is not valid.

### Z and M

In addition to X and Y coordinates, Single points (vertices) of
simple feature geometries can have

* a `Z` coordinate, denoting altitude, and/or
* an `M` value, denoting some "measure"

The `M` attribute shall be a property of the vertex. It sounds
attractive to encode a time stamp in it, e.g. to pack trajectories
in `LINESTRING`s. These become however invalid once the trajectory
self-intersects.

Both `Z` and `M` are found relatively rarely, and software support
to do something useful with them is (still) rather rare.

### Ten further geometry types

There are 10 more geometry types which are more rare, but increasingly find implementation:

| type | description                                        |
| ---- | -------------------------------------------------- |
| `CIRCULARSTRING` | The CIRCULARSTRING is the basic curve type, similar to a LINESTRING in the linear world. A single segment requires three points, the start and end points (first and third) and any other point on the arc. The exception to this is for a closed circle, where the start and end points are the same. In this case the second point MUST be the center of the arc, ie the opposite side of the circle. To chain arcs together, the last point of the previous arc becomes the first point of the next arc, just like in LINESTRING. This means that a valid circular string must have an odd number of points greated than 1. |
| `COMPOUNDCURVE` | A compound curve is a single, continuous curve that has both curved (circular) segments and linear segments. That means that in addition to having well-formed components, the end point of every component (except the last) must be coincident with the start point of the following component. |
| `CURVEPOLYGON` | Example compound curve in a curve polygon: CURVEPOLYGON(COMPOUNDCURVE(CIRCULARSTRING(0 0,2 0, 2 1, 2 3, 4 3),(4 3, 4 5, 1 4, 0 0)), CIRCULARSTRING(1.7 1, 1.4 0.4, 1.6 0.4, 1.6 0.5, 1.7 1) ) |
| `MULTICURVE` |  A MultiCurve is a 1-dimensional GeometryCollection whose elements are Curves, it can include linear strings, circular strings or compound strings.  |
| `MULTISURFACE` | A MultiSurface is a 2-dimensional GeometryCollection whose elements are Surfaces, all using coordinates from the same coordinate reference system. |
| `CURVE` | A Curve is a 1-dimensional geometric object usually stored as a sequence of Points, with the subtype of Curve specifying the form of the interpolation between Points |
| `SURFACE` | A Surface is a 2-dimensional geometric object |
| `POLYHEDRALSURFACE` | A PolyhedralSurface is a contiguous collection of polygons, which share common boundary segments  |
| `TIN` | A TIN (triangulated irregular network) is a PolyhedralSurface consisting only of Triangle patches.|
| `TRIANGLE` | A Triangle is a polygon with 3 distinct, non-collinear vertices and no interior boundary |

Note that `CIRCULASTRING`, `COMPOUNDCURVE` and `CURVEPOLYGON`
are not described in the SFA standard, but in the [SQL-MM part
3 standard](https://www.iso.org/standard/38651.html). The
descriptions above were copied from the [PostGIS
manual](http://postgis.net/docs/using_postgis_dbmanagement.html).

### Encodings

Part of the simple feature standard are two encodings: a text and a
binary encoding.  The text strings `POINT (0 1)` and so on indicate
text encodings, also known as _well-known text_ (WKT) encodings,
of simple feature geometries. They are meant to be human-readable.


## Simple features in `sf`

This section describes the implementation of simple feature
geometries in package `sf`.  It will first explain how single
simple feature geometries, explained in the previous section,
are represented in R objects of class `sfg`. Next, it will explain
how _sets_ of simple feature geometry objects are collected in a
list of class `sfc`. This list acts as a geometry _list-column_
in data.frame objects, of class `sf`.

### `sfg`: simple feature geometry

Point sets are stored as numeric matrix, with 2 (`XY`), 3 (`XYZ`
or `XYM`) or 4 (`XYZM`) columns, with a points in each row.
Individual simple feature geometry objects are implemented as:

* numeric vector for `POINT`,
* numeric matrix for `MULTIPOINT` and `LINESTRING`
* list of numeric matrices for `MULTILINESTRING` and `POLYGON`
* list of lists of numeric matrices for `MULTIPOLYGON`
* list of (typed) geometries for `GEOMETRYCOLLECTION`

All other geometry types follow this, using the simplest possible option. 
Note that matrices can have zero points, and lists can have zero
elements, in which case we have _empty_ geometries; more about this
in section \@ref(empty).

Objects have a class indicating their dimension, type, and a
superclass (`sfg`: simple feature geometry), and have no other
attributes than their S3 class:

```{r}
(pt = st_point(c(0,1)))
attributes(pt)
```

We see that in addition to `sfg` the `class` attribute has two values:

* `XY` telling the dimension of the point(s), can also be `XYZ`, `XYM` or `XYZM`
* `POINT` revealing the geometry type.

Examples of `XYZ` and `XYM` and `XYZM` geometries are found here:
```{r}
system.file("shape/storms_xyz_feature.shp", package="sf") %>%
	st_read()
system.file("shape/storms_xyzm_feature.shp", package="sf") %>% # badly named!
	st_read()
(pzm = st_point(c(1,2,3,4)))
```


A `MULTIPOINT` or a `LINESTRING` can be created by a matrix
```{r}
(m1 = rbind(c(8, 1), c(2, 5), c(3, 2)))
(mp = st_multipoint(m1))
(ls = st_linestring(m1))
```
Although these geometries contain the same points, they have entirely different meaning:
the point set is a zero-dimensional, the line a one-dimensional geometry:
```{r}
st_dimension(mp)
st_dimension(ls)
```
A `MULTILINESTRING` can be constructed from a list of matrices, representing vertices:
```{r}
m2 = rbind(c(22,20), c(18, 15))
(mls = st_multilinestring(list(m1, m2)))
```
A `POLYGON` consists of an outer ring, followed by zero or more
inner rings that denote holes in the outer ring:
```{r}
(ring1 = rbind(c(0,0), c(4,0), c(4,4), c(0,4), c(0,0)))
(p1 = st_polygon(list(ring1)))
(ring2 = ring1 + 5)
(ring3 = (ring1[5:1,] / 4) + 6)
(p2 = st_polygon(list(ring2, ring3)))
```
A `MULTIPOLYGON` can be constructed as a list of lists of matrices:
```{r}
(mpol = st_multipolygon(list(list(ring1), list(ring2, ring3))))
```
And finally, a `GEOMETRYCOLLECTION` can be constructed from a list of typed geometries:
```{r}
st_geometrycollection(list(pt, mp, ls, mpol))
```

#### WKT, WKB encodings

By default, package
`sf` prints the same number of digits as R, but this can be manipulated:
```{r}
st_point(c(1/3, 2/3))
print(st_point(c(1/3, 2/3)), digits = 16)
print(st_point(c(1/3, 2/3)), digits = 3)
```

An encoding that is more useful for machine-to-machine communication is
_well-known binary_. An example of a round-trip R $\rightarrow$ binary $\rightarrow$ R is
```{r}
(wkb = st_as_binary(st_point(c(1/3, 2/3))))
st_as_sfc(wkb)[[1]]
```
Object `r` is a raw vector, which is little useful in R. Binary
conversion is used to communicate geometries to external libraries
(GDAL, GEOS, liblwgeom) and and spatial databases because it is fast
and lossless. Whenever there is a choice, binary encoding should
be prefered over text encoding.


#### simple, valid, empty {#empty}

Methods `st_is_simple` and `st_is_valid` help detect non-simple and non-valid geometries:
```{r}
st_is_simple(st_linestring(rbind(c(0,0), c(1,1), c(1,0), c(0,1)))) # self-intersects
st_is_valid(st_polygon(list(rbind(c(1,1), c(0,0), c(1,1), c(2,2), c(2,1), c(1,1))))) # repeats
```

A very important concept in the feature geometry framework is that of the
empty geometry. We can think of an empty geometry as similar to the
`NA` value in R vectors: it is a placeholder, but a usable value is
not available. Empty geometries arise naturally when we do geometrical
operations (chapter \@ref(geommanip)), for instance when we want to
know where two disjoint geometries coincide:
```{r}
(e = st_intersection(st_point(c(0,0)), st_point(c(1,1))))
```

It is not entirely clear what the benefit is of having _typed_
empty geometries, but according to the simple feature standard
they are. They are detected by
```{r}
st_is_empty(e)
```

#### Conversion between geometry types

Up to the extent that a conversion is feasible, we can convert
simple feature geometries using the `st_cast` generic:
```{r}
methods(st_cast)
```

Conversion is required e.g. to be able to plot curved geometries.
`CURVE`, `COMPOUNDCURVE` and `CIRCULARSTRING` have `st_cast` methods to cast them to `LINESTRING`;
`MULTISURFACE` has an `st_cast` method to `MULTIPOLYGON`. An example, needed for plotting, is
```{r}
(ls <- st_as_sfc("CIRCULARSTRING(0 0,1 0,1 1)") %>% st_cast("LINESTRING"))
plot(ls, axes = TRUE)
```

It is convenient in other cases to analyse the point pattern
from a set of vertices in a linestring. However,
```{r}
rbind(c(0,0), c(1,1), c(1,0), c(0,1)) %>% st_linestring() %>% st_cast("POINT")
```
does not what we expect, because it will convert a single geometry into
a new single geometry. We can convert to a `MULTIPOINT`
```{r}
rbind(c(0,0), c(1,1), c(1,0), c(0,1)) %>% st_linestring() %>% st_cast("POINT")
```
but if we want to have a set of points, we need to work with sets (section \@(sfc)) first, because
we want a set with another cardinality:
```{r}
(p <- rbind(c(0,0), c(1,1), c(1,0), c(0,1)) %>% st_linestring() %>% st_sfc() %>% st_cast("POINT"))
```

#### `GEOMETRYCOLLECTION` handling

Single features can have a geometry that consists of several
subgeometries of different type, held in a `GEOMETRYCOLLECTION`.
This may sound like looking for trouble, but these arise rather
naturally when looking for intersections. For instance, the
intersection of two `LINESTRING` geometries may be the combination
of a `LINESTRING` and a `POINT`. Putting this intersection into a
single feature geometry needs a `GEOMETRYCOLLECTION`.

In case we end up with `GEOMETRYCOLLECTION` objects, the next question is
often what to do with them. One thing we can do is extract elements from them:
```{r}
pt <- st_point(c(1, 0))
ls <- st_linestring(matrix(c(4, 3, 0, 0), ncol = 2))
poly1 <- st_polygon(list(matrix(c(5.5, 7, 7, 6, 5.5, 0, 0, -0.5, -0.5, 0), ncol = 2)))
poly2 <- st_polygon(list(matrix(c(6.6, 8, 8, 7, 6.6, 1, 1, 1.5, 1.5, 1), ncol = 2)))
multipoly <- st_multipolygon(list(poly1, poly2))

j <- st_geometrycollection(list(pt, ls, poly1, poly2, multipoly))

st_collection_extract(j, "POLYGON")
st_collection_extract(j, "POINT")
st_collection_extract(j, "LINESTRING")
```
which sometimes results in a geometry set, sometimes in single geometries.

### `sfc`: sets of geometries {#sfc}

Rather than handling geometries individually, we typically handle
them as sets.
Package `sf` provides a dedicated class for this, called `sfc`
(for **s**imple **f**eature geometry list **c**olumn). We can create such a
list column with constructor function `st_sfc`:
```{r}
(sfc = st_sfc(st_point(c(0,1)), st_point(c(-3,2)), crs = 4326))
```
The default report from the `print` method for `sfc` gives

* the number of features geometries
* the feature geometry type (here: `POINT`)
* the feature geometry dimension (here: `XY`)
* the bounding box for the set
* the coordinate reference system for the set (`epsg` and `proj4string`: see chapter \@ref(crs))
* the first few geometries, as (abbreviated) WKT 

The class of the geometry list-column,
```{r}
class(sfc)
```
is again a combination of a specific class, and a superclass.
In addition to a `class`, the object has further attributes
```{r}
attributes(sfc) %>% names() %>% setdiff("class")
```
which are used to record _for the hole set_:

* a precision value (section \@ref(precision))
* the bounding box enclosing all geometries (for x and y)
* a coordinate reference system (section \@ref(crs))
* the number of empty geometries contained in the set

This means that all these properties are defined for the set,
and not for geometries individually.

As we've seen above, sets of geometries arise when we tear apart compound geometries, as in
```{r}
(p <- rbind(c(0,0), c(1,1), c(1,0), c(0,1)) %>% st_linestring() %>% st_sfc() %>% st_cast("POINT"))
```
Here, `st_sfc` creates a set of one `LINESTRING`, and the resulting set has size 4:
```{r}
length(p)
```
Going the other way around, we need `st_combine` to combine geometries into one:
```{r}
p %>% st_combine
p %>% st_combine %>% st_cast("LINESTRING")
```

More general, in practice we will almost always work with sets
of geometries, because in spatial data we typically associate an
observation with a feature, which has a geometry, and we work with
sets of observations. 

`sfc` objects are lists with each entry being an `sfg` object:
```{r}
p[[2]]
```
and we will use these lists as _list columns_ in `data.frame` or
`tibble` objects to represent simple features with geometries in
a list column. These objects are of class `sf` (section \@ref{sf}).

#### Feature sets with mixed geometries

Sets of simple features also consist of features with heterogeneous
geometries. In this case, the geometry type of the set is `GEOMETRY`:
```{r}
(g = st_sfc(st_point(c(0,0)), st_linestring(rbind(c(0,0), c(1,1)))))
```
These can be filtered by using `st_is`
```{r}
g %>% st_is("LINESTRING")
```
or, when working with `sf` objects, 
```{r}
st_sf(g) %>% filter(st_is(., "LINESTRING"))
```


### `sf`: geometries with attributes {#sf}

`sf` objects are `tibbles` or `data.frame`s with feature geometry
in a list column, and an `sf` class added:
```{r}
sf = st_sf(sfc)
class(sf)
```
Although there is nothing against simply using data.frames or tibbles
with `sfc` list columns, a number of methods have been written for
`sf` objects that make life even more convenient, including
plot methods to create maps.

In addition to the usual `data.frame` attributes, `sf` objects have two more attributes:
```{r}
sf %>% attributes() %>% names() %>% setdiff(c("row.names", "class", "names"))
```
They are:

* `sf_column`: a length one character vector with the name of the
(active) geometry list-column. Note that `sf` objects may contain multiple
geometry list-columns, but the one named here is used for all operations,
as the ``active'' geometry.
* `agr`: **a**ttribute-**g**eometry **r**elationships; this encodes for each of
the attributes how it relates to the geometry (in case of a non-point
geometry): is it constant throughout the geometry like a soil type,
is it an aggregate over the geometry like a population count,
or does it identify the geometry like a state name? This is explained
in more depth in section \@ref(agr).

## Tesselations: coverages, rasters {#geomraster}

A common case in spatial data analysis is that an area is split
(tesselated) in a number of non-overlapping regions. Although
this can be modelled by a sequence of simple feature geometries
(polygons), it is hard to guarantee for a set of simple feature
polygons that they overlap nowhere, or that there are no gaps
between them.

More fundamental ways of storing such polygons use a _topological_
model, examples of this are found in geographic information systems
like GRASS GIS or ArcGIS. Topological models store every boundary
between polygons only once, and register which polygon is on either
side of a boundary.

A simpler approach, associated with the term _raster data_,
is to tesselate each spatial dimension $d$ into regular cells,
formed e.g. by left-closed and right-open intervals $d_i$:
$$d_i = d_0 + [~ i \cdot \delta, (i+1) \cdot \delta~)$$
with $d_0$ an offset, $\delta$ the interval (cell or
pixel) size, and where the cell index $i$ is an arbitrary but
consecutive set of integers.

In arbitrary polygon tesselations, assigning points to polygons
when they fall on a boundary shared by two polygons is ambiguous.
Using left-closed "[" and right-open ")" intervals in regular
tesselations removes this ambiguity.

Tesselating the time dimension in this way is very common, and
reflects the implicit assumption underlying time series packages
such as `xts` in R.  Different models can be combined: one could
use simple feature polygons to tesselate space, and combine this
with a regular tesselation of time in order to cover a space-time
_vector datacube_.  Raster data and data cubes are discussed in
chapter \@ref(raster).

## Networks

Spatial networks are typically composed of linear (`LINESTRING`)
elements, but possess further topological properties describing
the network coherence:

* start and endpoints of a linestring may be connected to other linestring
start or end points, forming a set of nodes and edges
* edges may be directed, and allow for connections (flow, transport)
in only one way.

Several R packages (`osmar`, `stplanr`) have (limited) functionality
available for constructing network objects, and working with them,
e.g. computing shortest or fastest routes through a network.

## Geometries on the sphere

Geometries on the sphere are geometries made from geodetic
coordinates. The easiest of these concern small regions near the
equator not covering the date line, because in that case we can
ignore all problems and still do a good job. Reality is nastier.

The concept of a bounding box, defined from the coordinate ranges
breaks easily down, e.g. when crossing the date line:
```{r}
pts = rbind(c(-179,0), c(179,0), c(179,1), c(-179,1), c(-179,0))
date_line = st_sfc(st_polygon(list(pts)), crs = 4326)
st_bbox(date_line) %>% st_as_sfc()
```
or when a polygon contains one of the poles:
```{r}
pts = rbind(c(0,89), c(120,89), c(240,89), c(0,89))
pole = st_sfc(st_polygon(list(pts)), crs = 4326)
st_bbox(pole) %>% st_as_sfc()
```
where we see that in both cases the `st_bbox` implied polygon does
not cover the area intended.

### Straight lines?

The simple feature model assumes that linestrings and polygons
are formed of points connected by straight lines. When however
representing the Earth surface, what does a straight line mean?
The simple feature standard does not help much, here: it assumes
Carthesian space.  Technically speaking, a straight line between
two points on a sphere exists, but it crosses the sphere, which is
not very practical in most cases. The most common case is to use
_great circle_ segments to connect points: the shortest path that
follows the surface of the sphere or ellipsoid. This means that
(with longitude latitude coordinates) the line between `POINT(0 50)`
and `POINT(10 50)` does not cross `POINT(5 50)`. It also means that
the line between points on opposite sides of the sphere is ambiguous.
Also, the direction of a great circle segment, when defined as the
angle it has with meridians, is not constant.

### Ring direction for polygons

The simple feature standard is not conclusive about the direction
of points in a ring. It points out that exterior rings should be
counter clockwise, when seen from above, and interior rings (holes)
clockwise, but for instance `st_is_valid` does not invalidate
clockwise exterior rings:
```{r}
st_is_valid(st_polygon(list(rbind(c(0,0), c(0,1), c(1,1), c(0,0)))))
```
This may have several reasons: a lot of data may come with wrong
ring directions, and the distinction between exterior and interior
rings is already unambiguous by their order: the first is exterior,
anything following is interior.

On the sphere, any polygon divides the sphere surface in two
finite areas, meaning there is no longer an unambiguous "inside"
vs. "outside": does the polygon with longitude latitude coordinates
`POLYGON((0 0, 120 0, 240 0, 0 0))` denote the northern or the
southern hemisphere? One can still go two directions here:

* assume that in practice polygons never divide the Earth in two equal
halves, and take the smaller area as the "inside" 
* decide strongly about ring direction, e.g. counter-clockwise
(following the ring, standing on the Earth, the left-side of
the ring denotes the polygon interior)

Package `sf` comes with a large amount of functions that work both
for projected (Cartesian) data as for data defined in spherical
coordinates. Whenever it makes assumptions of Cartesian coordinates
for spherical coordinates it emits a warning. This is discussed
further in section \@ref(longlat).
