# Raster and vector datacubes {#raster}
 
Array data are data where values are indexed along multiple array
_dimensions_.  Raster and vector datacubes refer to array data,
where one or more of the dimensions refer to space, and often other
dimensions refer to time.

## Package `stars`

Athough package `sp` has always had limited support for raster data,
over the last decade R package `raster` has clearly been dominant
as the prime package for powerful, flexible and scalable raster analysis.
Its data model is that of a 2D raster, or a set of raster layers (a
"raster stack"). This follows the classical static GIS world view,
where the world is modelled as a set of layers, each representing
a different theme. A lot of data available today however is dynamic,
and comes as time series of rasters for different themes. A raster
stack does not meaningfully reflect this, requiring the user to
do shadow book keeping of which layer represents what.  Also, the
`raster` package does an excellent job in scaling computations up
to datasizes no larger than the local storage (the computer's hard
drives). Recent datasets however, including satellite imagery,
climate model or weather forecasting data, often no longer fit in
local storage. Package `spacetime` addresses the analysis of time
series of vector geometries or raster grid cells, but does not
extend to higher-dimensional arrays.

Here, we introduce a new package for raster analysis, called `stars`
(for scalable, spatiotemporal tidy arrays) that

* allows for representing dynamic raster stacks,
* in addition to regular grids handles rotated, sheared, rectilinear and curvilinear rasters,
* provides a tight integration with package `sf`,
* follows the tidyverse design principles,
* aims at being scalable, also beyond local disk size,
* also handles array data with non-raster spatial dimensions,  the _vector datacubes_,
* provides further integration of novel features in the GDAL
library than other R packages have given so far.

Vector data cubes include for instance time series for simple
features, or spatial graph data such as origin-destination matrices.
The wider concept of spatial vector and raster data cubes is
explained in section \@ref(datacubes)

## Raster data

As introduced in section \@ref(geomraster), raster data are spatial
datasets where observations are aligned on a regular grid usually
with square grid cells (in some coordinate reference system, chapter
\@ref(rs)). Raster datasets are used often to represent spatially
continuously varying phenomena such as temperature or elevation,
and also for observed imagery for instance obtained from satellites.

### Reading and writing raster data

Raster data typically are read from a file. We read an example
file of a regular, non-rotated grid from the package `stars`:
```{r}
tif = system.file("tif/L7_ETMs.tif", package = "stars")
library(stars)
x = read_stars(tif)
```
The dataset contains (a section of) a Landsat 7 scene, with the 6
30m-resolution bands (bands 1-5 and 7) for a region covering the
city of Olinda, Brazil. 
A short summary of the data is given by
```{r}
x
```
where we see the offset, cellsize, coordinate reference system,
and dimensions. The object `x` is a simple list of length one, holding
a three-dimensional array:
```{r}
length(x)
class(x[[1]])
dim(x[[1]])
```
and in addition holds an attribute with a dimensions table with all the metadata
required to know what the array values refer to, obtained by
```{r}
st_dimensions(x)
```
We can get the spatial extent of the array by
```{r}
st_bbox(x)
```
Raster data can be written to local disk using `write_stars`:
```{r}
write_stars(x, "x.tif")
```
where the format (in this case, GeoTIFF) is derived from the file
extension. As for simple features, reading and writing uses the GDAL
library; the list of available drivers for raster data is obtained
by 
```{r eval=FALSE}
st_drivers("raster")
```

### Plotting raster data
We can use the base plot method for `stars` objects, shown in figure \@ref(fig:firststars).
```{r firststars,fig.cap="6 30m Landsat bands downsampled to 90m for Olinda, Br."}
plot(x)
```
The default color scale uses grey tones, and stretches this such
that color breaks correspond to data quantiles over all bands.
A more familiar view is the rgb or false color composite:
```{r starsrgb, out.width = '100%', fig.cap = "two RGB composites", fig = 3, fig.show = 'hold'}
par(mfrow = c(1, 2))
plot(x, rgb = c(3,2,1), reset = FALSE, main = "RGB")    # rgb
plot(x, rgb = c(4,3,2), main = "False color (NIR-R-G)") # false color
```

### Analysing raster data

Element-wise mathematical operations on `stars` objects are just passed
on to the arrays. This means that we can call functions and create
expressions:
```{r}
log(x)
x + 2 * log(x)
```
or even mask out certain values:
```{r}
x2 = x
x2[x < 50] = NA
x2
```
or un-mask areas:
```{r}
x2[is.na(x2)] = 0
x2
```

Dimension-wise, we can apply functions to array dimensions of stars
objects just like `apply` does this to matrices. For instance, to
compute for each pixel the mean of the 6 band values we can do
```{r}
st_apply(x, c("x", "y"), mean)
```
A more meaningful function would e.g. compute the NDVI (normalized
differenced vegetation index):
```{r}
ndvi = function(x) (x[4]-x[3])/(x[4]+x[3])
st_apply(x, c("x", "y"), ndvi)
```
Alternatively, to compute for each band the mean of the whole image
we can do
```{r}
as.data.frame(st_apply(x, c("band"), mean))
```
which is so small it can be printed here as a `data.frame`. In these
two examples, entire dimensions disappear. Sometimes, this does not
happen; we can for instance compute the three quartiles for each band
```{r}
st_apply(x, c("band"), quantile, c(.25, .5, .75))
```
and see that this _creates_ a new dimension, `quantile`, with three values.
Alternatively, the three quantiles over the 6 bands for each pixel are
obtained by
```{r}
st_apply(x, c("x", "y"), quantile, c(.25, .5, .75))
```

### Raster varieties: rectilinear, curvilinear

Besides the regular raster with square cells and axes aligned with
$x$ and $y$, several other raster types exist. The ones supported
by package `stars` are shown in figure \@ref(fig:rastertypes).

```{r rastertypes, echo=FALSE,fig.cap="raster types supported by the stars package"}
x = 1:5
y = 1:4
d = st_dimensions(x = x, y = y, .raster = c("x", "y"))
m = matrix(runif(20),5,4)
r1 = st_as_stars(r = m, dimensions = d)

r = attr(d, "raster")
r$affine = c(0.2, -0.2)
attr(d, "raster") = r
r2 = st_as_stars(r = m, dimensions = d)

r = attr(d, "raster")
r$affine = c(0.1, -0.3)
attr(d, "raster") = r
r3 = st_as_stars(r = m, dimensions = d)

x = c(1, 2, 3.5, 5, 6)
y = c(1, 1.5, 3, 3.5)
d = st_dimensions(x = x, y = y, .raster = c("x", "y"))
r4 = st_as_stars(r = m, dimensions = d)

grd = st_make_grid(cellsize = c(10,10), offset = c(-130,10), n= c(8,5), crs=st_crs(4326))
r5 = st_transform(grd, "+proj=laea +lon_0=-70 +lat_0=35")

par(mfrow = c(2,3), mar = c(0.1, 1, 1.1, 1))
r1 = st_make_grid(cellsize = c(1,1), n = c(5,4), offset = c(0,0))
plot(r1, main = "regular")
plot(st_geometry(st_as_sf(r2)), main = "rotated")
plot(st_geometry(st_as_sf(r3)), main = "sheared")
plot(st_geometry(st_as_sf(r4, as_points = FALSE)), main = "rectilinear")
plot(st_geometry((r5)), main = "curvilinear")
```
The [data
model](https://r-spatial.github.io/stars/articles/data_model.html)
vignette of the package explains the models in detail, and points
out how they can be constructed.

There are several reasons why non-regular rasters occur. For one,
when the data is Earth-bound, a regular raster does not fit the Earth
surface, which is curved. Other reasons are:

* when we convert or transform a regular raster data into another coordinate reference system,
it will become curvilinear unless we resample; resampling always
goes at the cost of some loss of data and is not reversible.
* observation may lead to irregular rasters; e.g. for satellite swaths, we
may have a regular raster in the direction of the satellite (not
aligned with $x$ or $y$), and rectilinear perpendicular to that
(e.g. if the sensor discretizes the viewing _angle_ in equal sections)

### Handling large raster datasets

A common challenge with raster datasets is not only that they come
in large files (single Sentinel-2 tiles are around 1 Gb), but that
many of these files, potentially thousands, are needed to address
the area and time period of interest. At time of
writing this, the Copernicus program which runs all Sentinel
satellites publishes 160 Tb of images per day.  This means that a
classic pattern in using R, consisting of

* downloading data to local disc, 
* loading the data in memory, 
* analysing it

is not going to work.

Cloud-based Earth Observation processing platforms like Google Earth
Engine [@gorelick] or [Sentinel Hub](https://www.sentinel-hub.com/)
recognize this and let users work with datasets up to 20 petabyte
rather easily and with a great deal of interactivity. They share
the following properties:

* computations are posponed as long as possible (lazy evaluation),
* only the data you ask for are being computed and returned, and nothing more,
* storing intermediate results is avoided in favour of on-the-fly computations,
* maps with useful results are generated and shown quickly to allow for interactive model development.

This is similar to the `dbplyr` interface to databases
and cloud-based analytics environments, but differs in the aspect of
_what_ we want to see quickly: rather than the first $n$ records,
we want a quick _overview_ of the results, in the form of a map
covering the whole area, or part of it, but at screen resolution
rather than native (observation) resolution.

If for instance we want to "see" results for the United States on
screen with 1000 x 1000 pixels, we only need to compute results
for this many pixels, which corresponds roughly to data
on a grid with 3000 m x 3000 m grid cells.  For Sentinel-2
data with 10 m resolution, this means we can subsample with
a factor 300, giving 3 km x 3 km resolution.  Processing,
storage and network requirements then drop a factor $300^2 \approx 10^5$, compared
to working on the native 10 m x 10 m resolution. On the platforms
mentioned, zooming in the map triggers further computations on a
finer resolution and smaller extent.

A simple optimisation that follows these lines is how stars' plot
method works:  in case of plotting large rasters, it subsamples
the array before it plots, drastically saving time.  The degree
of subsampling is derived from the plotting region size and the
plotting resolution (pixel density). For vector devices, such as pdf,
R sets plot resolution to 75 dpi, corresponding to 0.3 mm per pixel.
Enlarging plots may reveal this, but replotting to an enlarged
devices will create a plot at target density.

### `stars` proxy objects

To handle datasets that are too large to fit in memory, `stars`
provides `stars_proxy` objects.  To demonstrate its use, we will
use the `starsdata` package, an R data package with larger datasets
(around 1 Gb total). It can be installed by
```{r eval=FALSE}
install.packages("starsdata", repos = "http://pebesma.staff.ifgi.de", type = "source")
```
We can "load" a Sentinel-2 image from it by
```{r}
granule = system.file("sentinel/S2A_MSIL1C_20180220T105051_N0206_R051_T32ULE_20180221T134037.zip", package = "starsdata")
file.size(granule)
base_name = strsplit(basename(granule), ".zip")[[1]]
s2 = paste0("SENTINEL2_L1C:/vsizip/", granule, "/", base_name, ".SAFE/MTD_MSIL1C.xml:10m:EPSG_32632")
(p = read_stars(s2, proxy = TRUE))
object.size(p)
```
and we see that this does not actually load _any_ of the pixel
values, but keeps the reference to the dataset and fills the
dimensions table. (The convoluted `s2` name is needed to point
GDAL to the right file inside the `.zip` file containing 115 files
in total).

The idea of a proxy object is that we can build expressions like
```{r}
p2 = p * 2
```
but that the computations for this are postponed. Only when we
really need the data, e.g. because we want to plot it, is `p *
2` evaluated.  We need data when either

* we want to `plot` data, or
* we want to write an object to disk, with `write_stars`, or
* we want to explicitly load an object in memory, with `st_as_stars`

In case the entire object does not fit in memory, `plot` and
`write_stars` choose different strategies to deal with this:

* `plot` fetches only the pixels that can be seen, rather than all
pixels available, and
* `write_stars` reads, processes, and writes data chunk by chunk.

Downsampling and chunking is implemented for spatially dense images,
not e.g. for dense time series, or other dense dimensions.

As an example,
```{r}
plot(p)
```
only fetches the pixels that can be seen on the plot device, rather
than the 10980 x 10980 pixels available in each band. The downsampling
ratio taken is
```{r}
floor(sqrt(prod(dim(p)) / prod(dev.size("px"))))
```
meaning that for every 19 x 19 sub-image in the original image,
only one pixel is read, and plotted. This value is still a bit too
high as it ignores the white space and space for the key on the
plotting device.

### Operations on proxy objects

A few dedicated methods are available for `stars_proxy` objects:
```{r}
methods(class = "stars_proxy")
```
We have seen `plot` and `print` in action; `dim` reads out
the dimension from the dimensions metadata table. 

The three methods that actually fetch data are `st_as_stars`,
`plot` and `write_stars`.  `st_as_stars` reads the actual data into a
`stars` object, its argument `downsample` controls the downsampling
rate. `plot` does this too, choosing an appropriate `downsample`
value from the device resolution, and plots the object. `write_stars`
writes a `star_proxy` object to disc.

All other methods for `stars_proxy` objects do not actually operate
on the raster data but add the operations to a _to do_ list,
attached to the object. Only when actual raster data are fetched,
e.g. by calling `plot` or `st_as_stars`, the commands in this list
are executed.

`st_crop` limits the extent (area) of the raster that will be
read. `c` combines `stars_proxy` objects, but still doesn't read
any data. `adrop` drops empty dimensions, `aperm` changes dimension
order.

`write_stars` reads and processes its input chunk-wise; it has an
argument `chunk_size` that lets users control the size of spatial
chunks.

## Vector Datacubes {#datacubes}

Data cubes are multi-dimensional array data, where array dimensions
are meaningfully related to categorical or continuous variables
that may include space and time [@lu2018multidimensional]. We have
seen raster data cubes so far, e.g.

* raster data naturally fit in two-dimensional arrays, 
* multi-spectral raster data fit in three-dimensional arrays (cubes), and 
* time series of multi-spectral raster data fit in four-dimensional arrays (hyper-cubes).

Besides Earth Observation/satellite imagery data, a large class
of datacubes come from modelling data, e.g. from oceanographic,
meteorologic or climate models, where dimensions may include

* latitude and longitude
* altitude, or depth
* pressure level (substituting altitude)
* time 
* time to forecast, in addition to time when a forecast was made

we can add to this as an additional dimension

* variable of interest (pressure, temperature, humidity, wind speed, salinity, ...)

when we accept that categorical variables also "take" a dimension. The
alternative would be to consider these as "fields", or "attributes"
of array records. Being able to swap dimensions to
attributes flexibly and vice-versa leads to powerful analysis, as e.g. shown
by the powerful array database SciDB [@brown2010overview].

We go from raster data cubes to _vector data cubes_ if we replace
the two or three raster dimensions with one dimension listing a
set of feature geometries (points, lines or polygons). One example
would be air quality data, where we could have $PM_{10}$ measurements
for

* a set of monitoring stations, and
* a sequence of time intervals

aligned in a vector data cube.  Another example would be demographic
or epidemiological data, where we have a time series of (population,
disease) counts, with number of persons

* by region, for $n$ regions
* by age class, for $m$ age classes, and
* by year, for $p$ years.

which forms an array with $n m p$ elements. 

R has strong native support for arbitrarily dimensioned arrays, and we can get
the value for year $i$, age class $j$ and year $k$ from array `a` by
```{r eval=FALSE}
a[i,j,k]
```
and e.g. the sub-array for age class $j$ by
```{r eval=FALSE}
a[,j,]
```

Thinking along the classical GIS
lines, where we would have either raster or vector data, one is left
with the question what to do when we have a raster time series data
cube (e.g.  a climate model forecast) and want to obtain a vector
time series data cube with aggregates of the model forecast over
polygons, as time series.  For spatial data science, support of
vector and raster data cubes is extremely useful, because many
variables are both spatially and temporaly varying, and because we
often want to either change dimensions or aggregate them out, but
in a fully flexible manner and order. Examples of changing dimensions are

* interpolating air quality measurements to values on a regular grid (raster)
* estimating density maps from points or lines, e.g. with the number of flights passing by per week within a range of 1 km
* aggregating climate model predictions to summary indicators for administrative regions
* combining Earth observation data from different sensors, e.g. Modis (250 m pixels, every 16 days) with Sentinel-2 (10 m,
every 5 days).

Examples of aggregating one ore more full dimensions are assessments of

* which air quality monitoring stations indicate unhealthy conditions
* which region has the highest increase in disease incidence
* global warming (e.g. in degrees per year)

### Example: aggregating air quality time series

Air quality data from package `spacetime` were
obtained from the [airBase European air quality data
base](https://www.eea.europa.eu/data-and-maps/data/aqereporting-8).
Downloaded were daily average PM$_{10}$ values for rural background
stations in Germany, 1998-2009.  

We can create a `stars` object from the `air` matrix, the `dates`
Date vector and the `stations` `SpatialPoints` objects by

```{r}
library(spacetime)
data(air) # this loads several datasets in .GlobalEnv
dim(air)
d = st_dimensions(station = st_as_sfc(stations), time = dates)
(aq = st_as_stars(list(PM10 = air), dimensions = d))
```
We can see from figure \@ref(fig:airst) that the time series are
quite long, but also have large missing value gaps.
Figure \@ref(fig:airmap) shows the spatial distribution measurement stations and
mean PM$_{10}$ values.

```{r airst, fig.cap="space-time diagram of PM$_{10}$ measurements by time and station", echo=TRUE, eval=TRUE, fig=TRUE}
image(aperm(log(aq), 2:1), main = "NA pattern (white) in PM10 station time series")
```

```{r airmap, fig.cap="locations of PM$_{10}$ measurement stations, showing mean values", echo=TRUE, eval=TRUE, fig=TRUE}
plot(st_as_sf(st_apply(aq, 1, mean, na.rm = TRUE)), reset = FALSE, pch = 16,
	ylim = st_bbox(DE)[c(2,4)])
plot(DE, add=TRUE)
```

We can now aggregate these station time series to area means,
mostly as a simple exercise.  For this, we use the `aggregate`
method for `stars` objects

```{r}
(a = aggregate(aq, st_as_sf(DE_NUTS1), mean, na.rm = TRUE))
```

and we can now for instance show the maps for six arbitrarily chosen days
(figure \@ref(fig:airagg)),

```{r airagg, fig.cap="areal mean PM$_{10}$ values, for six days"}
library(tidyverse)
a %>% filter(time >= "2008-01-01", time < "2008-01-07") %>% plot(key.pos = 4)
```
or a time series of mean values for a single state 
(figure \@ref(fig:airts)).
```{r airts, fig.cap="areal mean PM$_{10}$ values, for six days"}
suppressPackageStartupMessages(library(xts))
plot(as.xts(a)[,4], main = DE_NUTS1$NAME_1[4])
```

### Example: Bristol origin-destination datacube

The data used for this example come from [@geocomp], and concern
origin-destination (OD) counts: the number of persons going from
region A to region B, by transportation mode.  We have feature
geometries for the 102 origin and destination regions, shown in figure \@ref(fig:bristol1).
```{r bristol1, fig.cap="Origin destination data zones for Bristol, UK, with zone 33 (E02003043) colored red"}
library(spDataLarge)
plot(st_geometry(bristol_zones), axes = TRUE, graticule = TRUE)
plot(st_geometry(bristol_zones)[33], col = 'red', add = TRUE)
```
and the OD counts come in a table
with OD pairs as records, and transportation mode as variables:
```{r}
head(bristol_od)
```
We see that many combinations of origin and destination are implicit
zeroes, otherwise these two numbers would have been the same:
```{r}
nrow(bristol_zones)^2
nrow(bristol_od)
```

We will form a three-dimensional vector datacube with origin,
destination and transportation mode as dimensions. For this, we
first "tidy" the `bristol_od` table to have origin (o), destination (d),
transportation mode (mode), and count (n) as variables, using `gather`:
```{r}
# create O-D-mode array:
bristol_tidy <- bristol_od %>% select(-all) %>% gather("mode", "n", -o, -d)
head(bristol_tidy)
```
Next, we form the three-dimensional array `a`, filled with zeroes:
```{r}
od = bristol_tidy %>% pull("o") %>% unique
nod = length(od)
mode = bristol_tidy %>% pull("mode") %>% unique
nmode = length(mode)
a = array(0L,  c(nod, nod, nmode), 
	dimnames = list(o = od, d = od, mode = mode))
```
We see that the dimensions are named with the zone names (o, d)
and the transportation mode name (mode). 
Every row of `bristol_tidy` denotes an array entry, and we can
use this to to fill the non-zero entries of the `bristol_tidy` table
with their appropriate value (`n`):
```{r}
a[as.matrix(bristol_tidy[c("o", "d", "mode")])] = bristol_tidy$n
```

To be sure that there is not an order mismatch between the zones
in `bristol_zones` and the zone names in `bristol_tidy`, we can
get the right set of zones by:
```{r}
order = match(od, bristol_zones$geo_code) # it happens this equals 1:102
zones = st_geometry(bristol_zones)[order]
```
(It happens that the order is already correct, but it is good
practice to not assume this).

Next, with zones and modes we can create a stars dimensions object:
```{r}
library(stars)
(d = st_dimensions(o = zones, d = zones, mode = mode))
```

and finally build or stars object from `a` and `d`:
```{r}
(odm = st_as_stars(list(N = a), dimensions = d))
```

We can take a single slice through from this three-dimensional
array, e.g. for zone 33 (figure \@ref(fig:bristol1)), by `odm[,,33]`, 
and plot it:
```{r}
plot(odm[,,33] + 1, logz = TRUE)
```
Subsetting this way, we take all attributes (there is only one: N)
since the first argument is empty, we take all origin regions (second
argument empty), we take destination zone 33 (third argument),
and all transportation modes (fourth argument empty, or missing).

Why plotted this particular zone because it has the most travelers
as its destination. We can find this out by summing all origins and
travel modes by destination:
```{r}
d = st_apply(odm, 2, sum)
which.max(d[[1]])
```

Other aggregations we can carry out include: total transportation
by OD (102 x 102):
```{r}
st_apply(odm, 1:2, sum)
```
Origin totals, by mode:
```{r}
st_apply(odm, c(1,3), sum)
```
Destination totals, by mode:
```{r}
st_apply(odm, c(2,3), sum)
```
Origin totals, summed over modes:
```{r}
o = st_apply(odm, 1, sum)
```
Destination totals, summed over modes (we had this):
```{r}
d = st_apply(odm, 2, sum)
```
We take `o` and `d` together and plot them by
```{r}
x = (c(o, d, along = list(od = c("origin", "destination"))))
plot(x, logz = TRUE)
```

There is something to say for the argument that such maps
give the wrong message, as both amount (color) and polygon
size give an impression of amount. To take out the amount
in the count, we can compute densities (count / km$^2$), by

```{r}
library(units)
a = as.numeric(set_units(st_area(st_as_sf(o)), km^2))
dens_o = o / a
dens_d = d / a
plot(c(dens_o, dens_d, along = list(od = c("origin", "destination"))), logz = TRUE)
```


### Are datacubes tidy?

Yes! The _tidy data_ paper [@tidy] may suggest that such array
data should be processed not as an array, but in a long table where
each row holds (region, class, year, value), and it is always good
to be able to do this. For primary handling and storage however,
this is often not an option, because

* a lot of array data are collected or generated as array data, e.g.
by imagery or other sensory devices, or e.g. by climate models
* it is easier to derive the long table form from the array than
vice versa
* the long table form requires much more memory, since the space
occupied by dimension values is $O(nmp)$, rather than $O(n+m+p)$
* when missing-valued cells are dropped, the long table form loses
the implicit indexing of the array form

To put this argument to the extreme, consider for instance that
all image, video and sound data are stored in array form; few
people would make a real case for storing them in a long table
form instead.  Nevertheless, R packages like `tsibble` take this
approach, and have to deal with ambiguous ordering of multiple records
with identical time steps for different spatial features and index
them, which is solved for both _automatically_ by using the array form.

Package `stars` tries to follow the [tidy
manifesto](https://cran.r-project.org/web/packages/tidyverse/vignettes/manifesto.html)
to handle array sets, and has particularly developed support for the
case where one or more of the dimensions refer to space, and/or time.

## Exercises 

1. NDVI, normalized differenced vegetation index, is coputed as (NIR-R)/(NIR+R), with NIR the near infrared and R the red band.  Read the `L7_ETMs.tif` file into object `x`, and distribute the band dimensions over attributes by `split(x, "band")`. Then, compute NDVI by using an expression that uses the NIR (band 4) and R (band 3) attributes directly.
1. Compute NDVI for the S2 image, using `st_apply` and an a function `ndvi = function(x) (x[4]-x[3])/(x[4]+x[3])`. Plot the result, and write the result to a GeoTIFF. Explain the difference in runtime between plotting and writing.
1. Use `st_transform` to transform the `stars` object read from `L7_ETMs.tif` to EPSG 4326. Print the object. Is this a regular grid? Plot the first band using arguments `axes=TRUE` and `border=NA`, and explain why this takes such a long time.
1. Use `st_warp` to warp the `L7_ETMs.tif` object to EPSG 4326, and plot the resulting object with `axes=TRUE`. Why is the plot created much faster than after `st_transform`?
