# Spherical Geometries {#spherical}

The previous chapter discussed geometries defined on the plane,
$R^2$.  This chapter discusses what changes when we consider
geometries not on the plane, but on the sphere ($S^2$).

Although we learned in chapter \@ref(cs) that the shape of the Earth
is usually approximated by an ellipsoid, none of the libraries shown
in green in figure \@ref(fig:gdal-fig-nodetails) provide access
to a comprehensive set of functions using ellipsoidal geometries,
whereas the s2geometry [@R-s2; @s2geometry] library does provide it
using a sphere rather than an ellipsoid. When compared to a flat
(projected) space assumed in the previous chapter, a sphere is a
much better approximation to an ellipsoid.

## Straight lines

The basic premise of _simple features_ of chapter \@{geometries}
is that geometries are represented by sequences of points _connected
by straight lines_.  On $R^2$ (or any Cartesian space), this is is
trivial, but on a sphere straightlines do not exist. The shortest
line connecting two points is an arc of the circle through both
points and the center of the sphere, also called a _great circle
segment_. A consequence is that "the" shortest distance line
connecting two points on opposing sides of the sphere does not exist,
as any great circle segment connecting them has equal length.

## Ring direction

Any polygon on the sphere divides the sphere surface in two parts
with finite area: the inside and the outside. Using the "counter
clockwise rule" as was done for $R^2$ will not work, because the
direction interpretation depends on what is defined as inside. A
way out is to define the inside as the left (or right) side of the
polygon boundary when traversing its points in sequence. Reversal
of the node order then switches inside and outside.

## Full polygon

In addition to empty polygons, on a sphere one can define the
full polygon, which comprises its entire surface. This is useful,
for instance for computing the oceans as the geometric difference
between the full polygon and those of the land mass.

## Bounding box, rectangle, and cap

Where in $R^2$ one can easily define bounding boxes as the range
of the $x$ and $y$ coordinates, for geographic coordinates these
ranges are not of much use when geometries cross the antimeridian
(longitude +/- 180) or one of the poles. The assumption in $R^2$
that lower $x$ values are Westwards of higher ones does not hold
when crossing the antimeridian. An alternative to delineating
an area on a sphere that is more natural is the _bounding cap_,
defined by its center coordinates and a radius. For Antarctica,
as depicted in figure \@ref(fig:antarctica) (a) and (c), the
bounding box formed by coordinate ranges is
```{r caprect,echo=FALSE}
library(sf)
suppressPackageStartupMessages(library(maps))
# maps:
par(mfrow = c(2,2))
par(mar = c(1,1.2,1,1))
m = st_as_sf(map(fill=TRUE, plot=FALSE))
a = m[m$ID == "Antarctica", ]
st_bbox(a)
```
<details>
<summary style="color:blue;">Click for Code</summary>
```{r eval=FALSE}
library(sf)
suppressPackageStartupMessages(library(maps))
# maps:
par(mfrow = c(2,2))
par(mar = c(1,1.2,1,1))
m = st_as_sf(map(fill=TRUE, plot=FALSE))
a = m[m$ID == "Antarctica", ]
st_bbox(a)
```
</details>
which clearly does not contain the region (`ymin` being -90 and `xmax` 180).
Two geometries that do contain the region are the bounding cap:
```{r echo=FALSE}
library(s2)
s2_bounds_cap(a)
```
<details>
<summary style="color:blue;">Click for Code</summary>
```{r eval=FALSE}
library(s2)
s2_bounds_cap(a)
```
</details>
and the bounding _rectangle_:
```{r echo=FALSE}
s2_bounds_rect(a)
```
<details>
<summary style="color:blue;">Click for Code</summary>
```{r eval=FALSE}
s2_bounds_rect(a)
```
</details>
For an area spanning the antimeridian, here the Fiji island country,
the bounding box
```{r echo=FALSE}
st_bbox(m[m$ID == "Fiji",])
```
<details>
<summary style="color:blue;">Click for Code</summary>
```{r eval=FALSE}
st_bbox(m[m$ID == "Fiji",])
```
</details>
seems to span most of the Earth, as opposed to the bounding rectangle
```{r echo=FALSE}
s2_bounds_rect(m[m$ID == "Fiji",])
```
<details>
<summary style="color:blue;">Click for Code</summary>
```{r eval=FALSE}
s2_bounds_rect(m[m$ID == "Fiji",])
```
</details>
where a value `lng_lo` _larger_ than `lng_hi` indicates that the
bounding rectangle spans the antimeridian. This property could not
be infered from the coordinate ranges (bounding box) themselves.

## Validity on the sphere

Many global datasets are given in geographic coordinates but are
prepared in a way that they "work" when interpreted on the $R^2$
space [-180,180] $\times$ [-90,90]. This means that

* geometries crossing the antimeridian (longitude +/- 180) are cut in
halves, such that they no longer cross it (but nearly touch eachother)
* geometries including a pole, like Antarctica, are cut at +/- 180 and
make an excursion through -180,-90 and 180,90 (both representing the
Geographic South Pole). 

Figure \@ref(fig:antarctica) shows two different representation of
Antarctica, plotted with geographical coordinates taken as $R^2$
(top) and in a Polar Stereographic projection (bottom), without
(left) and with (right) en excursion through the Geographic South
Pole. In the projections as plotted, polygons polygons (b) and
(c) are valid; polygon (a) is not valid as it self-intersects,
polygon (d) is not valid because it traverses the same edge to the
South Pole twice.  On the sphere ($S^2$), polygon (a) is valid but
(b) is not, for the same reason as (d) is not valid.

```{r antarctica,echo=FALSE,fig.cap="different representations of Antarctica, (a, c): with a polygon _not_ passing through (-180 -90); (b, d): with a polygon passing through (-180 -90) and (180 -90)"}
library(sf)
suppressPackageStartupMessages(library(maps))
# maps:
par(mfrow = c(2,2))
par(mar = c(1,1.2,1,1))
m = st_as_sf(map(fill=TRUE, plot=FALSE))
m = m[m$ID == "Antarctica", ]
plot(st_geometry(m), asp = 2)
title("a (not valid)")
# ne:
library(rnaturalearth)
ne = ne_countries(returnclass = "sf")
ne = ne[ne$region_un == "Antarctica", "region_un"]
plot(st_geometry(ne), asp = 2)
title("b (valid)")
# 3031
m %>%
  st_geometry() %>%
  st_transform(3031) %>%
  plot()
title("c (valid)")
ne %>%
  st_geometry() %>%
  st_transform(3031) %>%
  plot()
title("d (not valid)")
```
<details>
<summary style="color:blue;">Click for Code</summary>
```{r eval=FALSE}
library(sf)
suppressPackageStartupMessages(library(maps))
# maps:
par(mfrow = c(2,2))
par(mar = c(1,1.2,1,1))
m = st_as_sf(map(fill=TRUE, plot=FALSE))
m = m[m$ID == "Antarctica", ]
plot(st_geometry(m), asp = 2)
title("a (not valid)")
# ne:
library(rnaturalearth)
ne = ne_countries(returnclass = "sf")
ne = ne[ne$region_un == "Antarctica", "region_un"]
plot(st_geometry(ne), asp = 2)
title("b (valid)")
# 3031
m %>%
  st_geometry() %>%
  st_transform(3031) %>%
  plot()
title("c (valid)")
ne %>%
  st_geometry() %>%
  st_transform(3031) %>%
  plot()
title("d (not valid)")
```
</details>
