# Attributes, Up- and Downscaling {#featureattributes}

Feature _attributes_ refer to the properties of features ("things")
that do not describe the feature's geometry. Feature attributes can
be _derived_ from geometry (e.g. length of a `LINESTRING`, area
of a `POLYGON`) but they can also refer to completely different
properties, such as 

* the name of a street or a county, 
* the number of people living in a country, 
* the type of a road
* the soil type in a polygon from a soil map.
* the opening hours of a shop
* the body weight or heart beat of an animal
* the NO$_2$ concentration measured at an air quality monitoring station

Although temporal properties of features are no less fundamental
than their spatial properties, the _simple feature access_ standard
does not give time a similar role as space; more on that in chapter
\@ref(raster).

Spatial data science software implementing simple features typically
organizes data in tables that contain both geometries and attributes
for features; this is true for `geopandas` in Python, `PostGIS`
tables in PostgreSQL, and `sf` objects in R.  The geometric
operations described in the previous chapter (\@ref(opgeom))
operate on geometries _only_, and may occasionally yield attributes
(predicates or measures), but will not modify attributes present.

In all these cases, while manipulating geometries, attribute _values_
remain unmodified. At first sight, that looks rather harmless. But
if we look into a simple case of replacing a county boundary with
a county centroid, we may see a warning is emitted:
```{r countycentroid,echo=FALSE}
library(sf)
library(dplyr)
system.file("gpkg/nc.gpkg", package="sf") %>%
	read_sf() %>%
	st_transform(32119) %>%
	select(BIR74, SID74, NAME) %>%
	st_centroid() %>%
	head(n = 1) -> x # save as x
st_geometry(x)[[1]]
```
<details>
<summary style="color:blue;">Click for Code</summary>
```{r eval=FALSE}
library(sf)
library(dplyr)
system.file("gpkg/nc.gpkg", package="sf") %>%
	read_sf() %>%
	st_transform(32119) %>%
	select(BIR74, SID74, NAME) %>%
	st_centroid() %>%
	head(n = 1) -> x # save as x
st_geometry(x)[[1]]
```
</details>
This warning is justified for the first two variables shown (total
births and number of SID disease cases, 1974) which, as such, are
_not associated with_ a feature whose geometry is `POINT (385605.4
300303.5)`. The third variable, `NAME` is however still the county
name for the point indicated, but the point geometry no longer _is_
the county geometry.

In section \@ref(support) we already described that for non-point
geometries, feature attribute values may either have _point support_,
meaning that the value applies to _every point_, or they may have
_block support_, meaning that the value _summarizes all points_
in the geometry. More complex options, e.g. inbetween these two
extremes, may also occur. This chapter will describe different
ways in which an attribute may relate to the geometry, its
consequences on analysing such data, and ways to derive attribute
data for different geometries (up- and downscaling).

## Attribute-geometry relationships {#agr}

Changing the feature geometry without changing the feature attributes
does change the _feature_, since the feature is characterised by
the combination of geometry and attributes. Can we, ahead of time,
predict whether the resulting feature will still meaningfully relate
to the attribute data when we replace all geometries for instance
with their convex hull or centroid? It depends.

Take the example of a road, represented by a `LINESTRING`, which has
an attribute property _road width_ equal to 10 m. What can we say about
the road width of an arbitray subsectin of this road? That depends
on whether the attribute road length describes, for instance the
road width everywhere, meaning that road width is constant along the
road, or whether it describes an aggregate property, such as minimum
or average road width.  In case of the minimum, for an arbitrary
subsection of the road one could still argue that the minimum
road width must be at least as large as the minimum road width for
the whole segment, but it may no longer be _the minimum_ for that
subsection. This gives us two "types" for the attribute-geometry
relationship (**AGR**):

* **constant** the attribute value is valid everywhere in or over the geometry
* **aggregate** the attribute is an aggregate, a summary value over the geometry

For polygon data, typical examples of **constant** AGR are

* land use for a land use polygon
* rock units or geologic strata in a geological map
* soil type in a soil map
* elevation class in a elevation map that shows elevation as classes
* climate zone in a climate zone map

Typical examples for the **aggregate** AGR are

* population, either as number of persons or as population density
* other socio-economic variables, summarised by area
* total emission of pollutants by region
* block mean NO$_2$ concentrations, as e.g. obtained by block kriging or a dispersion model that predicts areal means

A third type of AGR is that where an attribute identifies a feature
geometry. The example above is county `NAME`: the name identifies
the county, and is still the county `NAME` for any sub-area. 

* **identity** the attribute value uniquely identifies the geometry
as a whole, there are no other geometries with the same value

For arbitrary sub-areas, the attribute will lose the **identity**
property but becomes a **constant** attribute. An example is:

* any point inside a county is still part of the county and must have
the same value for county name, but it does not longer represent the
(entire) geometry corresponding to that county.


The challenge here is that spatial information (ignoring time for
simplicity) belongs to different phenomena types [e.g. @scheider2016],
including

* **fields**: where over _continuous_ space, every location corresponds to a single value, e.g. elevation
* **objects**: found a _discrete_ set of locations, e.g. houses or persons
* **aggregations**: e.g. sums, totals, averages of fields, or spatial densities of objects 

but that different spatial geometry types (points, lines, polygons,
raster cells) do not correspond to either of these:

* points may refer to samples of fields (air quality) or locations of objects
* lines may be used for objects (roads) or contours of a field
* raster pixels and polygons may reflect fields of a categorical
variable such as land use (_coverage_), but also aggregates such
as population density

Properly specifying attribute-geometry relationships, and warning
against their absence or cases when change of geometry (support)
implies a change of information can help avoiding a large class of
common spatial data analysis mistakes [@stasch2014] associated with
the support of spatial data.

## Spatial join

In regular (left, right or inner) joins, _joined_ records from a
pair of tables are reported when one or more selected attributes
match (are identical) in both tables. A spatial join is similar,
but the criterion to join records is not equality of attributes but
a spatial predicate. This leaves a wide variety of options in order
to define _spatially_ matching records, using binary predicates
listed in section \@ref(de9im). The concepts of "left", "right",
"inner" or "full" joins remain identical to the non-spatial join
as the options for handling records that have no spatial match.

When using spatial joins, each record may have many matched records,
yielding a large result table. A way to reduce this complexity may
be to select the matching record with the largest overlap with the
target geometry.  An example of this is shown (visually) in figure
\@ref(fig:largest).

(ref:foo) example of `st_join` with `largest = TRUE`: the label of the polygon in the top figure with the largest intersection with polygons in the bottom figure is assigned to the polygons of the bottom figure

```{r largest, out.width='60%', fig.cap='(ref:foo)', echo=FALSE}
# example of largest = TRUE:
nc <- st_transform(read_sf(system.file("shape/nc.shp", package="sf")), 2264)
gr = st_sf(
         label = apply(expand.grid(1:10, LETTERS[10:1])[,2:1], 1, paste0, collapse = " "),
         geom = st_make_grid(nc))
gr$col = sf.colors(10, categorical = TRUE, alpha = .3)
# cut, to check, NA's work out:
gr = gr[-(1:30),]
suppressWarnings(nc_j <- st_join(nc, gr, largest = TRUE))
# the two datasets:
opar = par(mfrow = c(2,1), mar = rep(0,4))
plot(st_geometry(nc_j))
plot(st_geometry(gr), add = TRUE, col = gr$col)
text(st_coordinates(st_centroid(st_geometry(gr))), labels = gr$label)
# the joined dataset:
plot(st_geometry(nc_j), border = 'black', col = nc_j$col)
text(st_coordinates(st_centroid(st_geometry(nc_j))), labels = nc_j$label, cex = .8)
plot(st_geometry(gr), border = 'green', add = TRUE)
par(opar)
```
<details>
<summary style="color:blue;">Click for Code</summary>
```{r eval=FALSE}
# example of largest = TRUE:
nc <- st_transform(read_sf(system.file("shape/nc.shp", package="sf")), 2264)
gr = st_sf(
         label = apply(expand.grid(1:10, LETTERS[10:1])[,2:1], 1, paste0, collapse = " "),
         geom = st_make_grid(nc))
gr$col = sf.colors(10, categorical = TRUE, alpha = .3)
# cut, to check, NA's work out:
gr = gr[-(1:30),]
suppressWarnings(nc_j <- st_join(nc, gr, largest = TRUE))
# the two datasets:
opar = par(mfrow = c(2,1), mar = rep(0,4))
plot(st_geometry(nc_j))
plot(st_geometry(gr), add = TRUE, col = gr$col)
text(st_coordinates(st_centroid(st_geometry(gr))), labels = gr$label)
# the joined dataset:
plot(st_geometry(nc_j), border = 'black', col = nc_j$col)
text(st_coordinates(st_centroid(st_geometry(nc_j))), labels = nc_j$label, cex = .8)
plot(st_geometry(gr), border = 'green', add = TRUE)
par(opar)
```
</details>

Another way to reduce the result set is to aggregate, after a join,
all matching records, and union their geometry; this will now be
discussed.

## Aggregating and Summarising

Aggregating records in a table (or `data.frame`) involves two
steps

* grouping records based on a grouping predicate, and
* applying an aggregation function to the attribute values of a
group to summarize them into a single number.

In SQL, this looks for instance like
```
SELECT GroupID, SUM(population) FROM table GROUP BY GroupID;
```
indicating that the aggregation function (`SUM`) and the grouping
predicate (`region`). R package `dplyr` for instance offers two
functions: `group_by` which specifies the group membership of records
and `summarize` which computes data summaries (such as `sum` or
`mean`) for each of the groups. R (base) function `aggregate` does
both in a single function that takes the data table, the grouping
predicate(s) and the aggregation function.

An example for the North Caroline counties is shown in figure
\@ref(fig:ncaggregation). Here, we group counties by their position
(according to the quadrant in which the county centroid is with
respect to geographical coordinate `POINT(-79, 35.5)`) and
count the number of disease cases per group. The result shows that
the geometries of the resulting group have been unioned (section
\@ref(bintrans)), which is necessary because the `MULTIPOLYGON`
formed by just putting all the county geometries together would
have many duplicate boundaries, and hence not be _valid_ (section
\@ref(valid)).

```{r ncaggregation, fig.cap="top: SID74 disease cases by county, bottom: summed values by county quadrant, county polygons unioned by county quadrant", echo=FALSE}
nc <- read_sf(system.file("gpkg/nc.gpkg", package="sf"))
plot(nc["SID74"], axes = TRUE)
sf_use_s2(TRUE)
# encode quadrant by two logicals:
nc$lng = st_coordinates(st_centroid(st_geometry(nc)))[,1] > -79
nc$lat = st_coordinates(st_centroid(st_geometry(nc)))[,2] > 35.5
nc.grp = aggregate(nc["SID74"], list(nc$lng, nc$lat), sum)
plot(nc.grp["SID74"], axes = TRUE)
```
<details>
<summary style="color:blue;">Click for Code</summary>
```{r eval=FALSE}
nc <- read_sf(system.file("gpkg/nc.gpkg", package="sf"))
plot(nc["SID74"], axes = TRUE)
sf_use_s2(TRUE)
# encode quadrant by two logicals:
nc$lng = st_coordinates(st_centroid(st_geometry(nc)))[,1] > -79
nc$lat = st_coordinates(st_centroid(st_geometry(nc)))[,2] > 35.5
nc.grp = aggregate(nc["SID74"], list(nc$lng, nc$lat), sum)
plot(nc.grp["SID74"], axes = TRUE)
```
</details>

Plotting collated county polygons is technically not a problem, but
for this case would raise the wrong suggestion that the group sums
relate to the counties, and not the group of counties. Subsequent
geometrical operations on such invalid polygon collections would
however fail (and require a `union` or `make_valid` first).

One particular property of aggregation in this way is that each
record is assigned to a single group; this has the advantage that
the sum of the group-wise sums equals the sum of the ungrouped data:
for variables that reflect _amount_, nothing gets lost and nothing
is added. The newly formed geometry is the result of unioning the
geometries of the records.

When we need an aggregate for a new area that is _not_ a union
of the geometries for a group of records, and we use a spatial
predicate then we may end up with a case where records are matched to
multiple groups. When taking the rectangles of figure \@ref(largest)
as the target areas, and we sum for each rectangle the disease cases
of the counties that _intersect_ with the rectangle, the sum of these
will be much larger:

```{r echo=FALSE}
nc <- st_transform(nc, 2264)
a = aggregate(nc["SID74"], gr, sum)
c(sid74_sum_counties = sum(nc$SID74), sid74_sum_rectangles = sum(a$SID74, na.rm = TRUE))
```
<details>
<summary style="color:blue;">Click for Code</summary>
```{r eval=FALSE}
nc <- st_transform(nc, 2264)
a = aggregate(nc["SID74"], gr, sum)
c(sid74_sum_counties = sum(nc$SID74), sid74_sum_rectangles = sum(a$SID74, na.rm = TRUE))
```
</details>

Choosing another predicate, e.g. _contains_ or _covers_ would on
the contrary result in much smaller values, because many counties
are not contained by _any_ the target geometries. However, there are
a few cases where this approach might be good or satisfactory:

* when we want to aggregate `POINT` geometries by a set of polygons,
and all points are contained by a polygon. If points fall on a
shared boundary than they are assigned to both polygons (this is
the case for DE-9IM-based GEOS library; the s2geometry library has
the option to define polygons as "semi-open", which implies that
points are assigned to single polygons when the polygons form a
coverage)
* when aggregating many very small polygons or raster pixels over
larger areas, e.g. averaging altitude from 30 m resolution raster
over North Carolina counties, the error made by multiple matches
may be insignificant.

A more comprehensive approach to aggregating spatial data associated
to areas to larger, arbitrary shaped areas is by using area-weighted
interpolation.

## Area-weighted interpolation

When we want to combine geometries and attributes of two datasets
such that we get attribute values of a source dataset summarised for
the geometries of a target, where source and target geometries are
unrelated, area-weighted interpolation may be a simple approach. In
effect, it considers the area of overlap of the source and target
geometries, and uses that to weight the source attribute values into
the target value [@goodchild; @thomas; @do]. Here, we follow the
notation of [@do].

Area-weighted interpolation computes for each of $q$ spatial target
areas $T_j$ a weighted average from the values $Y_i$ corresponding
to the $p$ spatial source areas $S_i$,

\begin{equation}
\hat{Y}_j(T_j) = \sum_{i=1}^p w_{ij} Y_i(S_i)
(\#eq:aw)
\end{equation}

where the $w_{ij}$ depend on the amount of overlap of $T_j$ and
$S_i$, $A_{ij} = T_j \cap S_i$.

Different options exist for choosing weights, including methods
using external variables (e.g. dasymetric mapping, [@mennis]).
Two simple approaches for computing weights that do not use external
variables arise, depending on whether the variable $Z$ is _intensive_
or _extensive_.

### Spatially extensive and intensive variables

An example of an extensive variable is _population count_. It
is associated with an area, and if that area is cut into smaller
areas, the population count is split accordingly: not necessary
proportional to area, because population is rarely uniform, but split
in such a way that the sum of the population count for the smaller
areas equals that of the total.  An example of a related variable
that is _intensive_ is population density. If an an area is split
into smaller areas, population density is not split similarly:
the sum of the population densities for the smaller areas is a
meaningless measure, as opposed to the average of the population
densities which will be similar to the density of the total area.

Extensive variables correspond to amounts, associated with a physical
size (length, area, volume); for spatially extensive variables,
if the area a value corresponds to is cut in parts, the values
associated with the sub-area are split accordingly. In other words:
the value is proportional to the support.  Intensive variables are
variables that do not have values proportional to support: if the
area is split, values may vary but _on average_ remain the same.
The corresponding example of an intensive variable is _population
density_: when we split an area into sub-areas, the sub-areas either
have identical population densities (in case population is uniformly
distributed) or, more realistically, have _varying_ population
densities that by necessity are both higher and lower than the
density of the total area.

When we assume that the extensive variable $Y$ is uniformly
distributed over space, the value $Y_{ij}$, derived from $Y_i$
for a sub-area of $S_i$, $A_{ij} = T_j \cap S_i$ of $S_i$ is

$$\hat{Y}_{ij}(A_{ij}) = \frac{|A_{ij}|}{|S_i|} Y_i(S_i)$$

where $|\cdot|$ denotes the spatial area.
For estimating $Y_j(T_j)$ we sum all the elements over area $T_j$:

\begin{equation}
\hat{Y}_j(T_j) = \sum_{i=1}^p \frac{|A_{ij}|}{|S_i|} Y_i(S_i)
(\#eq:awextensive)
\end{equation}

For an intensive variable, under the assumption that the variable
has a constant value over each area $S_i$, the estimate for a
sub-area equals that of the total,

$$\hat{Y}_{ij} = Y_i(S_i)$$

and we can estimate the value of $Y$ for a new spatial unit $T_j$
by an area-weighted average of the source values:

\begin{equation}
\hat{Y}_j(T_j) = \sum_{i=1}^p \frac{|A_{ij}|}{|T_j|} Y_i(S_i)
(\#eq:awintensive)
\end{equation}

### Dasymetric mapping

Dasymetric mapping distributes variables, such as population,
known at a course spatial aggregation level over finer spatial
units by using other variables that are associated with population
distribution, such as land use, building density, or road density.
The simplest approach to dasymetric mapping is obtained for
extensive variables, where the ratio $|A_{ij}| / |S_i|$ in
\@ref(eq:awextensive) is replaced by the ratio of another extensive
variable $X_{ij}(S_{ij})/X_i(S_i)$, which has to be known for both
the intersecting regions $S_{ij}$ and the source regions $S_i$.
[@do] discuss several alternatives for intensive $Y$ and/or $X$,
and cases where $X$ is known for other areas.

### Aggregation and disaggregation

The application of areal interpolation using \@ref(eq:aw) with
its realisations for extensive \@ref(eq:awextensive) and intensive
\@ref(eq:awintensive) variables allows moving information from any
source area $S_i$ to any target area $T_j$ as long as the two areas
have some overlap. This means that one can go arbitrarily to much
larger units (aggregation) or to much smaller units (disaggregation).
Of course this makes only sense to the extent that the assumptions
hold: over the source regions extensive variables need to be
uniformly distributed and intensive variables need to have constant
value.

The ultimate disaggregation involves retrieving (extracting)
point values from line or area data. For this, we cannot work with
equations \@ref(eq:awextensive) or \@ref(eq:awintensive) because
$|A_{ij}| = 0$ for points, but under the assumption of having a
constant value over the geometry, for intensive variables the value
$Y_i(S_i)$ can be assigned to points as long as all points can be
uniquely assigned to a single source area $S_i$. For polygon data,
this implies that $Y$ needs to be a coverage variable (section
\@ref(coverages)).

In cases where values associated with areas are **aggregate** values
over the area, the assumptions made by area-weighted interpolation
or dasymetric mapping -- uniformity or constant values over the
source areas -- are highly unrealistic. In such cases, these simple
approaches still be reasonable approximations, for instance when

* the source and target area are nearly identical
* the variability inside source units is very small, and the variable
is nearly uniform or constant.

In other cases, results obtained using these methods are merely
consequences of unjustified assumptions. Statistical aggregation
methods that can estimate quantities for larger regions from points
or smaller regions include

* design-based methods, which require that a probability sample is
available from the target region, with known inclusion probabilities, and
* model-based methods, which assume a random field model with spatially
correlated values (block kriging, section \@ref(blockkriging)).

Alternative disaggregation methods include 

* deterministic, smoothing-based approaches: kernel- or spline-based
smoothing methods [@toblerpyc; @martin89]
* statistical, model-based approaches: area-to-area and area-to-point
kriging [@kyriakidis04], see section \@ref(areatopoint)

## Exercises

* Add a variable to the `nc` dataset by `nc$State = "North Carolina"`. Which value should you attach to this variable for the attribute-geometry relationship (agr)?
* Create a new `sf` object from the geometry obtained by `st_union(nc)`, and assign `"North Carolina"` to the variable `State`. Which `agr` can you now assign to this attribute variable?
* Use `st_area` to add a variable with name `area` to `nc`. Compare the `area` and `AREA` variables in the `nc` dataset. What are the units of `AREA`? Are the two linearly related? If there are discrepancies, what could be the cause?
* Is the `area` variable intensive or extensive? Is its agr equal to `constant`, `identity` or `aggregate`?
* Find the name of the county that contains `POINT(-78.34046 35.017)`
* Find the names of all counties with boundaries that touch county `Sampson`.
* List the names of all counties that are less than 50 km away from county `Sampson`.

