# Geometric Manipulations {#geommanip}

Simple feature geometries can be queried for properties, combined
into new geometries, and combinations of geometries can be queried
for properties. This chapter will give an overview of the operations
offered by `sf`.

We can categorise operations in terms of what they take as input,
and what they give as output. In terms of input we have operations
that take

* a single geometry (unary operations)
* a pair of geometries (binary operations)
* a set of geometries (n-ary operations)

and in terms of what is returned, we distinguish

* predicates, returning a logical asserting a certain property is `TRUE`
* measures, returning a numeric value (possibly with measurement unit)
* operations that return a new geometry, or a set of geometries

We will now go through all combinations.

## Unary Operations 

### Unary predicates

`st_is_simple`
`st_is_valid`
`st_is_longlat` 

`st_is` 

`==` `!=` : `st_equals`

### Unary measures
`st_dimension` 
`st_area` 
`st_length` 
`st_geohash`
`st_geometry_type`

### Unary operations returning a geometry

`st_centroid`
`st_buffer`
`st_jitter` 
`st_wrap_dateline`

`st_boundary`
`st_convex_hull`
`st_line_merge` merges, on a per-feature basis, connecting `LINESTRING` elements of a `MULTILINESTRING` into longer `LINESTRING`s.
`st_make_valid`
`st_node`
`st_point_on_surface`
`st_polygonize`
`st_segmentize`
`st_simplify`
`lwgeom::st_split`
`st_transform`
`st_triangulate`
`st_voronoi`

`st_zm`

`st_collection_extract`
`st_cast`

Ops: + - (unary, binary)

Ops: * / (sfg, Matrix)

## Binary Operations 

### Binary predicates

`st_contains`
`st_contains_properly`
`st_covered_by`
`st_covers`
`st_crosses` `st_disjoint` `st_equals` `st_equals_exact` `st_intersects`
`st_is_within_distance` `st_within` `st_touches` `st_overlaps`

### Binary measures

`st_distance`
`st_relate`

### Binary operations returning a geometry

`st_intersection`  `&`
`st_union`  `|`
`st_difference`  '/'
`st_sym_differenc`  '%/%'

## N-ary Operations 

### N-ary operations returning a geometry

For n-ary operations, we have only two, n-ary `st_intersection` and `st_difference`,
which both return geometries, in case of `st_intersection` along with some properties
for each of the geometry.

`st_make_grid`
`st_graticule`

## Precision {#precision}

Geometrical operations, such as finding out whether a certain
point is on a line, may fail when coordinates are represented by
highly precise floating point numbers, such as 8-byte doubles in
R. A remedy might be to limit the precision of the coordinates
before the operation.  For this, a _precision model_ is adopted by
`sf`. It uses a precision value to round coordinates (X, Y, Z and M)
right before they are encoded as well-known binary, and passed on to
the libraries where this may have an effect (GEOS, GDAL, liblwgeom).
We demonstrate this by an R - WKB - R roundtrip.

Rounding can be done by three different options. With a negative
precision value, 8-byte doubles get converted to 4-byte floats and
back again:
```{r}
old = options(digits = 16)
(p = st_sfc(st_point(c(1e6/3, 1e4/3))))
p %>% st_set_precision(-1) %>% st_as_binary() %>% st_as_sfc() %>% `[[`(1)
```
With a positive precision $p$, each coordinate value $c$ is replaced by
$$c' = \mbox{round}(p \cdot c) / p$$
This implies that for instance with a precision of 1000, the number
of decimal placed to round to is 1/1000:
```{r}
p %>% st_set_precision(1000) %>% st_as_binary() %>% st_as_sfc() %>% `[[`(1)
```
With a precision of e.g. 0.001 or 0.05, rounding to the nearest `1/precision`,
i.e. 1000 or 20, is obtained:
```{r}
p %>% st_set_precision(0.001) %>% st_as_binary() %>% st_as_sfc() %>% `[[`(1) # nearest 1000
p %>% st_set_precision(0.05) %>% st_as_binary() %>% st_as_sfc()  %>% `[[`(1) # nearest 20
options(digits = old$digits)
```

In essence, these rounding methods bring the coordinates to
points on a regular grid, which is beneficial for geometric
computations. Of course, it also affects all computations like
areas and distances. Which precision values are best for which
application is often a matter of trial and error. A reproducible
example illustrating the need for setting precision is found
[here](https://github.com/r-spatial/sf/issues/275).

## Warnings for longitude/latitude geometries

In st_centroid.sfc(st_geometry(x), of_largest_polygon = of_largest_polygon) :
  st_centroid does not give correct centroids for longitude/latitude data

How serious is this?
