# Manipulating Geometries {#geommanip}

Simple feature geometries can be queried for properties, combined
into new geometries, and combinations of geometries can be queried
for properties. This chapter will give an overview of the operations
offered by `sf`, entirely focusing on _geometrical_ properties.
The next chapter, \@ref(featureattributes), focuses on the analysis
of non-geometrical feature properties, in relationship to their
geometries. Some of the material in this chapter also appeared as
[@rjsf]. 

Several of the concepts of geometric manipulations were introduced
in chapter \@{geometries}. This chapter gives a complete listing
of all geometries permitted on geometries, illustrating some of them.

We can categorise operations in terms of what they take as input,
and what they give as output. In terms of output we have operations
that give one or more

* predicates: a logical asserting a certain property is `TRUE`,
* measures: a value (e.g. a numeric value with measurement unit), or
* geometries

and in terms of what they operate on, we distinguish operations
that work on

* a single geometry (unary operations)
* pairs of geometries (binary operations)
* sets of geometries (n-ary operations)

Before we will go through all combinations, we make two observations:

* most functions are implemented as methods, and operate equally
on single geometry objects (`sfg`), geometry set objects (`sfc`)
or simple feature (`sf`) objects.
* also for binary and n-ary operations, `sfg` or `sf` objects are
accepted as input, and taken as a set of geometries.

## Predicates

Predicates return a logical, `TRUE`  or `FALSE` value, or a set of those.

### Unary predicates

`st_is_simple` returns whether a geometry is simple:
```{r}
st_is_simple(st_sfc(
	st_point(c(0,1)),
	st_linestring(rbind(c(0,0), c(1,1), c(0,1), c(1,0))))) # self-intersects
```

`st_is_valid` returns whether a geometry is valid
```{r}
st_is_valid(st_sfc(
	st_linestring(rbind(c(1,1), c(1,2))),
	st_linestring(rbind(c(1,1), c(1,1))))) # zero-length
```

`st_is_empty` returns whether a geometry is empty
```{r}
st_is_empty(st_point())
```

`st_is_longlat`  returns whether the coordinate reference system is geographic
\@{cs,rs}:
```{r}
demo(nc, ask = FALSE, echo = FALSE)
st_is_longlat(nc)
nc2 <- st_transform(nc, 3857) # to web Mercator
st_is_longlat(nc2)
st_is_longlat(st_point(0:1))
```

`st_is` is an easy way to check for the simple feature geometry type: 
```{r}
st_is(st_point(0:1), "POINT")
all(st_is(nc, "POLYGON"))
all(st_is(nc, "MULTIPOLYGON"))
```

Equality and inequality of geometries can be checked by `==` or `!=`; it
uses geometric equality, and is insensitive to the order of traversal of
nodes:
```{r}
st_sfc(st_point(0:1), st_point(1:2)) == st_sfc(st_point(0:1))
st_linestring(rbind(c(0,0), c(1,1))) == st_linestring(rbind(c(1,1), c(0,0)))
```
Under the hood, it uses `st_equals`, discussed by the  binary predicates.


### Binary predicates {#de9im}

Binary predicates result in a `TRUE` or `FALSE` value for every
pair of inputs.  For two sets of inputs with $n$ and $m$ geometries
respectively, this
results in an $n \times m$ logical matrix. Because $n$ and/or $m$
may be very large and the predicate matrix typically contains mostly
`FALSE` values, a sparse representation of it, a _sparse geometry
binary predicate_ (`sgbp`) object, is returned by all functions. They
are simply lists of indices of the `TRUE` values in each row:
```{r}
(r <- st_touches(nc2[1:2,], nc2))
str(r)
```

`sgbp` objects have the following methods:
```{r}
methods(class = 'sgbp')
```

For understanding predicates, the dimensionally
extended 9-intersection model (DE-9IM, [@de9im1,
@de9im2]) is adopted, which is explained in more detail on
[Wikipedia](https://en.wikipedia.org/wiki/DE-9IM).  Briefly, it
considers that every geometry has an interior, a boundary and an
exterior. For polygons this is trivial, for points the boundary
is an empty set, for linestrings the boundary is formed by the
end points and the interior by all non end points.  Also, any
geometry has a dimension of 0 (points), 1 (lines) or 2 (polygons)
or non-existent in the case of an empty geometry.

A relationship between two geometries A and B is expressed by
the _dimension_ of the overlap (intersections) of 9 intersections,
formed by the 9 pairs from the interior, boundary and exterior of A,
and the interior, boundary and exterior of B. We can query this relation
by using `st_relate`
```{r}
B = st_linestring(rbind(c(0,0), c(1,0)))
A = st_point(c(0.5, 0)) # halfway the line
st_relate(A, B)
```
In the relationship found, `0FFFFF102`, `F` indicates empty geometries, and
we see from 

* `0FF` that the (interior of the) point has 0-dimensional overlap with the interior of line (i.e., the overlap is a point), and no overlap with the boundary or the exterior of the line,
* `FFF` that the (empty) border of the point has nothing in common with the line, and
* `102` that the exterior of the point (all points except this one) have a 1-dimensional overlap with the interior of the line, a 0-dimensional overlap with the boundary of the line (its end points), and a 2-dimensional overlap with the exterior of the line.

We can query whether a particular relationship holds by giving `st_relate` a pattern. To check for instance whether point A overlaps with an end point of linestring B, we can use
```{r}
st_relate(A, B, pattern = "F0FFFFFFF") %>% as.matrix()
```
In these patterns, `*` can be used for _anything_, and `T` for non-empty (0, 1 or 2). 
The standard relationships below are all expressed as particular query patterns, the
Wikipedia page gives details on the patterns used.

The binary predicates provided by package `sf` are

|predicate|value                                                                      |inverse of|
|---------|---------------------------------------------------------------------------|----------|
|`st_contains`|None of the points of A are outside B| `st_within`|
|`st_contains_properly`|A contains B and B has no points in common with the boundary of A| |
|`st_covers`|No points of B lie in the exterior of A| `st_covered_by`|
|`st_covered_by`|inverse of `st_covers`| |
|`st_crosses`|A and B have some but not all interior points in common| |
|`st_disjoint` |A and B have no points in common| `st_intersects`|
|`st_equals` |A and B are geometrically equal; node order number of nodes may differ; identical to A contains B AND A within B|
|`st_equals_exact` |A and B are geometrically equal, and have identical node order| |
|`st_intersects`|A and B are not disjoint| `st_disjoint`|
|`st_is_within_distance`|A is closer to B than a given distance|
|`st_within`|None of the points of B are outside A| `st_contains`|
|`st_touches`|A and B have at least one boundary point in common, but no interior points|  |
|`st_overlaps`|A and B have some points in common; the dimension of these is identical to that of A and B|  |
|`st_relate`|given a pattern, returns whether A and B adhere to this pattern| |

### N-ary

Higher-order predicates are not supported by special functions.

## Measures

### Unary

Unary measures return a single value that describes a property of
the geometry:

|`function`|returns                                                       |
|----------|--------------------------------------------------------------|
|`st_dimension`|0 for points, 1 for linear, 2 for polygons, `NA` for empty geometries|
|`st_area`|the area for geometries|
|`st_length`|the lengths of linear geometries|
|`st_geohash`|the [geohash](https://en.wikipedia.org/wiki/Geohash) for geometries|
|`st_geometry_type`|the types of a set of geometries|

### Binary

`st_distance` returns the distances between pairs of geometries,
either as a vector with distances between the two first, the two
second, ... pairs, or as a matrix with all pairwise distances. The
result is numeric, or is of class `units` [@units] when distance
units can be derived from the coordinate reference system (chapter
\@{rs}):
```{r}
st_distance(nc[1:3,], nc[2:4,], by_element = TRUE) %>% setNames(NULL)
st_distance(nc[1:3,], nc[2:4,])
```

`st_relate` returns the relation pattern, as explained in section
\@{de9im}, or an `sgbp` object when given a pattern template to
match to.

### N-ary

No higher-order functions returning a measure are available.

## Geometry generating functions

### Unary

Unary operations work on a per-geometry basis, and for each geometry return a new geometry.
None of these functions operate on _more_ than one feature geometry.
Most functions are implemented as (S3) generic, with methods for `sfg`, `sfc` and `sf`;
their output is of the same class as their input:

* for `sfg` input, an `sfg` value is returned
* for `sfc` input, a new set of geometries is returned as `sfc`
* for `sf` objects, the same `sf` object is returned which has geometries replaced with the new ones.

|function|returns a geometry...                                                             |
|--------|----------------------------------------------------------------------------------|
|`st_centroid`|of type `POINT` with the geometry's centroid|
|`st_buffer`|that is this larger (or smaller) than the input geometry, depending on the buffer size|
|`st_jitter` |that was moved in space a certain amount, using a bivariate uniform distribution|
|`st_wrap_dateline`|cut into pieces that do no longer cover the dateline|
|`st_boundary`|with the boundary of the input geometry|
|`st_convex_hull`|that forms the convex hull of the input geometry (figure \@ref(fig:vor))|
|`st_line_merge`|after merging connecting `LINESTRING` elements of a `MULTILINESTRING` into longer `LINESTRING`s.|
|`st_make_valid`|that is valid |
|`st_node`|with added nodes to linear geometries at intersections without a node; only works on individual linear geometries|
|`st_point_on_surface`|with a (arbitrary) point on a surface|
|`st_polygonize`|of type polygon, created from lines that form a closed ring|
|`st_segmentize`|a (linear) geometry with nodes at a given density or minimal distance|
|`st_simplify`|simplified by removing vertices/nodes (lines or polygons)|
|`lwgeom::st_split`|that has been split with a splitting linestring|
|`st_transform`|transformed to a new coordinate reference system (chapter \@{rs})|
|`st_triangulate`|with triangulated polygon(s)|
|`st_voronoi`|with the voronoi tesselation of an input geometry (figure \@ref(fig:vor))|
|`st_zm`|with removed or added `Z` and/or `M` coordinates|
|`st_collection_extract`|with subgeometries from a `GEOMETRYCOLLECTION` of a particular type|
|`st_cast`|that is converted to another type|

```{r vor, fig.cap="left: convex hull (red) around a polygon (black); right: voronoi diagram (red) from a `MULTIPOINT` (black)", echo=TRUE, eval=TRUE, fig=TRUE}
par(mar = rep(0,4), mfrow = c(1, 2))
plot(st_geometry(nc)[1], col = NA, border = 'black')
plot(st_convex_hull(st_geometry(nc)[1]), add = TRUE, col = NA, border = 'red')
box()
set.seed(131)
mp = st_multipoint(matrix(runif(20), 10))
plot(mp)
plot(st_voronoi(mp), add = TRUE, col = NA, border = 'red')
box()
```

A number of operation can be applied directly to geometries
```{r}
(A = st_point(c(1,2)))
(B = st_linestring(rbind(c(2,2), c(3,4))))
-A
B + A
st_sfc(B + A) * matrix(c(1,0,0,2), 2, 2)
st_sfc(A, B) * c(3, 5) # scale first by 3, second by 5:
```

### Binary operations returning geometries

Binary functions that return a geometry include

|function|returns                                                    |infix operator|
|--------|-----------------------------------------------------------|:------------:|
|`st_intersection`|the overlapping geometries for pair of geometries|`&`|
|`st_union`|the combination of the geometries; also removes duplicate points, nodes or line pieces|`|`|
|`st_difference`|the geometries of the first after removing the overlap with the second geometry|`/`|
|`st_sym_differenc`|the combinations of the geometries after removing where they overlap|`%/%`|

When operating on two `sfg`, single geometries it is clear what
all these functions do: return a single geometry for this pair.
When given two {\em sets} of geometries (`sfc` or `sf` objects),
a new set of geometries is returned; for `st_intersection`
containing only the non-empty geometries, for all other operations
the geometries from all pairwise evaluation. In case the 
arguments are of class `sf`, the attributes of the objects are copied
over to all intersections to which each of the features contributed:

```{r}
a = st_sf(a = 1, geom = st_sfc(st_linestring(rbind(c(0,0), c(1,0)))))
b = st_sf(b = 1:3, geom = st_sfc(st_point(c(0,0)), st_point(c(1,0)), st_point(c(2,0))))
st_intersection(a, b)
```

When `st_intersection` or `st_difference` are called with a single
set of geometries (an `sfc` object), they perform an n-ary operation,
explained in the next section.

### N-ary operations returning a geometry

#### Union, c, and combine

Calling `st_union` with only a single argument leads either to computing
the union of all geometries, or applying union to each of the individual
geometries, depending on the setting of `by_feature`:
```{r}
st_union(b, by_feature = FALSE) # default
st_union(b, by_feature = TRUE) # default
```

The `c` method combines sets of geometries
```{r}
bb = st_geometry(b)
c(bb, bb)
```
or single geometries into single a new single geometry
```{r}
c(st_point(0:1), st_point(1:2))
```
and `st_combine` uses this to collapse features for different geometries into one:
```{r}
st_combine(c(bb, bb))
```
When using this on lines or polygons, it is easy to obtain invalid
geometries, and one needs to use `st_union` on the result.
```{r}
(x = st_combine(st_sfc(st_linestring(rbind(c(0,0), c(1,1))), st_linestring(rbind(c(1,0),c(0,1))))))
st_is_valid(x)
st_union(x) %>% st_is_valid()
```

#### N-ary intersection and difference

N-ary `st_intersection` and `st_difference` take a single argument,
but operate (sequentially) on all pairs, triples, quadruples etc.
Consider the plot in figure \@ref(fig:boxes): how do we identify
the box where all three overlap?  Using binary intersections, as of
gives us intersections of all pairs, double since `x` is passed
twice: 1-1, 1-1, 1-3, 2-1, 2-2, 2-3, 3-1, 3-2, 3-3:
```{r}
sq = function(pt, sz = 1) st_polygon(list(rbind(c(pt - sz), 
  c(pt[1] + sz, pt[2] - sz), c(pt + sz), c(pt[1] - sz, pt[2] + sz), c(pt - sz))))
x = st_sf(box = 1:3, st_sfc(sq(c(0,0)), sq(c(1.7, -0.5)), sq(c(0.5, 1))))
(ixx = st_intersection(x, x)) %>% nrow
lengths(st_overlaps(ixx, ixx))
```
When we use however
```{r}
(i = st_intersection(x))
```
we end up with a set of all seven distinct intersections, without overlaps.
```{r}
lengths(st_overlaps(i, i))
```
When given an `sf` object an `sf` is returned, with two additional
fields, one with the number of overlapping features, and a
list-column with the indexes of contributing feature geometries.

```{r boxes, fig.cap="left: three overlapping boxes -- how do we identify the small box where all three overlap? right: unique, non-overlapping n-ary intersections", echo=FALSE, eval=TRUE, fig=TRUE}
par(mar = rep(.1, 4), mfrow = c(1, 2))
plot(st_geometry(x), col = NA, border = sf.colors(3, categorical=TRUE), lwd = 3)
plot(st_intersection(st_geometry(x)), col = sf.colors(7, categorical=TRUE, alpha = .5))
```

Similarly, one can compute n-ary differences from a set $\{s_1, s_2,
s_3, ...\}$ by creating differences $\{s_1, s_2-s_1, s_3-s_2-s_1,
...\}$. This is done by
```{r}
(xd = st_difference(x))
```
The result is shown in figure \@ref(fig:diff), for `x` and for `x[3:1]`, to
make clear that the result here depends on order of the geometries.

```{r diff, fig.cap="difference between subsequent boxes, left: in original order; right: in reverse order", echo=FALSE, eval=TRUE, fig=TRUE}
par(mar = rep(.1, 4), mfrow = c(1, 2)) 
xg = st_geometry(x)
plot(st_difference(xg), col = sf.colors(3, alpha = .5, categorical=TRUE))
plot(st_difference(xg[3:1]), col = sf.colors(3, alpha = .5, categorical=TRUE))
```
Resulting geometries do not overlap:
```{r}
lengths(st_overlaps(xd, xd))
```

### Other geometry manipulators

`st_make_grid` creates a grid of square or hexagonal polygons, based on an input
bounding box and a grid cell size.

`st_graticule` creates a set of graticules, lines of constant latitude or longitude,
which can serve as a reference on small-scale (large area) maps.

## Precision {#precision}

Geometrical operations, such as finding out whether a certain
point is on a line, may fail when coordinates are represented by
highly precise floating point numbers, such as 8-byte doubles in
R. A remedy might be to limit the precision of the coordinates
before the operation.  For this, a _precision model_ is adopted by
`sf`. It uses a precision value to round coordinates (X, Y, Z and M)
right before they are encoded as well-known binary, and passed on to
the libraries where this may have an effect (GEOS, GDAL, liblwgeom).
We demonstrate this by an R - WKB - R roundtrip.

Rounding can be done in two different ways.  First, With a negative
precision value, 8-byte doubles get converted to 4-byte floats and
back again:
```{r}
(p = st_sfc(st_point(c(1e6/3, 1e4/3)), crs = 3857))
p %>% st_set_precision(-1) %>% st_as_binary() %>% st_as_sfc() %>% `[[`(1) %>% print(digits = 16)
```
Second, with a positive precision $p$, each coordinate value $c$
is replaced by 
$$c' = \mbox{round}(p \cdot c) / p$$
This implies that for instance with a precision of 1000, the number
of decimal places to round to is 1/1000, or to mm if the unit of
coordinates is metre:
```{r}
p %>% st_set_precision(1000) %>% st_as_binary() %>% st_as_sfc() %>% `[[`(1)
```
With a precision of e.g. 0.001 or 0.05, rounding to the nearest `1/precision`,
i.e. if the unit is m to the nearest 1000 m or 20 m, is obtained:
```{r}
p %>% st_set_precision(0.001) %>% st_as_binary() %>% st_as_sfc() %>% `[[`(1) # to nearest 1000
p %>% st_set_precision(0.05) %>% st_as_binary() %>% st_as_sfc()  %>% `[[`(1) # to nearest 20
```
As a convenience, precisions can also be specified as a `units` object, with
the unit to round to, e.g. to the nearest 5 cm:
```{r}
p %>% st_set_precision(units::set_units(5, cm)) %>% 
	st_as_binary() %>% 
	st_as_sfc() %>% `[[`(1) %>% print(digits = 10)
```
but this requires that the object, `p`, has a coordinate reference
system with known units.

In essence, these rounding methods bring the coordinates to
points on a regular grid, which is beneficial for geometric
computations. Of course, it also affects all computations like
areas and distances. Which precision values are best for which
application is often a matter of common sense
combined with trial and error. A reproducible example
illustrating the need for setting precision is found
[here](https://github.com/r-spatial/sf/issues/275).


## Warnings for longitude/latitude geometries {#longlat}

When working on geodetic coordinates (degrees longitude/latitude), package `sf`
gives warnings when it makes the assumption that coordinates are Cartesian, e.g. in
```{r}
i = st_intersects(nc[1,], nc[2,])
```
In many cases, making this assumption is not a problem. It might
be a problem when we have polygons that cover very large areas,
cover North or South pole, or have lines crossing or polygons
covering the dateline.
