# Feature attributes {#featureattributes}

Feature _attributes_ refer to the properties of features ("things")
that do not describe the feature's geometry. Feature attributes can
be _derived_ from geometry (e.g. lenght of a `LINESTRING`, area
of a `POLYGON`) but they can also refer to completely different
properties, such as 

* the name of a street or a county, 
* the number of people living in a country, 
* the type of a road
* the soil type in a polygon from a soil map.
* the opening hours of a shop

Although we believe that temporal properties of features are at no
less fundamental than their spatial properties, the simple feature
access standard and consequently the `sf` package does not give
time a similar role as space; more on that in chapter \@ref(array).

Most `sf` objects will contain both geometries and attributes for
features. The geometric operations described in the previous chapter
\@ref(geommanip) operate on geometries _only_. When applied to an
object of class `sf`, all _unary operations_ will, for functions
returning

* a predicate or a measure, ignore all attributes
* a geometr y, replace the object's geometry.

In all these cases, attribute _values_ remain unmodified. At first
sight, that looks rather harmless. But if we look into a simple
case of replacing a county boundary with a county centroid, as in
```{r}
library(sf)
library(dplyr)
system.file("gpkg/nc.gpkg", package="sf") %>%
	read_sf() %>%
	st_transform(32119) %>%
	select(BIR74, SID74, NAME) %>%
	st_centroid() %>%
	head(n = 1)
```
we receive a warning. This warning is justified for the first two
variables shown (total births and number of SID disease cases,
1974) which, as such, are _not associated with_ a feature whose
geometry is `POINT (385605.4 300303.5)`. The third variable,
`NAME` is however still the county name for the point indicated,
but the point geometry no longer _is_ the county geometry.

## Attribute-geometry relationship {#agr}

Changing the feature geometry without changing the feature attributes
does change the _feature_, since the feature is characterised by
the combination of geometry and attributes. Can we, ahead of time,
predict whether the resulting feature will still meaningfully relate
to the attribute data when we for instance replaced all geometries
with their convex hull or centroid? It depends.

Take the example of a road, represented by a `LINESTRING`, which has
an attribute property road width equal to 10 m. What can we say about
the road width of an arbitray subsectin of this road? That depends
on whether the attribute road length describes, for instance the
road width everywhere, meaning that road width is constant along the
road, or whether it describes an aggregate property, such as minimum
or average road width.  In case of the minimum, for an arbitrary
subsection of the road one could still argue that the minimum
road with must be at least as large as the minimum road width
for the whole segment, but it is no longer _the minimum_ for that
subsection. This gives us two "types" for the attribute-geometry
relationship (AGR):

* **constant** the attribute value is valid everywhere in or over the geometry
* **aggregate** the attribute is an aggregate, a summary value over the geometry

For polygon data, typical examples of **constant** AGR are

* land use for a land use polygon
* rock units or geologic strata in a geological map
* soil type in a soil map
* elevation class in a elevation map that shows elevation as classes
* klimate zone in a climate zone map

Typical examples for the **aggregate** AGR are

* population, either as number of persons or as population density
* other socio-economic data, summarised by area
* total emission of pollutants by region

A third type of AGR is that where an attribute identifies a feature
geometry. The example above is county `NAME`: the name identifies
the county, and is still the county `NAME` for any sub-area. 

* **identity** the attribute value uniquely identifies the geometry, there are no other geometries with the same value

Arbitrary sub-areas will lose the **identity** property but becomes
a **constant** attribute. An example is:

* any point inside a county is still part of the county and must have the same value for county name, but it does not longer represent the (total) geometry of that county.

We can specify the AGR of an attribute in an `sf` object by `st_set_agr`:
```{r}
nc <- system.file("gpkg/nc.gpkg", package="sf") %>%
	read_sf() %>%
	st_transform(32119)
nc1 <- nc %>% select(BIR74, SID74, NAME) %>%
	st_set_agr(c(BIR74 = "aggregate", SID74 = "aggregate", NAME = "identity"))
```
This helps to get rid of warnings that a particular attribute is assumed to be constant over a geometry, if it already is. The following no longer generates a warning
```{r}
nc1 %>% select(NAME) %>%
	st_centroid() %>%
	head(1)
```
and also changes AGR for `NAME` from `identity` to `constant` when replacing the geometry with the geometry's centroid:
```{r}
nc1 %>% select(NAME) %>%
	st_centroid() %>% 
	st_agr()
```

## Spatial join

## Aggregate and summarise

## Intersections

Suppose we have two datasets with different geometries and attributes, and we want to compute their
intersections:
```{r}
p1 = st_polygon(list(rbind(c(0,0), c(4,0), c(4,4), c(0,4), c(0,0))))
d1 = st_sf(a = c(3,1), geom = st_sfc(p1, p1 + c(4, 0)))
d2 = st_sf(b = c(4), geom = st_sfc(p1 * .75 + c(3, 2)))
plot(d1, xlim = c(0,8), ylim = c(0, 6), col = NA, border = 1, reset = FALSE)
plot(d2, col = NA, border = 'red', add = TRUE, lwd = 2)
```
What will the intersection of these two objects give?
```{r}
(i = st_intersection(d1, d2))
plot(d1, xlim = c(0,8), ylim = c(0, 6), col = NA, border = 1, lwd = 2, reset = FALSE)
plot(d2, col = NA, border = 'red', add = TRUE, lwd = 3)
plot(st_geometry(i), add = TRUE, col = grey(c(.7,.9)), , border = 'green', lwd = 1)
```
It gives the areas of intersection, along with the corresponding attributes of both objects, and a warning that attributes were assumed to be spatially constant.

## Area-weighted interpolation

Suppose we want to combine geometries and attributes of two datasets such, that we get attribute values of the first datasets summarised for the geometries of the second. There are various ways we can go for this. The simples one, building on the previous example, would be to obtain for the geometry of `d2` the attribute of `d1` that has the largest overlap with `d2`. This is obtained by
```{r}
st_join(d2, d1, largest = TRUE)
```

Another option would be to summarise the attribute, e.g. taking its mean, regardless the amount of overlap. This is obtained by
```{r}
aggregate(d1, d2, mean)
```

A third option is to apply area-weighted interpolation, meaning that we interpolate (average) the variable by taking into account the respective area contributions of overlap. This is done e.g. by
```{r}
d3 = st_sfc(p1 * .75 + c(3, 2), p1 * .75 + c(3,3))
st_interpolate_aw(d1, d3, extensive = FALSE)$a
st_interpolate_aw(d1, d3, extensive = TRUE)$a
```

### Spatially intensive and extensive variables

The difference between the two examples for area-weighted interpolation is how the final weighted sum (value times area of intersection) is normalised: by the target area (extensive), or by the sum of the area covered (intensive, `extensive = FALSE`). Spatially intensive variables are variables for which the value, when we split an area, does not _principally_ change. An example might be temperature, elevation, or population density. Spatially extensive variables are variables for which the value is also split, according to the area. Examples are population (amount), or area.

