# Reference Systems {#rs}


## Units of measurement {#units}

```{r echo=FALSE}
Sys.setenv(UDUNITS2_XML_PATH="")
```

### Quantities 

@cobbmoore wrote that "Data are not just numbers, they are numbers
with a context", and "in data analysis, context provides meaning".

The VIM (International vocabulary of metrology, @vim)
defines a _quantity_ as a "property of a phenomenon, body, or
substance, where the property has a magnitude that can be expressed
as a number and a reference", where "[a] reference can be a measurement
unit, a measurement procedure, a reference material, or a combination
of such."

One could argue whether all data is constitued of quantities, but
there is no need to argue that proper data handling requires that
numbers are accompanied by information on what the numbers mean,
and what they are about. 

A measurement system consist of _base units_ for base quantities, and
_derived units_ for derived quantities.  The SI measurement system
consist of the seven base units length (metre, m), mass (kilogram,
kg), time (second, s), electric current (ampere, A), thermodynamic
temperature (kelvin, K), amount of substance (mole, mol), and
luminous intensity (candela, cd).  Derived units are composed
of products of integer powers of base units; exampes are speed
($\mbox{m}~\mbox{s}^{-1}$) and density ($\mbox{kg}~\mbox{m}^{-3}$).

Many data variables have units that are not expressed as SI base
units or derived units. @hand discusses many such measurement scales,
e.g. those used to measure intelligence in social sciences, in the
context of measurement units.

### Unitless measures

The special case of unitless units can refer to either cases where
units cancelled out (e.g. mass fraction: kg/kg, or angle measured
in rad: m/m) or to cases where objects or events were counted
(e.g. 5 apples).  Adding an angle to a count of apples would not make
sense; adding 5 apples to 3 oranges may make sense if the result is
reinterpreted, e.g. as _pieces of fruit_. @flater discusses systems
for proper handling of unitless quantities; handling counts could for
instance link to domain-specific ontologies pointing out which things
were counted, and perhaps identifying super-classes, like _fruit_.

### Units in R

The `units` R package [@unitsrj] provides units
of measurement support for R, by interfacing the
[udunits2](https://www.unidata.ucar.edu/software/udunits/) units
database and C library. It allows for setting, converting and deriving 
units:
```{r}
library(units)
(a = set_units(1:3, m))
a_km = set_units(1:3, km)
a + a_km
b = set_units(4:6, s)
a / b
```
and raises errors in case of meaningless operations
```{r error=TRUE}
a + b
```

### Datum

For many quantities, the natural origin of values is zero. This
works for amounts, and differences between amounts results in
meaningful negative values. For locations and times, differences
have a natural zero interpretation: length and duration.  Absolute
location (position) and time need a fixed origin, from which we can
meaningfully measure other absolute space-time points: a datum.  For space,
a datum involves more than one dimension. The combination of a datum
and a measurement unit (scale) is a a _reference system_.  The next
two sections will deal with temporal and spatial reference systems,
and how they are handled in R.

## Temporal Reference Systems

R has two native classes for time-related data: `POSIXt` and `Date`,
which are used for specifying dates, and times.

### `Date`

`Date` objects are numeric vectors of class `Date`, which contain
the number of days since (or in case negative: before) Jan 1, 1970:
```{r}
(d = as.Date("1970-02-01"))
as.numeric(d)
```
We see that the `print` method as well as the `as.Date` method use
ISO 8601 [@iso8601] character strings, which is standard used to
read and write dates. We can also modify this to local conventions
by specifying a `format`:
```{r}
(d = as.Date("01.02.1970", format = "%d.%m.%Y"))
format(d, format = "%d.%m.%Y")
```
The default for `format` may depend on the locale settings of
the computer used. The help page of `?as.Date` contains further
discussion of date systems, and calendars used.

### POSIXt

`POSIXt` is an R native class for specifying times. It has two
subclasses: `POSIXct` represents time as a numeric, representing decimal
seconds since 1970-01-01 00:00 UTC, and `POSIXlt` contains the same
information as a list with all time components (second, minute,
hour, day of month, month, year) in list components:
```{r}
t = as.POSIXct("1970-01-01 01:00:00", tz = "UTC")
as.numeric(t)
names(unclass(as.POSIXlt(t)))
unclass(as.POSIXlt(t))$hour
```

### Time zones

Time zones can be seen as local modifiers of time: where time as
numeric value is stored with respect to UTC (universal coordinated
time), a local time zone is used to format it, and a time zone can
be specified for creation, and formatting:
```{r}
(t = as.POSIXct("1970-01-01 00:00:00", tz = "PST"))
```
this adds a time zone modifier to `t` that redefines the time origin, as
```{r}
as.numeric(t)
```
To convert `POSIXt` to `Date` we can use `as.Date`; this converts
to the _local_ date. Converting back to `POSIXct` looses the time
of day and time zone information.
```{r}
(t = as.POSIXct("1970-01-01 23:00:00", tz = "PST"))
as.Date(t)
format(as.POSIXct(as.Date(t)), tz = "UTC")
```

Working with local time zones is sometimes confusing when the data
we work with were not referenced to this time zone. It may help
to set
```{r}
Sys.setenv(TZ="UTC")
```
at the start of an R script. The effect is, at lease on some
platforms, that R thinks it is working in a UTC time zone.  This way,
the scripts will produce identical outputs, no matter in which time
zone it is run.

## Coordinate Reference Systems {#crs}

We follow @lott2015geographic when defining the following concepts (italics indicate literal quoting):

* a **coordinate system** is a _set of mathematical rules for specifying how coordinates are to be assigned to points_
* a **datum** is a _parameter or set of parameters that define the position of the origin, the scale, and the orientation of a coordinate system_, 
* a **geodetic datum** is a _datum describing the relationship of a two- or three-dimensional coordinate system to the Earth_, and
* a **coordinate reference system** is a _coordinate system that is related to an object by a datum; for geodetic and vertical datums, the object will be the Earth._

A readable text that further explains these concepts is @iliffelott.

Essentially it boils down to the Earth not following a regular shape. The topography of the Earth is of course known to vary strongly, but also the surface formed by constant gravity at mean sea level, the geoid, is also irregular. A commonly used model that is fit to the geoid is an ellipsoid of revolution, which is an ellipse with two identical minor axes. This model can be fit locally to be highly precise, can be fixed for particular tectonic plates (like ETRS89), or can be globally fit (like WGS84). 

The definitions above state that coordinates in degrees longitude
and latitude can only have a meaning, i.e. can only be understood
as Earth coordinates when the datum they relate to is given.

Recomputing coordinates in a new datum is called _coordinate
transformation_.

### Projections

When we look at geographical data on a paper map or a screen, or
on any _flat_ device, we see the values _projected_ -- they are no
longer arranged on an a sphere or ellipsoid.  Even if we plot degrees
longitude/latitude data on a flat x/y plane with unit aspect ratio,
we use a projection, called _plate carrée_. 

Note that even for projected data, the data that _were_ projected
are associated with a reference ellipsoid (datum). Going from one
projection to the other without changing datum is called _coordinate
conversion_, and usually passes through the geodetic coordinates of
the datum involved; up to numerical errors this process is lossless
and invertible.

### Describing Coordinate Reference Systems

@lott2015geographic describes a standard for encoding coordinate
reference system using _well known text_; the standard is referred
to as WKT2. GDAL and PROJ support this encoding (FIXME: verify this
is true by the time this book goes into print).

Traditionally, PROJ used a string representation to encode coordinate
reference systems (datums) and projections, called the proj4string.
Some of these come from a catalogue (originally) compiled by the
European Petroleum Survey Group (now: International Association of
Oil & Gas Producers), and have a so-called epsg code.

Package `sf` provides a `crs` class which is initialised either by
an epsg code, like
```{r}
st_crs(4326)
```
or by a PROJ string, like
```{r}
st_crs("+proj=longlat")
```
and can be missing valued
```{r}
st_crs()
```
A number of methods are available for `crs` objects:
```{r}
methods(class = "crs")
```
The `Ops` methods are convenience functions, e.g.
```{r}
st_crs(4326) == st_crs("+proj=longlat")
```
but there not all cases semantically identical `crs` objects will
yield equality in this test.

`st_as_text` converts a `crs` object into WKT, we print it using `cat`:
```{r}
cat(st_as_text(st_crs(4326), pretty = TRUE))
```

`st_crs` is also a generic, with methods for
```{r}
methods(st_crs)
```
The method for `sfc` objects can drill further into the underlying
data, by adding an argument; a few of these are printed by
```{r}
st_crs(st_sfc(crs = 4326), parameters = TRUE)[c(1:4, 8)]
```
where we see that udunits and GDAL units are integrated. The
major axis lengths (`SemiMajor`) and inverse flattening are used
e.g. to compute great circle distances on the ellipsoid, using the
algorithm from @karney2013algorithms. This algorithm is implemented
in PROJ, and interfaced by `lwgeom::st_geod_distance`, which is
called from `sf::st_distance` when objects have geodetic coordinates.
