# Large data sets {#large}

This chapter describes how large spatial and spatiotemporal datasets
can be handled with R, with a focus on packages `sf` and `stars`.
For practical use, we classify large data sets as either

* too large to fit in working memory, or
* also too large to fit on the local hard drive, or
* also too large to download it to locally managed compute
infrastructure (such as network attached storage)

these three categories correspond very roughly to Gigabyte-,
Terabyte- and Petabyte-sized data sets.

## Vector data: `sf` {#largesf}

### Reading from disk

Function `st_read` reads vector data from disk, using GDAL, and
then keeps the data read in working memory. In case the file is
too large to be read in working memory, several options exist to
read parts of the file. The first is to set argument `wkt_filter`
with a WKT text string containing a geometry; only geometries from
the target file that intersect with this geometry will be returned.
An example is

```{r}
library(sf)
(file = system.file("gpkg/nc.gpkg", package="sf"))
bb = "POLYGON ((-81.7 36.2, -80.4 36.2, -80.4 36.5, -81.7 36.5, -81.7 36.2))"
nc.1 = st_read(file, wkt_filter = bb)
```

The second option is to use the `query` argument to `st_read`,
which can be any query in "OGR SQL" dialect, which can be used to
select features from a layer, and limit fields. An example is:

```{r}
q = paste("select BIR74,SID74,geom from 'nc.gpkg' where BIR74 > 1500")
nc.2 = st_read(file, query = q)
```
note that `nc.pgkg` is the _layer name_, which can be obtained from `file` using `st_layers`.
Sequences of records can be read using `LIMIT` and `OFFSET`, to read records 51-60 use
```{r}
q = paste("select BIR74,SID74,geom from 'nc.gpkg' LIMIT 10 OFFSET 50")
nc.2 = st_read(file, query = q)
```

Further query options include selection on geometry type, polygon
area. When the dataset queried is a spatial database, then the query
is passed on to the database and not interpreted by GDAL; this means
that more powerful features will be available.  Further information
is found in the GDAL documentation under "OGR SQL dialect".

Very large file files or directories that are zipped can be read
without the need to unzip them, using the `/vsizip` (for zip),
`/vsigzip` (for gzip) or `/vsitar` (for tar files) prefix to files;
this is followed by the path to the zip file, and then followed by
the file inside this zip file. Reading files this way may come at
some computatational cost.

### Reading from databases, dbplyr

```{r, echo=FALSE}
has_PG <- any("PostgreSQL" %in% st_drivers()$name)
```

```{r echo=FALSE, eval=has_PG}
nc = read_sf(file)
write_sf(nc, "PG:dbname=postgis", "nc")
```

Although GDAL has support for several spatial databases, and as
mentioned above it passes on SQL in the `query` argument to the
database, it is sometimes beneficial to directly read and 
to a spatial database using the R database drivers for this. An
example of this is:
```{r, eval=has_PG}
pg <- DBI::dbConnect(
    RPostgres::Postgres(),
    host = "localhost",
    dbname = "postgis")
st_read(pg, query = "select BIR74,wkb_geometry from nc limit 3")
```
A spatial query might look like
```{r, eval=has_PG}
q = "SELECT BIR74,wkb_geometry FROM nc WHERE \
  ST_Intersects(wkb_geometry, 'SRID=4267;POINT (-81.49826 36.4314)');"
st_read(pg, query = q)
```
Here, the intersection is done in the database, and uses the spatial
index typically present.

The same mechanism works when using `dplyr` with a database backend:
```{r, eval=has_PG}
library(dplyr, warn.conflicts = FALSE)
nc_db = tbl(pg, "nc")
```

Spatial queries can be formulated and are passed on to the database:
```{r, eval=has_PG}
nc_db %>% 
	 filter(ST_Intersects(wkb_geometry, 'SRID=4267;POINT (-81.49826 36.4314)')) %>%
	 collect()
nc_db %>% filter(ST_Area(wkb_geometry) > 0.1) %>% head(3)
```

It should be noted that PostGIS' `ST_Area` computes the same
area as the `AREA` field in `nc`, which is the meaningless value
obtained by assuming the coordinates are projected, although they
are ellipsoidal.

### Reading from online resources or web services

GDAL drivers support reading from online recourses, by prepending
`/vsicurl/` before the URL starting with e.g. `https://`. A number of
similar drivers specialized for particular clouds include `/vsis3`
for Amazon S3, `/vsigs`  for Google Cloud Storage, `/vsiaz` for
Azure, `/vsioss` for Alibaba Cloud, or `/vsiswift` for OpenStack
Swift Object Storage. These prepositions can be combined e.g. with
`/vsizip/` to read a zipped online resource. Depending on the
file format used, reading information this way may always involve
reading the entire file, or reading it multiple times, and may not
always be the most efficient way of handling resources. A format
like "cloud-optimized geotiff" (COG) has been specially designed
to be efficient and resource-friendly in many cases, e.g. for only
reading the metadata, or for only reading overviews (low-resolutions
versions of the full imagery) or spatial segments. COGs can also
be created using the GeoTIFF driver of GDAL, and setting the right
dataset creation options in a `write_stars` call.

### API's, OpenStreetMap

Although online resource do not have to be stored files but could be
created server-side on the fly, typical web services for geospatial
data create data on the fly, and give access to this through an API.
As an example, data from [OpenStreetMap](https://openstreetmap.org/)
can be bulk downloaded and read locally, e.g. using the GDAL vector
driver, but more typical a user wants to obtain a small subset of
the data or use the data for a small query. Several R pacages exist
that query openstreetmap data: 

* Package `OpenStreetMap` downloads data as raster tiles, typically
  used as backdrop or reference for plotting other features
* Package `osmdata` downloads vector data as points, lines or polygons
  in `sf` or `sp` format
* Package `osmar` returns vector data, but in addition the network
  topology (as an `igraph` object) that contains how road elements
  form a network, and has functions that compute the shortest route.

When provided with a correctly formulated API call in the URL the
highly configurable GDAL OSM driver (in `st_read`) can read an
".osm" file (xml) and returns a dataset with five layers: `points`
that have significant tags, `lines` with non-area "way" features,
`multilinestrings` with "relation" features, `multipolygons` with
"relation" features  and `other_relations`. A simple and very small
bounding box query to OpenStreetMap could look like

```{r,eval=FALSE}
download.file(
  "https://openstreetmap.org/api/0.6/map?bbox=7.595,51.969,7.598,51.970"
  "data/ms.osm", method = "auto")
```
and from this file we can read the layer `lines`, and plot its
first attribute by
```{r overpass, fig.cap = "OpenStreetMap vector data"}
o = read_sf("data/ms.osm", "lines")
p = read_sf("data/ms.osm", "multipolygons")
bb = st_bbox(c(xmin=7.595, ymin = 51.969, xmax = 7.598, ymax = 51.970),
	crs = 4326)
plot(st_as_sfc(bb), axes = TRUE, lwd = 2, lty = 2, cex.axis = .5)
plot(o[,1], lwd = 2, add = TRUE)
plot(st_geometry(p), border = NA, col = '#88888888', add = TRUE)
```
the result of which is shown in figure \@ref(fig:overpass).
The overpass API provides a more generic and powerful query
functionality to OpenStreetMap data.

## Raster data: `stars`

A common challenge with raster datasets is not only that they come
in large files (single Sentinel-2 tiles are around 1 Gb), but that
many of these files, potentially thousands, are needed to address
the area and time period of interest. At time of writing this, the
Copernicus program that runs all Sentinel satellites publishes 160
Tb of images per day.  This means that a classic pattern in using R,
consisting of

* downloading data to local disc, 
* loading the data in memory, 
* analysing it

is not going to work.

Cloud-based Earth Observation processing platforms like Google Earth
Engine [@gorelick] or [Sentinel Hub](https://www.sentinel-hub.com/)
recognize this and let users work with datasets up to the petabyte
range rather easily and with a great deal of interactivity. They
share the following properties:

* computations are posponed as long as possible (lazy evaluation),
* only the data you ask for are being computed and returned, and nothing more,
* storing intermediate results is avoided in favour of on-the-fly computations,
* maps with useful results are generated and shown quickly to allow for interactive model development.

This is similar to the `dbplyr` interface to databases and
cloud-based analytics environments, but differs in the aspect of
_what_ we want to see quickly: rather than the first $n$ records
of a `dbplyr` table, we want a quick _overview_ of the results,
in the form of a map covering the whole area, or part of it, but
at screen resolution rather than native (observation) resolution.

If for instance we want to "see" results for the United States on
screen with 1000 x 1000 pixels, we only need to compute results
for this many pixels, which corresponds roughly to data
on a grid with 3000 m x 3000 m grid cells.  For Sentinel-2
data with 10 m resolution, this means we can subsample with
a factor 300, giving 3 km x 3 km resolution.  Processing,
storage and network requirements then drop a factor $300^2 \approx 10^5$, compared
to working on the native 10 m x 10 m resolution. On the platforms
mentioned, zooming in the map triggers further computations on a
finer resolution and smaller extent.

A simple optimisation that follows these lines is how stars' plot
method works:  in case of plotting large rasters, it subsamples
the array before it plots, drastically saving time.  The degree
of subsampling is derived from the plotting region size and the
plotting resolution (pixel density). For vector devices, such as pdf,
R sets plot resolution to 75 dpi, corresponding to 0.3 mm per pixel.
Enlarging plots may reveal this, but replotting to an enlarged
devices will create a plot at target density.

### `stars` proxy objects

To handle datasets that are too large to fit in memory, `stars`
provides `stars_proxy` objects.  To demonstrate its use, we will
use the `starsdata` package, an R data package with larger datasets
(around 1 Gb total). It can be installed by
```{r eval=FALSE}
install.packages("starsdata", repos = "http://pebesma.staff.ifgi.de", 
	type = "source")
```
We can "load" a Sentinel-2 image from it by
```{r}
f = "sentinel/S2A_MSIL1C_20180220T105051_N0206_R051_T32ULE_20180221T134037.zip"
granule = system.file(file = f, package = "starsdata")
file.size(granule)
base_name = strsplit(basename(granule), ".zip")[[1]]
s2 = paste0("SENTINEL2_L1C:/vsizip/", granule, "/", base_name, 
	".SAFE/MTD_MSIL1C.xml:10m:EPSG_32632")
(p = read_stars(s2, proxy = TRUE))
object.size(p)
```
and we see that this does not actually load _any_ of the pixel
values, but keeps the reference to the dataset and fills the
dimensions table. (The convoluted `s2` name is needed to point
GDAL to the right file inside the `.zip` file containing 115 files
in total).

The idea of a proxy object is that we can build expressions like
```{r}
p2 = p * 2
```
but that the computations for this are postponed. Only when we
really need the data, e.g. because we want to plot it, is `p *
2` evaluated.  We need data when either

* we want to `plot` data, or
* we want to write an object to disk, with `write_stars`, or
* we want to explicitly load an object in memory, with `st_as_stars`

In case the entire object does not fit in memory, `plot` and
`write_stars` choose different strategies to deal with this:

* `plot` fetches only the pixels that can be seen, rather than all
pixels available, and
* `write_stars` reads, processes, and writes data chunk by chunk.

Downsampling and chunking is implemented for spatially dense images,
not e.g. for dense time series, or other dense dimensions.

As an example, the output of `plot(p)`, shown in figure \@ref(fig:plotp)
```{r plotp, fig.cap="Plot of downsampled 10 m bands of a Sentinel-2 scene", echo=!knitr::is_latex_output()}
plot(p)
```
only fetches the pixels that can be seen on the plot device, rather
than the 10980 x 10980 pixels available in each band. The downsampling
ratio taken is
```{r}
(ds = floor(sqrt(prod(dim(p)) / prod(dev.size("px")))))
```
meaning that for every `r ds` $\times$ `r ds` sub-image in the
original image, only one pixel is read, and plotted. This value is
still a bit too low as it ignores the white space and space for
the key on the plotting device.

### Operations on proxy objects

A few dedicated methods are available for `stars_proxy` objects:
```{r}
methods(class = "stars_proxy")
```
We have seen `plot` and `print` in action; `dim` reads out
the dimension from the dimensions metadata table. 

The three methods that actually fetch data are `st_as_stars`,
`plot` and `write_stars`.  `st_as_stars` reads the actual data into a
`stars` object, its argument `downsample` controls the downsampling
rate. `plot` does this too, choosing an appropriate `downsample`
value from the device resolution, and plots the object. `write_stars`
writes a `star_proxy` object to disc.

All other methods for `stars_proxy` objects do not actually operate
on the raster data but add the operations to a _to do_ list,
attached to the object. Only when actual raster data are fetched,
e.g. by calling `plot` or `st_as_stars`, the commands in this list
are executed.

`st_crop` limits the extent (area) of the raster that will be
read. `c` combines `stars_proxy` objects, but still doesn't read
any data. `adrop` drops empty dimensions, `aperm` changes dimension
order.

`write_stars` reads and processes its input chunk-wise; it has an
argument `chunk_size` that lets users control the size of spatial
chunks.

## Very large data cubes

At some stage, data sets need to be analysed that are so large that
downloading them is no longer feasible; even when local storage would
be sufficient, network bandwidth may become limiting. Examples are
satellite image archives such as those from Landsat and Copernicus
(Sentinel-x), or model computations such as the ERA5 [@era5], a
model reanalysis of the global atmosphere, land surface and ocean
waves from 1950 onwards. In such cases it may be most helpful to
either gain access to (typically: rent) virtual machines in a cloud
where these data are available and nearby (i.e., the data should be
stored in the same data center as where the virtual machine is),
or to use a system that lets the user carry out such computations
without having to worry about virtual machines. Both options will
be discussed.

### Finding and processing assets

When working on a virtual machine on a cloud, a first task is usally
to find the assets (files) to work on. It looks attractive to obtain
a file listing, and then parse file names such as
```
S2A_MSIL1C_20180220T105051_N0206_R051_T32ULE_20180221T134037.zip
```
for their metadata including the date of acquisition and the code
of the spatial tile covered. Obtaining such a file listing however
is usually computationally very demanding, as is the processing of
the result, when the number of tiles runs in the many millions.

A solution to this is to use a catalogue. The recently developed
and increasingly deployed STAC, short for _spatiotemporal asset
catalogue_, provides an API that can be used to query image
collections by properties like bounding box, date, band, and cloud
coverage. The R package `rstac` [@R-rstac] provides an R interface
to create queries, and manage the information returned.

Processing the resulting files may involve creating a data cube at
a lower spatial and/or temporal resolution, from images that may
span a range of coordinate reference systems (e.g., several UTM
zones). An R package that can do that is gdalcubes [@R-gdalcubes;
@appel2019demand], which can also directly use STAC output
[@appelblog].

### Processing data: GEE, openEO

Platforms that do not require the management and programming of
virtual machines _in_ the cloud but provide direct access to the
imagery managed include GEE, openEO, and the climate data store.

Google Earth Engine (GEE) is a cloud platform that allows users
to compute on large amounts of Earth Observation data as well
as modelling products [@gorelick]. It has powerful analysis
capabilities, including most of the data cube operations explained
in section \@ref(dcoperations). It has an IDE where scripts can
be written in JavaScript, and a Python interface to the same
functionality. The code of GEE is not open source, and cannot be
extended by arbitrary user-defined functions in languages like
Python or R. R package `rgee` [@R-rgee] provides an R client
interface to GEE.

Cloud-based data cube processing platforms built entirely around
open source software are arising, several of which using the openEO
API [@openEO]. This API has room for user-defined functions (UDFs)
written in Python or R that are being passed on through the API and
executed at the pixel level, e.g. to aggregate or reduce dimensions.
UDFs in R represent the data chunk to be processed as `stars`
object, in Python `xarray` objects are used.

Other platforms include the Copernicus climate data store [@cds]
or atmosphere data store, which allow processing of atmospheric
or climate data from ECMWF, including ERA5. An R package with an
interface to both data stores is `ecmwfr` [@R-ecmwfr].

## Exercises 

Use R to solve the following exercises.

1. For the S2 image (above), find out in which order the bands are
using `st_get_dimension_values()`, and try to find out (e.g. by internet
search) which spectral bands / colors they correspond to.
1. Compute NDVI for the S2 image, using `st_apply` and an an appropriate 
`ndvi` function.  Plot the result to screen, and then write the
result to a GeoTIFF. Explain the difference in runtime between
plotting and writing.
1. Plot an RGB composite of the S2 image, using the `rgb` argument
to `plot()`, and then by using `st_rgb()` first.
1. select five random points from the bounding box of `S2`, and extract
the band values at these points; convert the object returned to an `sf`
object.
1.  For the 10 km radius circle around `POINT(390000  5940000)`, use
`aggregate` to compute the mean pixel values of the S2 image
when downsampling the images with factor 30, and on the original
resolution. Compute the relative difference between the results.
1. Use `hist` to compute the histogram on the downsampled S2 image.
Also do this for each of the bands. Use `ggplot2` to compute a
single plot with all four histograms.
1.  Use `st_crop` to crop the S2 image to the area covered by the 10 km circle.
Plot the results. Explore the effect of setting argument `crop = FALSE`
1.  With the downsampled image, compute the logical layer where all four
bands have pixel values higher than 1000. Use a raster algebra expression
on the four bands (use `split` first), or use `st_apply` for this.
