# Base and grid plots {#plot}

With base plots we mean the `plot` methods as offered by base R.
Higher-level plots created with `ggplot2`, which are built on top
of the `grid` package, are discussed in chapter \@ref(ggplot2).

## Base plots

The nice thing about base plots is that they work relatively simple,
and can be built incrementally: after plotting a first element, it
is possible to add further elements, like titles, axes, legends,
annotations, and so on. Making more complex plot such as faceted
plots is more of a challenge with base plot. They may also be
faster than e.g. `ggplot2`. This chapter focuses on the base `plot`
methods for `sf` and `stars` objects, and in particular where they
deviate from what one expects from `base` R plots: default colors,
color key placement, and multiple maps.

### `plot.sf` defaults

When we plot an `sf` object with multiple attributes, we get to see
multiple attributes, as shown in figure \@ref(fig:plotsf); the warning indicates that
the number of attributes plotted has been limited (`max.plot`),
that by default no key is shown when multiple attributes are plotted,
and that `factor` or `character` columns are plotted with a categorical
scale, and `numeric` variables by a continuous color scale.
```{r plotsf,out.width='100%',fig.cap="default plot with multiple attributes", fig=TRUE}
system.file("gpkg/nc.gpkg", package="sf") %>% read_sf -> nc
plot(nc)
```
When the attributes values have a common reference system it makes sense
to plot a key, and this can be done by
```{r plotkey,out.width='60%',fig.cap="two attributes sharing a key", fig=TRUE}
suppressPackageStartupMessages(library(dplyr))
nc %>% select(SID74, SID79) %>% plot(key.pos = 4) # 1: below; 4: right
```

### Controlling color, color breaks, key

The color of each feature geometry, be it a point, line or polygon,
can be controlled by passing a single value to argument `col`, or to
pass a vector with colors of length equal to the number of features.
In that case, one does not get an automatic key, since the mapping
of values to the key is unclear.

One can get further control over the key by specifying either of

* `breaks`, to one of the `classInt::classIntervals` styles ("fixed",
"sd", "equal", "pretty", "quantile", "kmeans", "hclust", "bclust",
"fisher", or "jenks"), or to the numeric vector of (increasing)
class break values, 
* `nbreaks` to set the number of color breaks, and
* `pal`, to pass a _function_ that generates a palette, e.g. `rainbow`
or `viridis::viridis`.
* `logz` which causes legend values to be plotted as 10-powers
* `at` which controls the values plotted along the key
* `key.pos` plots the key beneath (1), left (2), over (3) or to the
right (4) of the map(s), and is omitted if `NULL`.

An example is shown in figure \@ref(fig:viridis).

```{r viridis,out.width='60%',fig.cap="controlling key color and breaks; log transform causes zero values to remain uncolored; custom key size parameters",fig=TRUE}
plot(nc["SID74"], logz = TRUE, pal = viridis::viridis,
    breaks = c(0,.5,1,1.5,2,2.5), at = c(0,.5,1,1.5,2,2.5),
	key.width = lcm(1.3), key.length = 1)
```

For factor variables, the key shows factor levels, and `key.length`
and/or `key.width` may need to be further controlled to get them
look good.

### Incrementally adding plot elements {#reset}

We can add elements to maps without keys by simply adding `add
= TRUE`.  When a map has a key, the initial `plot` command needs
to have `reset = FALSE`.

The reason for this is as follows.
When plotting a map with a key on the side, in base plot one needs
to cut up the plotting region in a region for the map, and a region
for the key, similar to using `par(mfrow = c(1,2))`.
In order to not keep the plot region splitting active for subsequent,
unrelated plots, it is removed before the plot function returns. This
means that _by default_ one cannot add elements to a plot.

In case we want to add elements to a plot, we need to instruct `plot`
not to reset, and then can add elements if we set `add = TRUE`, as shown
in figure \@ref(fig:figadd).
```{r figadd,out.width='60%',fig.cap="adding plot elements to a map with a key; graticule and axes",fig=TRUE}
plot(nc["SID74"], pal = viridis::viridis, reset = FALSE, graticule = TRUE, axes = TRUE)
plot(st_point(c(-81.498,36.434)), col = 'red', pch = 1, cex = 4, lwd = 3, add = TRUE)
layout(matrix(1)) # manually reset the plotting region split-up
```

### plotting graticules and axes

A graticule (section \@ref(graticule)) is added if `graticule =
TRUE` is set; figure @\ref(fig:figadd) gives an example.  Axes are by
default omitted: they take space, and often other map elements, such
as boundaries or coast lines, are sufficient for orientation. Axes
can be added with `axes = TRUE`. They are given in values of the
coordinate reference system of the data, unless a graticule is added,
in which case they correspond to the graticule values.

### `plot.stars` defaults

The base plot method for `stars` objects follows a number of
parameters that were also discussed above for `plot.sf`. These are:
`key.pos`, `logz`, `axes`, `reset`, `key.width`, `key.length`
`nbreaks`,  and `breaks`.

The default for `breaks` is `"quantile"`; along with the default
`join_zlim = TRUE` this results in quantile values taken from all
of the images shown in a composite plot to yeild color breaks. The
result of this is shown in figure \@ref(fig:firststars), the default
`stars` plot for a multi-band raster. If `join_zlim` is `FALSE`,
color breaks are computed separately for each subplot, and a joint
legend is not possible, hence, no legend is drawn. For large datasets
with more than 10000 pixels, a (regular) sample of size 10000 is
taken to determine quantiles, rather than using the entire dataset.
The `col` argument provides the pallette used, and not a color
vector for each pixel; by default a linear grey scale is used.
`box_col` is the color used for the boxes around subplots.

With the default of `downsample = TRUE`, `plot.stars` does not many
more pixels than can be seen on the plotting device.

As shown in the left subfigure of \@ref(fig:starsrgb), the `rgb`
parameter can be used to specify three bands for creating rgb
colors. `maxColorValue`, by default the maximum of all bands,
can be used to specify the maximum pixel values for scaling pixel
values into rgb components.

The effect of setting `text_values` and `interpolate` are shown in
figure \@ref(fig:starstext) showing a 10 x 10 cell raster. `text_values` prints cell values as text,
e.g. for numeric examples. `interpolate` interpolates `rgb` values, and
seems to limit its extent to the raster cell centres, rather than corners.
```{r starstext,out.width='60%',fig.cap="stars plots with text values (left) and interpolated rgb colors (right); both maps plot a 10 x 10 raster"}
tif = system.file("tif/L7_ETMs.tif", package = "stars")
library(stars)
x = read_stars(tif)
par(mfrow = c(1, 2))
plot(x[,1:10,1:10,1], text_values=TRUE, key.pos = NULL, col = viridis::viridis(10), 
	reset = FALSE)
plot(x[,1:10,1:10,1:3], rgb=1:3, interpolate = TRUE, reset = FALSE)
```

Some rasters cannot be plotted by R's `image`, but have to be plotted
as features (polygons or points); examples are rotated, sheared
and curvilinear grids. Figure \@ref(fig:rastersheared) shows a sheared
grid, and illustrates `st_as_sf` for conversion to simple features,
`merge` for whether to merge polygons with identical values, and
`as_points` to plot raster cells as points.

```{r rastersheared,fig.cap="sheared rasters, plotted as simple features"}
par(mfrow = c(1, 3))
xs = adrop(x[,1:10,1:10,1])
attr(attr(xs, "dimensions"), "raster")$affine = c(1, 3)
plot(xs, col = viridis::viridis(10), key.pos = NULL, reset = FALSE)
plot(st_as_sf(xs, as_points = FALSE, merge = FALSE), pal = viridis::viridis, 
    key.pos = NULL, reset = FALSE)
plot(st_as_sf(xs, as_points = TRUE, merge = FALSE), key.pos = NULL, 
    pal = viridis::viridis, pch = 16, reset = FALSE)
```

## Combining base plots

The explanation about the need for parameter `reset` and the examples
above with multiple subplots setting it to `FALSE` already show the
limitations of this approach: if we want a key next to each map,
we cannot easily combine maps in subplots, using `sf`'s or `stars`'
`plot` methods. If we can do without the automatic legend, all
is fine.

A better approach to combine maps into a composite is using facet
plots, as will be shown in chapter \@ref(ggplot2) about `ggplot2`.

## Grid plots and viewports

## Combining base and grid plots

