# Plotting spatial data {#plotting}

Together with timelines, maps belong to the most powerful graphs,
perhaps because we can immediately relate where we are, or have
been, on the space of the plot. Two recent books on visualisation
[@Healy; @Wilke] contain chapters on visualising geospatial data or
maps. Here, we will not try to preach the do's and don'ts of maps,
but rather point out a number of possibilities how to do things,
point out challenges along the way and ways to mitigate them.

## Every plot is a projection {#transform}

The world is round, but plotting devices are flat.  As mentioned
in section \@ref(projections), any time we visualise, in any
way, the world on a flat device, we project: we convert ellipsoidal
coordinates into Cartesian coordinates. This includes the cases
where we think we "do nothing" (figure \@ref(fig:world), left),
or where show the world "as it is", e.g. as seen from space (figure
\@ref(fig:world), right).

```{r world,fig.cap="Earth country boundaries; left: mapping long/lat linearly to $x$ and $y$ (plate carrée); right: ``as seen from space'' (orthographic)", fig=TRUE, echo=knitr::is_latex_output(), message=FALSE}
library(sf)
library(rnaturalearth)
#sf_use_s2(FALSE) # FIXME:
w <- ne_countries(scale = "medium", returnclass = "sf")
suppressWarnings(st_crs(w) <- st_crs(4326))
layout(matrix(1:2, 1, 2), c(2,1))
par(mar = rep(0, 4))
plot(st_geometry(w))

# sphere:
library(s2)
g = as_s2_geography(TRUE) # Earth
co = s2_data_countries()
oc = s2_difference(g, s2_union_agg(co)) # oceans
b = s2_buffer_cells(as_s2_geography("POINT(-30 -10)"), 9800000) # visible half
i = s2_intersection(b, oc) # visible ocean
co = s2_intersection(b, co)
plot(st_transform(st_as_sfc(i), "+proj=ortho +lat_0=-10 +lon_0=-30"), col = 'lightblue')
plot(st_transform(st_as_sfc(co), "+proj=ortho +lat_0=-10 +lon_0=-30"), col = NA, add = TRUE)
```

The left plot of figure \@ref(fig:world) was obtained by
```{r eval=FALSE}
library(sf)
library(rnaturalearth)
w <- ne_countries(scale = "medium", returnclass = "sf")
plot(st_geometry(w))
```
and we see that this is the default projection for data with ellipsoidal coordinates, as indicated by
```{r}
st_is_longlat(w)
```

The projection taken in figure \@ref(fig:world) (left) is the
equirectangular (or equidistant cylindrical) projection, which maps
longitude and latitude linearly to the $x$ and $y$ axis, keeping
an aspect ratio of 1. Were we to do this for smaller areas not on
the equator, it makes sense to choose a plot ratio such that one
distance unit E-W equals one distance unit N-S on the center of
the plotted area, and this is the default behaviour of the `plot()`
method for unprojected `sf` or `stars` datasets, as well as the
default for `ggplot2::geom_sf()` (section \@rer(geomsf)).

We can also carry out this projection before plotting. Say we want to
plot Germany, then after loading the (rough) country outline,
we use `st_transform` to project:
```{r}
DE = st_geometry(ne_countries(country = "germany", returnclass = "sf"))
DE.eqc = st_transform(DE, "+proj=eqc +lat_ts=51.14 +lon_0=90w")
```
The `eqc` refers to the "equidistant cylindrical" projection of PROJ;
the projection parameter here is `lat_ts`, the latitude of true
scale (i.e., one length unit N-S equals one length unit E-W),
which was here chosen as the middle of the bounding box latitudes
```{r}
print(mean(st_bbox(DE)[c("ymin", "ymax")]), digits = 4)
```
When we now plot both maps (figure \@ref(fig:eqc)), they look
identical up to the values along the axes: degrees for ellipsoidal
(left), and metres for projected (Cartesian) coordinates.

```{r eqc,out.width='60%', fig.cap="Germany in equirectangular projection: with axis units degrees (left) and metres in the equidistant cylindrical projection (right)"}
par(mfrow = c(1, 2))
plot(DE, axes = TRUE)
plot(DE.eqc, axes = TRUE)
```

### What is a good projection for my data?

There is unfortunately no silver bullet here. Projections that
maintain all distances do not exist; only globes do. The most
used projections try to preserve

* areas (equal area),
* directions (conformal, e.g. Mercator),
* some properties of distances (e.g. equirectangular preserves distances along meridians, azimuthal equidistant preserves distances to a central point)

or some compromise of these. Parameters of projections decide what
is shown in the center of a map and what on the fringes, which
areas are up and which are down, and which areas are most enlarged.
All these choices are in the end political decisions.

It is often entertaining and at times educational to play around with
the different projections and understand their consequences. When
the primary purpose of the map however is not to entertain or educate
projection varieties, it may be preferrable to choose a well-known or
less surprising projection, and move the discussion which projection
should be preferred to a decision process on its own.

## Plotting points, lines, polygons, grid cells

Since maps are just a special form of plots of statistical data,
the usual rules hold. Frequently occuring challenges include:

* polygons may be very small, and vanish when plotted, 
* depending on the data, polygons for different features may well
overlap, and be visible only partially; using transparent fill
colors may help indentify them
* when points are plotted with symbols, they may easily overlap and be hidden; density maps (chapter \@ref(pointpatterns)) may be more helpful
* lines may be hard to read when coloured and may overlap regardless line width

### Colors

When plotting polygons filled with colors, one has the choice to plot
polygon boundaries, or to suppress these. If polygon boundaries draw
too much attention, an alternative is to colour them in a grey tone,
or another color that doesn't interfere with the fill colors. When
suppressing boundaries entirely, polygons with (nearly) identical
colors will no longer be visually distinguishable. If the property
indicating the fill color is constant over the region, such as land
cover type, then this is not a problem but if the property is an
aggregation, the region over which it was aggregated gets lost,
and by that the proper interpretation: especially for extensive
variables, e.g. the amount of people living in a polygon, this
strongly misleads. But even with polygon boundaries, using filled
polygons for such variables may not be a good idea.

The use of continuous color scales that have no noticable color
breaks for continuously varying variables may look attractive,
but is often more fancy than useful:

* it impracticle to match a color on the map with a legend value
* colors ramps often stretch non-linearly over the value range,
making it hard to convey magnitude

Only for cases where the identification of values is less
important than the continuity of the map, such as the coloring of
a high resolution digital terrain model, it does serve its goal.
Good colors scales are e.g. found in packages `RColorBrewer`
[@R-RColorBrewer], viridis [@R-viridis] or colorspace
[@R-colorspace; @colorspace].

### Color breaks: `classInt` {#classintervals}

When plotting continuous geometry attributes using a limited set
of colors (or symbols), classes need to be made from the data.
R package `classInt` [@R-classInt] provides a number of methods to
do so. The default method is "quantile":
```{r}
library(classInt)
# set.seed(1) needed ?
r = rnorm(100)
(cI <- classIntervals(r))
cI$brks
```
it takes argument `n` for the number of intervals, and a `style`
that can be one of "fixed", "sd", "equal", "pretty", "quantile",
"kmeans", "hclust", "bclust", "fisher" or "jenks".  Style "pretty"
may not obey `n`; if `n` is missing, ‘nclass.Sturges’ is used;
two other methods are available for choosing `n` automatically. If 
the number of observations is greater than 3000, a 10\% sample is used
to create the breaks for "fisher" and "jenks".

### Graticules and other navigation aids {#graticule}

Graticules are lines on a map that follow constant latitude or
longitude values. On figure \@ref(fig:world) left they are drawn
in grey. Graticules are often drawn in maps to give reference where
something is. In our first map in figure \@ref(fig:first-map) we can
read that the area plotted is near 35$^o$ North and 80$^o$ West.
Had we plotted the lines in the projected coordinate system, they
would have been straight and their actual numbers would not have
been very informative, apart from giving an interpretation of size
or distances when the unit is known, and familiar to the map reader.
Graticules, by that, also shed light on which projection
was used: equirectangular or Mercator projections will have straight
vertical and horizontal lines, conic projections have straight but
diverging meridians, equal area may have curved meridians

The real navigation aid on figure \@ref(fig:world) and most other
maps are geographical features like the state outline, country
outlines, coast lines, rivers, roads, railways and so on. If these
are added sparsely and sufficiently, a graticule can as well be
omitted. In such cases, maps look good without axes, tics, and labels,
leaving up a lot of plotting space to be filled with actual map data.

## Base `plot`

The `plot` method for `sf` and `stars` objects try to make quick,
useful, exploratory plots; for production quality, alternatives
with more control and/or better defaults are offered by packages
`ggplot2` [@R-ggplot2] or `tmap` [@R-tmap; @tmap].

By default, the plot method tries to plot "all" it is given.
This means that

* given a geometry only (`sfc`), the geometry is plotted, without colors,
* given a geometry and an attribute, the geometry is colored according to
  the values of the attribute, using a qualitative color scale for `factor`
  or `logical` attributes and a continuous scale otherwise,
* given multiple attributes, multiple maps are plotted, each with a color
  scale but a legend is omitted by default, as color assignment is
  done on a per sub-map basis,
* for `stars` objects with multiple attributes, only the first
  attribute is plotted; for three-dimensional raster cubes, all
  slices over the third dimension are plotted.

### Adding to plots with legends

The `plot` methods for `stars` and `sf` objects may show a color key
on one of the sides (e.g., figure \@ref(fig:first-map)). To do this
with `base::plot`, the plot region is split in two and two plots are
created: one with the map, and one with the legend.  By default, the
`plot` function resets the graphics device (using `layout(matrix(1))`
so that subsequent plots are not hindered by the device being split
in two. If one wants to _add_ to an existing plot having a color
legend, this is however what is needed, and resetting the plotting
device needs to be prevented by setting argument `reset = FALSE`,
and use `add = TRUE` in a subsequent call to `plot`, an example
is 

```{r figreset, fig.cap = "Annotating base plots that have a legend"}
sf_use_s2(TRUE)
library(sf)
nc = read_sf(system.file("gpkg/nc.gpkg", package="sf"))
plot(nc["BIR74"], reset = FALSE, key.pos = 4)
plot(st_buffer(nc[1,1], units::set_units(10, km)), col = 'NA', 
	 border = 'red', lwd = 2, add = TRUE)
```
which is shown in figure \@ref(fig:figreset).  Annotating `stars`
plots can be done in the same way when a _single_ stars layer is
shown. Annotating `stars` plots with multiple cube slices can be
done by adding a "hook" function that will be called on every slice
shown, as in
```{r starshook, fig.cap = "annotated multi-slice stars plot"}
library(stars)
r = read_stars(system.file("tif/L7_ETMs.tif", package = "stars"))
circ = st_bbox(r) %>% st_as_sfc() %>% st_sample(5) %>% st_buffer(300)
hook = function() plot(circ, col = NA, border = 'yellow', add = TRUE)
plot(r, hook = hook, key.pos = 4)
```
and as shown in figure \@ref(fig:starshook).

Base plot methods have access to the resolution of the screen device
and hence the base plot method for `stars` and `stars_proxy` object
will downsample dense rasters and only plot pixels at a density
that makes sense for the device available.

### Projections in base plots

The base `plot` method plots data with ellipsoidal coordinates
using the equirectangular projection, using a latitude parameter
equal to the middle latitude of the data bounding box (figure
\@ref(fig:eqc)). To control this parameter, either a projection
to another equirectangular can be applied before plotting, or the
parameter `asp` can be set to override, e.g. `asp=1` would lead to
plate carrée (figure \@ref(fig:world) left). Subsequent plots need
to be in the same coordinate reference system in order to make
sense with overplotting, this is not being checked.

### Colors and color breaks

In base plots, `nbreaks` can be used to set the number of color
breaks, and `breaks` either to the numeric vector with actual breaks,
or to a value for the `style` argument in `classInt::classIntervals`.

## Maps with `ggplot2`

Package `ggplot2` [@R-ggplot2; @ggplot2] can be used to create
more complex an nicer looking graphs; it has a geometry `geom_sf`
that was developed in conjunction with the development of `sf`, and
helps creating beautiful maps; an introduction to this is found in
[@moreno], a first example is shown in figure \@ref(fig:figgather).
The code used for this plot is:
```{r}
suppressPackageStartupMessages(library(tidyverse))
nc.32119 = st_transform(nc, 32119) 
year_labels = c("SID74" = "1974 - 1978", "SID79" = "1979 - 1984")
nc.32119 %>% select(SID74, SID79) %>% gather(VAR, SID, -geom) -> nc2
```
```{r eval=FALSE}
ggplot() + geom_sf(data = nc2, aes(fill = SID)) + 
  facet_wrap(~VAR, ncol = 1, labeller = labeller(VAR = year_labels)) +
  scale_y_continuous(breaks = 34:36) +
  scale_fill_gradientn(colors = sf.colors(20)) +
  theme(panel.grid.major = element_line(color = "white"))
```
where we see that two attributes had to be stacked (`gather`)
before plotting them as facets: this is the idea of "tidy" data,
and the `gather` method for `sf` objects automatically stacks the
geometry column too.

Because `ggplot2` creates graphics _objects_ before plotting them,
it can control the cooordinate reference system of all elements
involved, and will transform or convert all subsequent objects to
the coordinate reference system of the first. It will also draw a
graticule for the (default) thin white lines on a grey background,
and uses a datum (by default: WGS84) for this. `geom_sf()` can be
combined with

For package `stars`, a `geom_stars` has, at the moment of writing
this, very limited scope: it wraps `geom_sf` for vector data
cubes, and calls `geom_raster` for regular rasters and `geom_rect`
for rectilinear rasters. It downsamples if the user specifies a
downsampling rate, but has no access to the screen dimensions to
automatically do this. This may be just enough, for instance figure
\@ref(fig:ggplotstars) is created by the following commands:
```{r ggplotstars,fig.cap="Simple raster plot with ggplot2"}
library(ggplot2)
library(stars)
r = read_stars(system.file("tif/L7_ETMs.tif", package = "stars"))
ggplot() + geom_stars(data = r) +
		facet_wrap(~band) + coord_equal() +
		theme_void() +
        scale_x_discrete(expand=c(0,0)) + 
        scale_y_discrete(expand=c(0,0)) +
		scale_fill_viridis_c()
```
More elaborate `ggplot2`-based plots with `stars` objects may be
obtained using package `ggspatial` [@R-ggspatial]; non-compatible
but nevertheless `ggplot2`-style plots can be created with `tmap`.

## Maps with `tmap` {#tmap}

Package `tmap` [@R-tmap; @tmap] takes a fresh look on plotting
spatial data in R; it resembles `ggplot2` in the sense that it
composes graphics objects before printing, by building on the `grid`
package, and by concatenating map elements with a `+` between them,
but otherwise it is entirely independent from, and incompatible
with, `ggplot2`. It has a number of options that allow for highly
professional looking maps, and many defaults have been carefully
chosen. To recreate figure \@ref(fig:figgather), for instance,
we use
<!-- ```{r tmapnc, fig.cap = "tmap: using tm_polygons with two attributes"}
-->

```{r tmapnc, fig.cap = "tmap: using ... with two attributes"}
library(tmap)
tm_shape(nc.32119) + tm_polygons(c("SID74", "SID79"))
```
to create figure \@ref(fig:tmapnc) and
```{r tmapnc2, fig.cap = "tmap: using ..."}
tm_shape(nc2) + tm_polygons("SID") + tm_facets(by = "VAR")
```
to create figure \@ref(fig:tmapnc2).

Package `tmap` also has support for `stars` objects, an example created with
```{r tmapstars, fig.cap = "Simple raster plot with tmap"}
tm_shape(r) + tm_raster("L7_ETMs.tif")
```
is shown in figure \@ref(fig:tmapstars). More examples of the use of **tmap** are given in sectiion \@ref(area).


## Interactive maps: `leaflet`, `mapview`, `tmap`

Interactive maps as shown in figure \@ref(fig:mapviewfigure) can be
created with R packages `leaflet`, `mapview` or `tmap`. `mapview`
adds a number of capabilities to `leaflet` including a map legend,
configurable pop-up windows when clicking features, support for
raster data, and scalable maps with very large feature sets using
the filegeobuf file format, as well as facet maps that react
synchronously to zoom and pan actions.  Package `tmap` has the nice
option that after giving
```{r eval=FALSE}
tmap_mode("view")
```
all usual `tmap` commands are applied to an interactive html/leaflet widget, 
whereas after
```{r eval=FALSE}
tmap_mode("plot")
```
again all output is sent to R own graphics device.

