# ggplot2 {#ggplot2}

Function `ggplot` from package `ggplot2` [@wickham2016ggplot2]
provides a high-level interface to creating graphs, essentially
by composing all their ingredients and constraints in a
single expression. It implements the "grammar for graphics" by
@wilkinson2006grammar, and is the plotting package of choice in the
`tidyverse`.

Before `ggplot` 3.0.0 came out, the approach to plotting spatial
data was to _fortify_ it, meaning encode all the geometries as atomic
vectors in a `data.frame`, keeping an `ID` column to register which
coordinate belonged to which geometry, and repeating all non-geometry
attributes for each geometry _coordinate_.  This worked reasonably
well for points and lines, but not very well for polygons with holes.

Since `ggplot` 3.0.0 and package `sf`, this runs much smoother;
`ggplot` has received a `geom_sf` function that could take an `sf`
object and calls `st_as_grob` on each feature geometry to get an
object that can directly be added to the plot. In addition to that,
it takes care of automated datum transformations or projections
if different objects have differing coordinate reference systems,
and adds by default a graticule and degree axis tic labels.

@moreno1, @moreno2 and @moreno3 published three guest blogs on
r-spatial.org explaining the capabilities of ggplot for making
beautiful maps with `sf` and `ggplot2`.

## `geom_sf`

We will introduce the affordances of `geom_sf` here step by step.
We use a projected version of `nc` 
```{r}
library(tidyverse)
library(sf)
system.file("gpkg/nc.gpkg", package="sf") %>% read_sf() %>% 
    st_transform(32119) -> nc.32119
```
and create a first `ggplot` by
```{r}
ggplot() + geom_sf(data = nc.32119) 
```
It is attractive to think that
```{r eval=FALSE}
ggplot(nc.32119) + geom_sf()
```
would also work, but it doesn't -- it only works if the geometry
column is named `geometry`, which is not always the case (`sf`
objects may have more than one geometry column).

If we want to get rid of the axis tics and grid lines, we could use
```{r}
ggplot() + geom_sf(data = nc.32119) + theme_void() +
  theme(panel.grid.major = element_line(color = "white"))
```

A first ggplot2 plot with polygons colored by attributes 
(as in figure \@ref(fig:firstggplot)) is created by
```{r}
ggplot() + geom_sf(data = nc.32119) + aes(fill = BIR74) +
    scale_fill_gradientn(colors = viridis::viridis(20))
```

### facet plots

### multiple geometries in a single map

### graticule control

## using `stars` objects with `ggplot2`

```{r}
library(stars)
library(viridis)
system.file("tif/L7_ETMs.tif", package = "stars") %>% read_stars() -> x
ggplot() + geom_tile(data = as.data.frame(x), aes(x = x,y = y, fill = L7_ETMs.tif)) + 
	coord_equal() + 
	scale_fill_viridis() + 
	facet_wrap(~band) + 
	theme_void()
```



