# ggplot2 {#ggplot2}

Function `ggplot` from package `ggplot2` [@wickham2016ggplot2]
provides a high-level interface to creating graphs, essentially
by composing all their ingredients and constraints in a
single expression. It implements the "grammar for graphics" by
@wilkinson2006grammar, and is the plotting package of choice in the
`tidyverse`.

Before `ggplot` 3.0.0 came out, the approach to plotting spatial
data was to _fortify_ it, meaning encode all the geometries as atomic
vectors in a `data.frame`, keeping an `ID` column to register which
coordinate belonged to which geometry, and repeating all non-geometry
attributes for each geometry _coordinate_.  This worked reasonably
well for points and lines, but not very well for polygons with holes.

Since `ggplot` 3.0.0 and package `sf`, this runs much smoother;
`ggplot` has received a `geom_sf` function that could take an `sf`
object and calls `st_as_grob` on each feature geometry to get an
object that can directly be added to the plot. In addition to that,
it takes care of automated datum transformations or projections
if different objects have differing coordinate reference systems,
and adds by default a graticule and degree axis tic labels.

@moreno1, @moreno2 and @moreno3 published three guest blogs on
r-spatial.org explaining the capabilities of ggplot for making
beautiful maps with `sf` and `ggplot2`.

## `geom_sf`

We will introduce the properties of `geom_sf` here step by step.
We use a projected version of `nc` 
```{r}
library(tidyverse)
library(sf)
system.file("gpkg/nc.gpkg", package="sf") %>% read_sf() %>% 
    st_transform(32119) -> nc.32119
```
and create a first `ggplot` by
```{r}
ggplot() + geom_sf(data = nc.32119) 
```
It is attractive to think that
```{r eval=FALSE}
ggplot(nc.32119) + geom_sf()
```
would also work, but it doesn't -- it only works if the geometry
column is named `geometry`, which is not always the case (`sf`
objects may have more than one geometry column).

If we want to get rid of the axis tics and grid lines, we could use
```{r}
ggplot() + geom_sf(data = nc.32119) + theme_void() +
  theme(panel.grid.major = element_line(color = "white"))
```

A simple ggplot2 plot with polygons colored by attributes 
is created by
```{r}
ggplot() + geom_sf(data = nc.32119) + aes(fill = BIR74) +
    scale_fill_gradientn(colors = viridis::viridis(20))
```

### facet plots

Facet plots are a powerful means to compare maps, because they
keep all the plotting parameters constant (spatial extent, scale,
color breaks).  One would perhaps wish that it were possible to
directly plot multiple attribute columns over facet maps.  For this,
we first have to reorganise the data such that the target variable is
a single column, another column indicates the facet, and geometries
are repeated accordingly.  For this, `tidyr::gather` can be used; an
example is given in figure \@ref(fig:firstgather).

### multiple geometries in a single map

Multiple geometries with `geom_sf` can be created by adding consecutive
`geom_sf` geometries; we can add for instance the county centroids to
a plot by
```{r}
ggplot() + geom_sf(data = nc.32119) + geom_sf(data = st_centroid(nc.32119))
```

When subsequent `sf` objects have a different coordinate reference
system from the first object, `geom_sf` will transform them to the
reference system of the first object.

### Fine tuning

In case `sf` objects have multiple geometry list-columns, the
"active" list column is selected by default. This can be overriden
by supplying another column name in `aes(geometry = my_column)` `

To add labels to geometries, one can use `geom_sf_label`. From the examples of `geom_sf`:
```{r}
nc <- sf::st_read(system.file("gpkg/nc.gpkg", package = "sf"), quiet = TRUE)
nc_3857 <- sf::st_transform(nc, "+init=epsg:3857")
ggplot() + 
    geom_sf(data = nc_3857[1:3, ], aes(fill = AREA)) + 
	geom_sf_label(data = nc_3857[1:3, ], aes(label = NAME))
```
`geom_sf_text` can be used to add simple text annotations without decoration.

Graticules are drawn by default. Since they are different for
every projection, it is hard to anticipate how strongly they will
be curved. `geom_sf` takes a parameter `ndiscr`, by default set to
100, which can be increased when graticulas show up unexpectedly
as non-smooth lines.

`label_graticules` can be used to control which lines of the graticule are
labeled. In addition, `label_axes` controls on which axes particular
graticule lines will be labeled.

## `geom_stars`

Package `stars` comes with a `geom_stars` function that is much
more limited in scope than `geom_sf`. In essence, it creates a call
to 

* `geom_raster` in case of raster data with a regular grid, 
* to `geom_tile` for other raster data, or 
* to `geom_sf` if the `stars` object has a simple feature geometry dimension rather than raster dimensions, or has a curvilinear raster

`geom_raster` also creates the mapping of dimension names to x
and y-coordinates and set the first attribute name as the `fill`
variable. This means that the aspect ratio still needs to be
controlled (`coord_equal()`) and that a `facet_wrap` is needed
to display multiple rasters. An example is shown in figure
\@ref(fig:geomraster).

(ref:geomstars) example of `geom_stars`

```{r geomraster,fig.cap='(ref:geomstars)'}
library(stars)
library(ggplot2)
library(viridis)
system.file("tif/L7_ETMs.tif", package = "stars") %>% read_stars() -> x
g = ggplot() + 
	coord_equal() + 
	scale_fill_viridis() + 
	theme_void() +
    scale_x_discrete(expand=c(0,0)) +
    scale_y_discrete(expand=c(0,0))
g + geom_stars(data = x) + 
	facet_wrap(~band)
```

`geom_stars` has a parameter, `downsample`, which can be used to downsample
particular dimensions. Here we downsample a 90m x 90m raster to a 900m x 900m
raster:
```{r}
g + geom_stars(data = x, downsample = c(10,10,1)) + 
	facet_wrap(~band)
```

```{r}
library(spacetime)
data(air) # this loads several datasets in .GlobalEnv
d = st_dimensions(station = st_as_sfc(stations), time = dates)
aq = st_as_stars(list(PM10 = air), dimensions = d)
# ggplot() + geom_stars(data = aq[,,3000])
aq.sf = st_as_sf(aq[,,3000], long=TRUE)
ggplot() + 
   geom_sf(data = st_as_sf(DE_NUTS1)) + 
   geom_sf(data = aq.sf, mapping = aes(col = PM10)) + 
   ggtitle(aq.sf$time[1])
```

