# Spatial Interpolation {#sec-interpolation}

\index{spatial interpolation}
\index{interpolation}

Spatial interpolation is the activity of estimating values of spatially
continuous variables (fields) for spatial locations where they have not
been observed, based on observations.  The statistical methodology
for spatial interpolation, called geostatistics, is concerned with
the modelling, prediction, and simulation of spatially continuous
phenomena.  The typical problem is a missing value problem: we
observe a property of a phenomenon $Z(s)$ at a limited number
of sample locations $s_i, i = 1,...,n$, and are interested in
the property value at all locations $s_0$ covering an area of
interest, so we have to predict it for unobserved locations. This
is also called _kriging_, or Gaussian Process prediction.
In case $Z(s)$ contains a white noise component $\epsilon$, as in
$Z(s)=S(s)+\epsilon$ (possibly reflecting measurement error) an
alternative but similar goal is to predict or simulate $S(s)$ rather
than $Z(s)$, which may be called _spatial filtering_ or _smoothing_.

In this chapter we will show simple approaches for handling
geostatistical data, demonstrate simple interpolation methods,
and explore modelling spatial correlation, spatial prediction and
simulation. @sec-stgeostatistics focuses on more complex
multivariate and spatiotemporal geostatistical models.  We will
use package **gstat** [@R-gstat; @gstatcg], which offers a fairly
wide palette of models and options for non-Bayesian geostatistical
analysis.  Bayesian methods with R implementations are found in
@DiggleTawnMoyeed:98, @Diggle:2007, @blangiardo2015spatial, and
@wikle2019spatio. An overview and comparison of methods for large
datasets is given in @Heaton2018.

\index{spatial autocorrelation in fields}

```{r echo=FALSE, eval=FALSE}
# after running the "Spatiotemporal Geostatistics" chapter, write NO2 means by
write_csv(right_join(a2, tb), "no2.csv")
```

```{r echo=FALSE}
CI <- Sys.getenv("CI") == "true" # TRUE when inside GitHub Actions
```

## A first dataset

We can read station mean NO$_2$ values, a dataset that is prepared 
in @sec-stgeostatistics, by loading it from package **gstat** using
```{r, message=FALSE}
library(tidyverse) |> suppressPackageStartupMessages()
no2 <- read_csv(system.file("external/no2.csv", 
    package = "gstat"), show_col_types = FALSE)
```
and convert it into an `sf` object with an appropriate UTM projection using
```{r}
library(sf)
crs <- st_crs("EPSG:32632")
st_as_sf(no2, crs = "OGC:CRS84", coords = 
	c("station_longitude_deg", "station_latitude_deg")) |>
	st_transform(crs) -> no2.sf
```

Next, we can load country boundaries and plot these data using `ggplot`, shown
in @fig-plotDE.
```{r plot}
read_sf("data/de_nuts1.gpkg") |> st_transform(crs) -> de
```

```{r fig-plotDE, echo = !knitr::is_latex_output()}
#| fig.cap: "Mean NO$_2$ concentrations in air for rural background stations in Germany, in 2017"
#| code-fold: true
ggplot() + geom_sf(data = de) + 
	geom_sf(data = no2.sf, mapping = aes(col = NO2))
```

\index{interpolation!target grid}
\index{st\_as\_stars!for generating a grid}

If we want to interpolate, we first need to decide where. This
is typically done on a regular grid covering the area of
interest. Starting with the country outline in object `de` we can
create a regular 10 km $\times$ 10 km grid over Germany by

```{r buildgridgermany}
library(stars) |> suppressPackageStartupMessages()
st_bbox(de) |>
  st_as_stars(dx = 10000) |>
  st_crop(de) -> grd
grd
```

Here, we chose grid cells not too fine, so that we still see
them in plots.

Perhaps the simplest interpolation method is inverse distance weighted
interpolation, which is a weighted average, using weights inverse
proportional to distances from the interpolation location:

\index{interpolation!inverse distance weighted}
\index{inverse distance weighted interpolation}

$$
\hat{z}(s_0) = \frac{\sum_{i=1}^{n} w_i z(s_i)}{\sum_{i=1}^n w_i}
$$

with $w_i = |s_0-s_i|^{-p}$, and the inverse distance power $p$ typically
taken as 2, or optimised using cross-validation. We can compute
inverse distance interpolated values using `gstat::idw`, 
```{r, message=FALSE}
library(gstat)
i <- idw(NO2~1, no2.sf, grd)
```

and plot them in @fig-idw.

```{r fig-idw, echo=!knitr::is_latex_output()}
#| fig.cap: "Inverse distance weighted interpolated values for NO$_2$ over Germany"
#| code-fold: true
ggplot() + geom_stars(data = i, 
					  aes(fill = var1.pred, x = x, y = y)) + 
    xlab(NULL) + ylab(NULL) +
	geom_sf(data = st_cast(de, "MULTILINESTRING")) + 
	geom_sf(data = no2.sf)
```

## Sample variogram 

\index{variogram!sample}
\index{sample variogram}

In order to make spatial predictions using geostatistical methods, we first need to identify a model for the mean and for the spatial correlation. In the simplest model, $Z(s) = m + e(s)$, the mean is an unknown constant $m$, and in this case the spatial correlation can be modelled using the variogram, $\gamma(h) = 0.5 E (Z(s)-Z(s+h))^2$. For processes with a finite variance $C(0)$, the variogram is related to the covariogram or covariance function through $\gamma(h) = C(0)-C(h)$.

The sample variogram is obtained by computing estimates of $\gamma(h)$ for distance intervals, $h_i = [h_{i,0},h_{i,1}]$:
$$
\hat{\gamma}(h_i) = \frac{1}{2N(h_i)}\sum_{j=1}^{N(h_i)}(z(s_i)-z(s_i+h'))^2, \ \ h_{i,0} \le h' < h_{i,1}
$$ {#eq-samplevariogram}

with $N(h_i)$ the number of sample pairs available for distance interval $h_i$.
Function `gstat::variogram` computes sample variograms,
```{r computevariogram}
v <- variogram(NO2~1, no2.sf)
```
and the result of plotting this is shown in @fig-vgm.
```{r fig-vgm, echo=!knitr::is_latex_output()}
#| fig.cap: "Sample variogram plot"
#| code-fold: true
plot(v, plot.numbers = TRUE, xlab = "distance h [m]",
	 ylab = expression(gamma(h)),
	 xlim = c(0, 1.055 * max(v$dist)))
```
Function `variogram` chooses default for maximum distance (`cutoff`: one-third of the length of the bounding box diagonal) and (constant) interval widths (`width`: cutoff divided by 15). These defaults can be changed by
```{r controlcutoff}
v0 <- variogram(NO2~1, no2.sf, cutoff = 100000, width = 10000)
```
shown in @fig-vgm2.
```{r fig-vgm2, echo=!knitr::is_latex_output()}
#| fig.cap: "Sample variogram plot with adjusted cutoff and lag width"
#| code-fold: true
plot(v0, plot.numbers = TRUE, xlab = "distance h [m]",
	 ylab = expression(gamma(h)),
	 xlim = c(0, 1.055 * max(v0$dist)))
```
Note that the formula `NO2~1` is used to select the variable of interest from the data file (`NO2`), and to specify the mean model: `~1` specifies an intercept-only (unknown, constant mean) model.

## Fitting variogram models

\index{variogram!model}

In order to progress towards spatial predictions, we need a variogram _model_ $\gamma(h)$ for (potentially) all distances $h$, rather than the set of estimates derived above. If we would connect these estimates with straight lines, or assume they reflect constant values over their respective distance intervals, it would lead to statistical models with non-positive definite covariance matrices, which would block using them in prediction.

\index{variogram!model!WLS fitting}
\index[function]{fit.variogram}

To avoid this, we fit parametric models $\gamma(h)$ to the estimates $\hat{\gamma}(h_i)$, where we take $h_i$ as the mean value of all the $h'$ values involved in estimating $\hat{\gamma}(h_i)$. We can fit for instance a model with an exponential variogram by
```{r}
v.m <- fit.variogram(v, vgm(1, "Exp", 50000, 1))
```
shown by the solid line in @fig-fitvariogrammodel.
```{r fig-fitvariogrammodel, echo = !knitr::is_latex_output()}
#| fig.cap: "Sample variogram (circles) with models fitted using weighted least squares (solid line) and maximum likelihood estimation (dashed line)"
#| code-fold: true
fit.variogram_ml <- function(formula, data, init, ...) {
  stopifnot(nrow(init) <= 2, inherits(data, "sf"), inherits(formula, "formula"),
    inherits(init, "variogramModel"))
  if (nrow(init) == 2)
    stopifnot("Nug" %in% init$model)

  # convert from parameter vector to "variogramModel" class:
  # x is c(sill, range) or: c(sill, range, nugget)
  get_model <- function(x, model, min_range = 1e-10) {
    sill <- x[1]
    range <- max(x[2], min_range)
    nugget <- if (length(x) == 3)
		    x[3]
      else
		    0.
    m <- vgm(sill, model, range, nugget)
  }

  # with A <- chol(Q), solve Q x = b for x:
  ch_solve <- function(A, b) {
    backsolve(A, forwardsolve(A, b, upper.tri = TRUE, transpose = TRUE))
  }

  # negative log likelihood, excluding the constant:
  nll <- function(x, d, res, model, ...) {
    m <- get_model(x, model, ...)
    Q <- variogramLine(m, dist_vector = d, covariance = TRUE)
    Qc <- chol(Q)
    det <- 2 * sum(log(diag(Qc)))
    det + t(res) %*% ch_solve(Qc, res)
  }

  # distance matrix, for optim stability rescaled to [0,1] range
  d <- st_distance(data) |> units::drop_units()
  max_d <- max(d)
  d <- d / max_d
  # residuals y - X beta: scale to sd 1
  res <- residuals(lm(formula, data))
  v <- var(res)
  res <- res/sqrt(v)
  if (nrow(init) == 2) {
    o.init <- c(init$psill[2], init$range[2], init$psill[0])
    model <- as.character(init$model[2])
  } else {
    o.init <- c(init$psill[1], init$range[1])
    model <- as.character(init$model[1])
  }
  o.init[2] <- o.init[2] / max_d # scale to [0,1]
  o.init[-2] <- o.init[-2] / v   # scale to sd 1
  o <- optim(o.init, nll, d = d, res = res, model = model, 
			lower = rep(0, length(o.init)), method = "L-BFGS-B", ...)$par
  o[2] <- o[2] * max_d # scale back to distance units
  o[-2] <- o[-2] * v # scale back to variance v
  get_model(o, model)
}
# use WLS fit model v.m for initial parameters:
v.ml <- fit.variogram_ml(NO2~1, no2.sf, v.m)
# plot(v, v.ml, plot.numbers = TRUE)
# plot(v, v.m, plot.numbers = TRUE) ## draws a single model; draw 2 models in single plot:
par(xaxs = "i", yaxs = "i")
plot(gamma ~ dist, v, 
	 xlim = c(0, 1.075 * max(v$dist)), ylim = c(0, 1.05 * max(v$gamma)),
     xlab = "distance h [m]", ylab = expression(gamma(h)))
lines(variogramLine(v.m, 1.075 * max(v$dist)), lty = 1, col = 'blue')
lines(variogramLine(v.ml, 1.075 * max(v$dist)), lty = 2, col = 'blue')
text(v$dist, v$gamma, v$np, pos = 4)
```
The fitting for the drawn line was done by weighted least squares,
minimising 
$$
\sum_{i=1}^{n}w_i(\gamma(h_i)-\hat{\gamma}(h_i))^2,
$$ {#eq-wlsweights}
with weights $w_i$ by default equal to $N(h_i)/h^2$. Other weight
options are available through argument `fit.method`.

\index{variogram!model!maximum likelihood estimation}

As an alternative to weighted least squares fitting, one can
use maximum likelihood (ML) or restricted maximum likelihood
parameter estimation [@kitanidis1985maximum], which for
this case leads to a relatively similar fitted model, shown
as the dashed line in @fig-fitvariogrammodel. An advantage of
ML-type approaches is that they do not require choosing distance
intervals $h_i$ in @eq-samplevariogram or weights $w_i$ in
@eq-wlsweights. Disadvantages are that they lean on stronger
assumptions of multivariate normally distributed data, and for
larger datasets require iteratively solving linear systems of size
equal to the number of observations; @Heaton2018 compare approaches
dedicated to fitting models to large datasets.

## Kriging interpolation {#sec-kriging}

\index{interpolation!kriging}
\index{kriging!ordinary}
\index{ordinary kriging}

Typically, when we interpolate a variable, we do that on
points on a regular grid covering the target area. We first create
a `stars` object with a raster covering the target area, and `NA`s
outside it.

Kriging involves the prediction of $Z(s_0)$ at arbitrary locations
$s_0$.  We can krige NO$_2$ by using `gstat::krige`, with the model
for the trend, the data, the prediction grid, and the variogram
model as arguments (@fig-krigeovergermany) by:

\index{krige}

```{r}
k <- krige(NO2~1, no2.sf, grd, v.m)
```
```{r fig-krigeovergermany, echo = !knitr::is_latex_output()}
#| fig.cap: "Kriged NO$_2$ concentrations over Germany"
ggplot() + geom_stars(data = k, aes(fill = var1.pred, x = x, y = y)) + 
    xlab(NULL) + ylab(NULL) +
	geom_sf(data = st_cast(de, "MULTILINESTRING")) + 
	geom_sf(data = no2.sf) +
	coord_sf(lims_method = "geometry_bbox")
```

## Areal means: block kriging {#sec-blockkriging}

\index{kriging!block}
\index{kriging!areal means}

Computing areal means can be done in several ways. The simplest is to take
the average of point samples falling inside the target polygons:
```{r}
a <- aggregate(no2.sf["NO2"], by = de, FUN = mean)
```
A more complicated way is to use _block kriging_ [@jh78], which
uses _all_ the data to estimate the mean of the variable over the
target areas. With `krige`, this can be done by giving the target
areas (polygons) as the `newdata` argument:
```{r krigeNO2regionmeans}
b <- krige(NO2~1, no2.sf, de, v.m)
```
we can now merge the two maps into a single object to create a single plot (@fig-aggregations):
```{r}
b$sample <- a$NO2
b$kriging <- b$var1.pred
```

```{r fig-aggregations, echo = !knitr::is_latex_output()}
#| fig.cap: "Aggregated NO$_2$ values from simple averaging (left) and block kriging (right)"
b |> select(sample, kriging) |> 
		pivot_longer(1:2, names_to = "var", values_to = "NO2") -> b2
b2$var <- factor(b2$var, levels = c("sample", "kriging"))
ggplot() + geom_sf(data = b2, mapping = aes(fill = NO2)) + facet_wrap(~var) +
	 scale_fill_gradientn(colors = sf.colors(20))
```
We see that the signal is similar, but that the sample means from simple 
averaging are more variable than the block kriging values; this may be due to
the smoothing effect of kriging: data points outside the aggregation
area receive weight, too.

\index{kriging!standard errors}

To compare the standard errors of means, for the sample mean we
can get a rough guess of the standard error by $\sqrt{(\sigma^2/n)}$:
```{r}
SE <- function(x) sqrt(var(x)/length(x))
a <- aggregate(no2.sf["NO2"], de, SE)
```
which would have been the actual estimate in design-based inference (@sec-design)
if the sample were obtained by spatially random sampling.
The block kriging variance is the model-based estimate and is
a by-product of kriging. We can compare the two in @fig-aggrSE where
we see that the simple averaging approach gives more variability and mostly
larger values for prediction errors of areal means, compared to
block kriging.

```{r fig-aggrSE, echo = !knitr::is_latex_output()}
#| fig.cap: "Standard errors for mean NO$_2$ values obtained by simple averaging (left) and block kriging (right)"
#| code-fold: true
b$sample <- a$NO2
b$kriging <- sqrt(b$var1.var)
b |> select(sample, kriging) |> 
		pivot_longer(1:2, names_to = "var", 
					 values_to = "Standard_error") -> b2
b2$var <- factor(b2$var, levels = c("sample", "kriging"))
ggplot() +
    geom_sf(data = b2, mapping = aes(fill = Standard_error)) +
    facet_wrap(~var, as.table = FALSE) + 
    scale_fill_gradientn(colors = sf.colors(20))
```

## Conditional simulation

\index{kriging!conditional simulation}
\index{conditional simulation}

In case one or more conditional realisation of the field $Z(s)$
are needed rather than their conditional mean, we can obtain this
by _conditional simulation_. A reason for wanting this may be the
need to estimate areal mean values of $g(Z(s))$ with $g(\cdot)$
a non-linear function; a simple example is the areal fraction where
$Z(s)$ exceeds a threshold.

The default approach used by `gstat` is to use the sequential
simulation algorithm for this. This is a simple algorithm that
randomly steps through the prediction locations and at each location:

* carries out a kriging prediction
* draws a random variable from the normal distribution with mean and variance equal to the kriging variance
* adds this value to the conditioning dataset
* finds a new random simulation location

until all locations have been visited. 

This is carried out by `gstat::krige` when `nsim` is set to a
positive value:
```{r condsim}
set.seed(13341)
(s <- krige(NO2~1, no2.sf, grd, v.m, nmax = 30, nsim = 6))
```
where `set.seed()` was called here to allow reproducibility.

It is usually needed to constrain the (maximum) number of nearest
neighbours to include in kriging estimation by setting `nmax`
because the dataset grows each step, leading otherwise quickly
to very long computing times and large memory requirements. Resulting
conditional simulations are shown in (@fig-plotkrigingvalues).

```{r fig-plotkrigingvalues, echo=!knitr::is_latex_output(), message=FALSE}
#| fig.cap: "Six conditional simulations for NO$_2$ values"
#| code-fold: true
library(viridis)
g <- ggplot() + coord_equal() +
	scale_fill_viridis() +
    theme_void() +
	scale_x_discrete(expand=c(0,0)) +
	scale_y_discrete(expand=c(0,0))
g + geom_stars(data = s[,,,1:6]) + facet_wrap(~sample)
```

Alternative methods for conditional simulation have recently been
added to `gstat`, and include `krigeSimCE` implementing the circular
embedding method [@JSSv055i09], and `krigeSTSimTB` implementing
the turning bands method [@schlather]. These are of particular
of interest for larger datasets or conditional simulations of
spatiotemporal data.

## Trend models

\index{kriging!trend model}

Kriging and conditional simulation, as used so far in this
chapter, assume that all spatial variability is a random process,
characterised by a spatial covariance model. In case we have other
variables that are meaningfully correlated with the target variable,
we can use them in a linear regression model for the trend,
$$
Z(s) = \sum_{j=0}^p \beta_j X_j(s) + e(s)
$$
with $X_0(s) = 1$ and $\beta_0$ an intercept, but with the other
$\beta_j$ regression coefficients. Adding variables typically
reduces both the spatial correlation in the residual $e(s)$, as
well as its variance, and leads to more accurate predictions and
more similar conditional simulations. As an example, we will use
population density to partly explain variation in NO$_2$.

### A population grid

\index{population density grid}

As a potential predictor for NO$_2$ in the air, we use population
density. NO$_2$ is mostly caused by traffic, and traffic is more intense
in densely populated areas.
Population density is obtained from the [2011 census](https://www.zensus2011.de/DE/Home/Aktuelles/DemografischeGrunddaten.html) and is downloaded as a csv file with the number of inhabitants per 100 m $\times$ 100 m grid cell. We can aggregate these data to the target grid cells by summing the inhabitants:
```{r vroom1, message=FALSE, eval=!CI}
v <- vroom::vroom("aq/pop/Zensus_Bevoelkerung_100m-Gitter.csv")
v |> filter(Einwohner > 0) |> 
	select(-Gitter_ID_100m) |>
	st_as_sf(coords = c("x_mp_100m", "y_mp_100m"), crs = 3035) |>
	st_transform(st_crs(grd)) -> b
a <- aggregate(b, st_as_sf(grd, na.rm = FALSE), sum)
```
```{r vroom0, echo=FALSE, eval=CI}
load("data/ch12.rda") # instead of reading v & processing b and a, load object a
```

Now we have the population counts per grid cell in `a`. To get to
population density, we need to find the area of each cell; for cells
crossing the country border, this will be less than 10 $\times$ 10 km:
```{r, message=FALSE}
grd$ID <- 1:prod(dim(grd)) # to identify grid cells
ii <- st_intersects(grd["ID"],
  st_cast(st_union(de), "MULTILINESTRING"), as_points = FALSE)
grd_sf <- st_as_sf(grd["ID"], na.rm = FALSE)[lengths(ii) > 0,]
st_agr(grd_sf) = "identity"
iii <- st_intersection(grd_sf, st_union(de))
grd$area <- st_area(grd)[[1]] + 
    units::set_units(grd$values, m^2)
grd$area[iii$ID] <- st_area(iii)
```
Instead of doing the two-stage procedure above, first finding cells that
have a border crossing it then computing its area, we could also directly
use `st_intersection` on all cells, but that takes considerably longer.
From the counts and areas we can compute densities (@fig-popdens)
and verify totals 
```{r}
grd$pop_dens <- a$Einwohner / grd$area
sum(grd$pop_dens * grd$area, na.rm = TRUE) # verify
sum(b$Einwohner)
```
which indicates strong agreement. Using `st_interpolate_aw` would
have given an exact match.
```{r fig-popdens, echo = !knitr::is_latex_output()}
#| fig.cap: "Population density for 100 m $\\times$ 100 m grid cells"
#| code-fold: true
g + geom_stars(data = grd, aes(fill = sqrt(pop_dens), x = x, y = y))
```
We need to divide the number of inhabitants by the number of 100
m $\times$ 100 m grid cells contributing to it, in order to convert population
counts into population density.

\index{st\_extract}
To obtain population density values at monitoring network stations,
we can use `st_extract`:
```{r}
grd |>
  select("pop_dens") |>
  st_extract(no2.sf) |>
  pull("pop_dens") |> 
  mutate(no2.sf, pop_dens = _) -> no2.sf
```

We can then investigate the linear relationship between NO$_2$ and
population density at monitoring station locations:
```{r}
summary(lm(NO2~sqrt(pop_dens), no2.sf))
```
for which the corresponding scatterplot is shown in @fig-no2scat.
```{r fig-no2scat, echo=!knitr::is_latex_output()}
#| fig.cap: "Scatter plot of 2017 annual mean NO$_2$ concentration against population density, for rural background air quality stations"
#| code-fold: true
plot(NO2 ~ sqrt(pop_dens), no2.sf)
abline(lm(NO2 ~ sqrt(pop_dens), no2.sf))
```

\index{variogram!residual}
Prediction under this new model involves first modelling a residual
variogram 
```{r}
no2.sf <- no2.sf[!is.na(no2.sf$pop_dens),]
vr <- variogram(NO2~sqrt(pop_dens), no2.sf)
vr.m <- fit.variogram(vr, vgm(1, "Exp", 50000, 1))
```
```{r fig-predictusingpopulationdensity,echo=!knitr::is_latex_output()}
#| fig.cap: "Residual variogram after subtracting population density trend"
#| code-fold: true
plot(vr, vr.m, plot.numbers = TRUE)
```
which is shown in @fig-predictusingpopulationdensity.
Subsequently, kriging prediction (@fig-residualkriging) is done by 
```{r}
kr <- krige(NO2 ~ sqrt(pop_dens), no2.sf, 
			grd["pop_dens"], vr.m)
```

```{r fig-residualkriging, echo=!knitr::is_latex_output(), message=FALSE}
#| fig.cap: "Kriging NO$_2$ values using population density as a trend variable"
#| code-fold: true
k$kr1 <- k$var1.pred
k$kr2 <- kr$var1.pred
st_redimension(k[c("kr1", "kr2")], 
	along = list(what = c("kriging", "residual kriging"))) |>
	setNames("NO2") -> km
g + geom_stars(data = km, aes(fill = NO2, x = x, y = y)) + 
	geom_sf(data = st_cast(de, "MULTILINESTRING")) + 
	geom_sf(data = no2.sf) + facet_wrap(~what) +
	coord_sf(lims_method = "geometry_bbox")
```
where, critically, the `pop_dens` values are now available for
prediction locations in the `newdata` object `grd`.  

Compared to (ordinary) kriging We see some clear differences:
the map using population density in the trend follows the extremes of
the population density rather than those of the measurement stations,
and has a value range that extends that of ordinary kriging. It
should be taken with a large grain of salt however, since the
stations used were filtered for the category "rural background",
indicating that they only represent conditions of lower populations
density. The scatter-plot of @fig-no2scat  reveals that the the
population density at the locations of stations is much more limited
than that in the population density map, and hence the right-hand
side map is based on strongly extrapolating the relationship shown
in @fig-no2scat.

## Exercises 

1. Create a plot like the one in @fig-residualkriging 
that has the inverse distance interpolated map of 
@fig-idw added on the left side.
2. Create a scatter-plot of the map values of the idw and kriging
map, and a scatter-plot of map values of idw and residual kriging.
3. Carry out a _block kriging_, predicting block averages for blocks
centred over grid cells, by setting the `block` argument in
`krige()`, and do this for block sizes of 10 km (grid cell size),
50 km, and 200 km. Compare the resulting maps of estimates for these
three block sizes with those obtained by point kriging, and
do the same thing for all associated kriging standard errors.
4. Based on the residual kriging results obtained above, compute
maps of the lower and upper boundary of a 95% confidence interval,
when assuming that the kriging error is normally distributed,
and show them in a plot with a single (joint) legend.
5. Compute and show the map with the probabilities that NO$_2$ point
values exceed the level of 15 ppm, assuming normally distributed
kriging errors.

```{r echo=FALSE, eval=!CI}
rm(v, b)
save(list = ls(), file = "ch12.RData")
```

{{< include ga.qmd >}}
