# Interpolation and Geostatistics

Geostatistics is concerned with the modelling, prediction and
simulation of spatially continuous phenomena. The typical problem
is a missing value problem: we observe a property of a phenomenon
$Z(s)$ at a limited number of sample locations $s_i, i = 1,...,n$, 
and are interested in the property value at all locations $s_0$, 
so we have to predict it for unobserved locations.

In this chapter we will show simple approaches for handling
geostatistical data, will demonstrate simple interpolation methods,
explore modelling spatial correlation, spatial prediction and
simulation. We will use package `gstat` [@R-gstat, @gstatcg],
which offers a fairly wide palette of models and options for
geostatistical analysis.

## Preparing the air quality dataset

The dataset we work with is an air quality dataset obtained from
the European Environmental Agency (EEA). European member states report
air quality measurements to this Agency. So-called _validated_
data are quality controlled by member states, and are reported on a
yearly basis. They form the basis for policy compliancy evaluations.

The EEA's [air quality
e-reporting](https://www.eea.europa.eu/data-and-maps/data/aqereporting-8)
website gives access to the data reported by European member states.
We decided to download hourly (time series) data, which is the data primarily measured.
A web form helps convert simple selection criteria into an http request. The following URL
```
https://fme.discomap.eea.europa.eu/fmedatastreaming/AirQualityDownload/AQData_Extract.fmw?CountryCode=DE&CityName=&Pollutant=8&Year_from=2017&Year_to=2017&Station=&Samplingpoint=&Source=E1a&Output=TEXT&UpdateDate=
```
was created to select all validated (`Source=E1a`) $NO_2$
(`Pollutant=8`) data for 2017 (`Year_from`, `Year_to`) from Germany
(`CountryCode=DE`).  It returns a text file with a set of URLs to CSV
files, each containing the hourly values for the whole period for a
single measurement station. These files were downloaded and converted
to the right encoding using the `dos2unix` command line utility.

In the following, we will read all the files into a list,
```{r}
files = list.files("aq", pattern = "*.csv", full.names = TRUE)
r = lapply(files[-1], function(f) read.csv(f))
```
then convert the time variable into a `POSIXct` variable, and time order them
```{r}
r = lapply(r, function(f) {
		f$t = as.POSIXct(f$DatetimeBegin) 
		f[order(f$t), ] 
	}
) 
```
and we deselect smaller datasets that do not contain hourly data:
```{r}
r = r[sapply(r, nrow) > 1000]
names(r) =  sapply(r, function(f) unique(f$AirQualityStationEoICode))
length(r) == length(unique(names(r)))
```
and then combine all files using `xts::cbind`, so that they are matched based on time:
```{r}
library(xts)
r = lapply(r, function(f) xts(f$Concentration, f$t))
aq = do.call(cbind, r)
```
We can now select those stations for which we have 75\% of the hourly values measured, i.e. drop those with more than 25\% hourly values missing:
```{r}
# remove stations with more than 75% missing values:
sel = apply(aq, 2, function(x) sum(is.na(x)) < 0.75 * 365 * 24)
aqsel = aq[, sel] # stations are in columns
```

Next, the [station metadata](http://ftp.eea.europa.eu/www/AirBase_v8/AirBase_v8_statistics.zip) was read 
and filtered for rural background stations in Germany by
```{r}
library(tidyverse)
read.csv("aq/AirBase_v8_stations.csv", sep = "\t", stringsAsFactors = FALSE) %>% 
	as_tibble  %>% 
	filter(country_iso_code == "DE", station_type_of_area == "rural", 
				 type_of_station == "Background") -> a2
```
These stations contain coordinates, and an `sf` object with (static) station metadata is created by
```{r}
library(sf)
library(stars)
a2.sf = st_as_sf(a2, coords = c("station_longitude_deg", "station_latitude_deg"), crs = 4326)
```

We now subset the air quality data to stations that are of type rural background:
```{r}
sel =  colnames(aqsel) %in% a2$station_european_code
aqsel = aqsel[, sel]
```

We can compute station means, and join these to stations locations by
```{r}
tb = tibble(NO2 = apply(aqsel, 2, mean, na.rm = TRUE), station_european_code = colnames(aqsel))
crs = 32632
right_join(a2.sf, tb) %>% st_transform(crs) -> no2.sf 
# load German boundaries
data(air, package = "spacetime")
de <- st_transform(st_as_sf(DE_NUTS1), crs)
ggplot() + geom_sf(data = de) +  geom_sf(data = no2.sf, mapping = aes(col = NO2))
```

## Sample variogram 

In order to make spatial predictions using geostatistical methods, we first need to identify a model for the mean and for the spatial correlation. In the simplest model, $Z(s) = m + e(s)$, the mean is an unknown constant $m$, and in this case the spatial correlation can be modelled using the variogram, $\gamma(h) = 0.5 E (Z(s)-Z(s+h))^2$. For processes with a finite variance $C(0)$, the variogram is related to the covariogram or covariance function through $\gamma(h) = C(0)-C(h)$.

The sample variogram is obtained by computing estimates of $\gamma(h)$ for distance intervals, $h_i = [h_{i,0},h_{i,1}]$:
$$
\hat{\gamma}(h_i) = \frac{1}{2N(h_i)}\sum_{j=1}^{N(h_i)}(z(s_i)-z(s_i+h'))^2, \ \ h_{i,0} \le h' < h_{i,1}
$$

with $N(h_i)$ the number of sample pairs available for distance interval $h_i$.
Function `gstat::variogram` computes sample variograms.
```{r}
library(gstat)
v = variogram(NO2~1, no2.sf)
plot(v, plot.numbers = TRUE)
```
This chooses default maximum distance (`cutoff`: one third of the length of the bounding box diagonal) and (constant) interval widths (`width`: cutoff divided by 15). These defaults can be changed, e.g. by
```{r}
library(gstat)
v0 = variogram(NO2~1, no2.sf, cutoff = 100000, width = 10000)
plot(v0, plot.numbers = TRUE)
```
Note that the formula `NO2~1` is used to select the variable of interest from the data file (`NO2`), and to specify the mean model: `~1` refers to an intercept-only (unknown, constant mean) model.

## Fitting variogram models

In order to progress toward spatial predictions, we need a variogram _model_ $\gamma(h)$ for (potentially) all distances $h$, rather than the set of estimates derived above: in case we would for instance connect these estimates with straight lines, or assume they reflect constant values over their respective distance intervals, this would lead to statisical models with non-positive covariance matrices, which is a complicated way of expressing that you are in a lot of trouble.

To avoid these troubles we fit parametric models $\gamma(h)$ to the estimates $\hat{\gamma}(h_i)$, where we take $h_i$ as the mean value of all the $h'$ values involved in estimating $\hat{\gamma}(h_i)$. For this, when we fit a model like the exponential variogram,
```{r}
v.m = fit.variogram(v, vgm(1, "Exp", 50000, 1))
plot(v, v.m, plot.numbers = TRUE)
```
the fitting is done by minimizing $\sum_{i=1}^{n}w_i(\gamma(h_i)-\hat{\gamma}(h_i))^2$, with $w_i$ by default equal to $N(h_i)/h^2$, other fitting schemes are available through argument `fit.method`.

## Kriging interpolation

Kriging involves the prediction of $Z(s_0)$ at arbitrary locations $s_0$. Typically, when
we interpolate a variable, we do that on points on a regular grid covering the target area.
When we take the grid points such that they align with the points of our graphics device (screen), 
then nobody will notice as long as the graphic device is not resized.

We first create a `stars` object with a raster covering the target area, and NA's outside it:
```{r}
# build a grid over Germany:
bb = st_bbox(de)
dx = seq(bb[1], bb[3], 10000)
dy = seq(bb[4], bb[2], -10000) # decreases!
st_as_stars(matrix(0, length(dx), length(dy))) %>%
  st_set_dimensions(1, dx) %>%
  st_set_dimensions(2, dy) %>%
  st_set_dimensions(names = c("x", "y")) %>%
  st_set_crs(crs) -> grd
i = st_intersects(grd, de)
grd[[1]][lengths(i)==0] = NA
#grd %>% st_dimensions()
grd
```

Then, we can krige by using `gstat::krige`, with the model for the trend, the data,
the prediction grid, and the variogram model:
```{r}
k = krige(NO2~1, no2.sf, grd, v.m)
ggplot() + geom_stars(data = k, aes(fill = var1.pred, x = x, y = y)) + 
	geom_sf(data = st_cast(de, "MULTILINESTRING")) + 
	geom_sf(data = no2.sf)
```

## Areal means: block kriging
```{r}
a = aggregate(no2.sf["NO2"], de, mean)
b = krige(NO2~1, no2.sf, de, v.m)
b$sample = a$NO2
b$kriging = b$var1.pred
b %>% select(sample, kriging) %>% gather(var, NO2, -geometry) -> b2
ggplot() + geom_sf(data = b2, mapping = aes(fill = NO2)) + facet_wrap(~var) +
	 scale_fill_gradientn(colors = sf.colors(20))
```

```{r}
SE = function(x) sqrt(var(x)/length(x))
a = aggregate(no2.sf["NO2"], de, SE)
b$sample = a$NO2
b$kriging = sqrt(b$var1.var)
b %>% select(sample, kriging) %>% gather(var, NO2, -geometry) -> b2
ggplot() + geom_sf(data = b2, mapping = aes(fill = NO2)) + facet_wrap(~var) +
	 scale_fill_gradientn(colors = sf.colors(20))
```

## Conditional and unconditional simulation

```{r}
library(viridis)
s = krige(NO2~1, no2.sf, grd, v.m, nmax = 30, nsim = 10)
g = ggplot() + coord_equal() +
	scale_fill_viridis() +
    theme_void() +
	scale_x_discrete(expand=c(0,0)) +
	scale_y_discrete(expand=c(0,0))
g + geom_stars(data = s[,,,1:6]) + facet_wrap(~sample)
```

## Trend models

Kriging and conditional simulation, as used so far in this
chapter, assume that all spatial variability is a random process,
characterized by a spatial covariance model. In case we have other
variables that are meaningfully correlated with the target variable,
we can use them in a linear regression model for the trend,
$$
Z(s) = \sum_{j=0}^p \beta_j X_p(s) + e(s)
$$
with $X_0(s) = 1$ and $\beta_0$ an intercept, but with the other
$\beta_j$ regression coefficients. This typically reduces both the
spatial correlation in the residual $e(s)$, as well as its variance,
and leads to more accurate predictions and more similar conditional
simulations.

### A population grid

As a potential predictor for NO2 in the air, we use population
density. NO2 is mostly caused by traffic, and traffic is stronger
in more densely populated areas.
Population density is obtained from the [2011 census](https://www.zensus2011.de/DE/Home/Aktuelles/DemografischeGrunddaten.html), and is downloaded as a csv file with the number of inhabitants per 100 m grid cell. We can aggregate these data to the target grid cells by summing the inhabitants:
```{r}
library(vroom)
v = vroom("aq/pop/Zensus_Bevoelkerung_100m-Gitter.csv")
v %>% filter(Einwohner > 0) %>% 
	select(-Gitter_ID_100m) %>%
	st_as_sf(coords = c("x_mp_100m", "y_mp_100m"), crs = 3035) %>%
	st_transform(st_crs(grd)) -> b
a = aggregate(b, st_as_sf(grd, na.rm = FALSE), sum)
```
Now we have the population counts per grid cell in `a`. To get to
population density, we need to find the area of each cell; for cells
crossing the country border, this will be less than 10 x 10 km:
```{r}
grd$ID = 1:prod(dim(grd)) # so we can find out later which grid cell we have
ii = st_intersects(grd["ID"], st_cast(st_union(de), "MULTILINESTRING"))
grd_sf = st_as_sf(grd["ID"], na.rm = FALSE)[lengths(ii) > 0,]
iii = st_intersection(grd_sf, st_union(de))
grd$area = st_area(grd)[[1]] + units::set_units(grd$A1, m^2) # NA's
grd$area[iii$ID] = st_area(iii)
```
Instead of doing the two-stage procedure above: first finding cells that
have a border crossing it, then computing its area, we could also directly
use `st_intersection` on all cells, but that takes considerably longer.
From the counts and areas we can compute densities, and verify totals:
```{r}
grd$pop_dens = a$Einwohner / grd$area
sum(grd$pop_dens * grd$area, na.rm = TRUE) # verify
sum(b$Einwohner)
g + geom_stars(data = grd, aes(fill = sqrt(pop_dens), x = x, y = y))
```
We need to divide the number of inhabitants by the number of 100
m grid points contributing to it, in order to convert population
counts into population density.

To obtain population density values at monitoring network stations,
we can use
```{r}
(a = aggregate(grd["pop_dens"], no2.sf, mean))
no2.sf$pop_dens = st_as_sf(a)[[1]]
summary(lm(NO2~sqrt(pop_dens), no2.sf))
```
and the corresponding scatterplot is shown in \@ref(fig:no2scat).
```{r no2scat, out.width = '50%', fig.cap="Scatter plot of 2017 annual mean NO2 concentration against population density, for rural background air quality stations", eval=TRUE, fig=TRUE, echo=FALSE}
plot(NO2~sqrt(pop_dens), no2.sf)
abline(lm(NO2~sqrt(pop_dens), no2.sf))
```

Prediction under this new model involves first modelling a residual
variogram, by
```{r}
no2.sf = no2.sf[!is.na(no2.sf$pop_dens),]
vr = variogram(NO2~sqrt(pop_dens), no2.sf)
vr.m = fit.variogram(vr, vgm(1, "Exp", 50000, 1))
plot(vr, vr.m, plot.numbers = TRUE)
```
and subsequently, kriging prediction is done by
```{r}
kr = krige(NO2~sqrt(pop_dens), no2.sf, grd["pop_dens"], vr.m)
k$kr1 = k$var1.pred
k$kr2 = kr$var1.pred
st_redimension(k[c("kr1", "kr2")], 
	along = list(what = c("kriging", "residual kriging"))) %>%
	setNames("NO2") -> km
g + geom_stars(data = km, aes(fill = NO2, x = x, y = y)) + 
	geom_sf(data = st_cast(de, "MULTILINESTRING")) + 
	geom_sf(data = no2.sf) + facet_wrap(~what)
```
where, critically, the `pop_dens` values are now available for
prediction locations in object `grd`.  We see some clear differences:
the map with population density in the trend follows the extremes of
the population density rather than those of the measurement stations,
and has a range that extends that of the former. It should be taken
with a large grain of salt however, since the stations used were
filtered for the category "rural background", indicating that they
represent conditions of lower populations density. The scatter plot
of Figure \@ref(fig:no2scat) reveals that the the population density
at the locations of stations is much more limited than that in the
population density map, and hence the right-hand side map is based on
strongly extrapolating the relationship shown in \@ref(fig:no2scat).

## Multivariable geostatistics

Multivariable geostatics involves the _joint_ modelling, prediction
and simulation of multiple variables, 
$$Z_1(s) = X_1 \beta_1 + e_1(s)$$
$$...$$
$$Z_n(s) = X_n \beta_n + e_n(s).$$
In addition to having observations, trend models, and variograms
for each variable, the _cross_ variogram for each pair of residual
variables, describing the covariance of $e_i(s), e_j(s+h)$,
is required. If this cross covariance is non-zero, knowledge of
$e_j(s+h)$ may help predict (or simulate) $e_i(s)$.  This is
especially true if $Z_j(s)$ is more densely sample than $Z_i(s)$.
Prediction and simulation under this model are called cokriging
and cosimulation. Examples using gstat are found when running the
demo scripts
```{r eval=FALSE}
library(gstat)
demo(cokriging)
demo(cosimulation)
```

In case the different variables considered are observed at
the same set of locations, for instance different air quality
parameters, then the statistical _gain_ of using cokriging
as opposed to direct (univariable) kriging is often modest,
when not negligible. A gain may however be that the prediction
is truly multivariable: in addition to the prediction vector
$\hat{Z(s_0)}=(\hat{Z}_1(s_0),...,\hat{Z}_n(s_0))$ we get the full
covariance matrix of the prediction error [@ver1993multivariable].
This means for instance that if we are interested in some
linear combination of $\hat{Z}(s_0)$, such as $\hat{Z}_2(s_0)
- \hat{Z}_1(s_0)$, that we can get the standard error of that
combination because we have the correlations between the prediction
errors.

Although sets of direct and cross variograms can be computed
and fitted automatically, multivariable geostatistical modelling
becomes quickly hard to manage when the number of variables gets
large, because the number of direct and cross variograms required
is $n(n+1)/2$.

In case different variables refer to the same variable take at
different time steps, one could use a multivariable (cokriging)
prediction approach, but this would not allow for e.g. interpolation
between two time steps. For this, and for handling the case of
having data observed at many time instances, one can also model
its variation as a function of continuous space _and_ time, as of
$Z(s,t)$, which we will do in the next section.

## Spatiotemporal interpolation

[@RJ-2016-014]

- predict missing values -> multivariate arima would have been enough/better?
- predict a sequence of hours
- smooth in between, e.g. by 10-min interpolations


