# Older R Spatial Packages {#older}

## Retiring `rgdal` and `rgeos` {#retire}

R users who have been around a bit longer, in particular before
packages like `sf` and `stars` were developed, may be more familiar
with older packages like `maptools`, `sp`, `rgeos`, and `rgdal`. A
fair question is whether they should migrate existing code and/or
existing R packages depending on these packages.  The answer is: yes.

Unless someone steps up to volunteer maintaining packages `maptools`,
`rgdal` and `rgeos`, the plan is to retire packages by the end of
2023. Retirement means that maintenance will halt, and that as a
consequence the packages will sooner or later disapper from CRAN.
One reason for retirement is that their maintainer has retired,
another that their role has been superseded by the newer packages.
We hold it not for very likely that a new maintainer will take over,
in part because much of the code of these packages has over a few
decades gradually evolved along with developments in the GEOS,
GDAL and PROJ libraries, and now contains numerous constructs that
are no longer necessary and make it hard to read.

Before `rgeos` and `rgdal` retire, existing ties that package `sp`
has to `rgdal` and `rgeos` can and will be replaced by ties to
package `sf`. This only involves validation of coordinate reference
system identifiers, and checking whether rings are holes or exterior
rings. Theoretically one could replace `rgdal` and `rgeos` with
packages that would call into `sf` for their ties to the GEOS,
GDAL and PROJ libraries but that would involve a major effort.

## links and differences between sf and sp

There are a number of differences between `sf` and `sp`. The most
notable is that `sp` classes are formal, S4 classes where `sf` uses
the (more) informal S3 class hierarchy. `sf` objects derive from
data.frames, or from tibbles, and as such can be used easier with
much of the other infrastructure of R and e.g. with the tidyverse
package family. `sf` objects keep geometry in a list-column, meaning
that a geometry is _always_ a list element. Package `sp` used data
structures much less strictly, and for instance all coordinates of
`SpatiaPoints` or `SpatialPixels` are kept in matrices, which is
much more performant for certain problems but is not possible with 
a list-column. Conversion from an `sf` object `x` to its `sp`
equivalent is done by
```{r eval=FALSE}
library(sp)
y = as(x, "Spatial")
```
and the conversion the other way around is done by
```{r eval=FALSE}
x0 = st_as_sf(y)
```

There are some limitations to conversions like this:

* `sp` does not distinguish between `LINESTRING` and
`MULTILINESTRING` geometries, or between `POLYGON` or `MULTIPOLYGON`,
so e.g. a `LINESTRING` will after conversion to `sp` come back as a
`MULTILINESTRING`
* `sp` does have no representation for `GEOMETRYCOLLECTION`
geometries, or `sf` objects with geometries _not_ in the "big seven"
(section \@ref(seven))
* `sf` or `sfc` objects of geometry type `GEOMETRY`, with mixed
geometry types, cannot be converted into `sp` objects
* attribute-geometry relationship attributes get lost when converting
  to `sp`
* `sf` objects with more than one geometry list-column will, when
converting to `sp`, loose their secondary list-column(s).


## migration code and packages

The wiki page of the GitHub site for sf, found at

    https://github.com/r-spatial/sf/wiki/Migrating

contains a list of methods and functions in `rgeos`, `rgdal` and
`sp` and the corresponding `sf` method or function. This may help
converting existing code or packages.

A simple approach to migrate code is when only `rgdal::readOGR` is
used to read `file`. As an alternative, one might use
```{r, eval = FALSE}
x = as(sf::read_sf("file"), "Spatial")
```
however possible arguments to `readOGR`, when used, would need more
care.

An effort by us is underway to convert all code of our earlier book
"Applied Spatial Data Analysis with R" (with Virgilio Gomez-Rubio,
@asdar) to run entirely without `rgdal`, `rgeos` and `maptools` and
where possible without `sp`. The scripts are found at
https://github.com/rsbivand/sf_asdar2ed .

## Package raster and terra

Package `raster` has been a workhorse package for analysing raster
data with R since 2010, and has since then grown into a package for
"Geographic Data Analysis and Modeling" [@R-raster], indicating that
it is used for all kinds of spatial data. The `raster` package uses
`sp` objects for vector data, and `rgdal` to read and write data to
formats served by the GDAL library. Its successor package `terra`,
for "Spatial Data Analysis" [@R-terra], "is very similar to the
'raster' package; but [...] can do more, is easier to use, and
[...]  is faster". The `terra` package comes with its own classes
for vector data, but accepts many `sf` objects, with similar
restrictions as listed above for conversion to `sp`.  Package `terra`
has its own direct links to GDAL, GEOS and PROJ so no longer needs
other packages for that. 

Raster maps, or stacks of them from package `raster` or `terra`
can be converted to `stars` objects using `st_as_stars()`. Package
`sf` contains an `st_as_sf()` method for `SpatVector` objects
from package `terra`.  Migration from `raster` to `terra` may
become more important once `rgdal` is no longer easily installable
(section \@ref(retire)).

The online book "Spatial Data Science with R", written by Robert
Hijmans and found at https://rspatial.org/terra details the `terra`
approach to spatial data analysis. Package `sf` and `stars` and
several other r-spatial packages discussed in this book reside on the
`r-spatial` GitHub organisation (note the hyphen between `r` and
`spatial`, which is absent on Hijmans' organisation), which has a
blog site, with links to this book, found  at https://r-spatial.org/ . 

Packages `sf` and `stars` on one hand and `terra` on the other
have many goals in common, but try to reach them in slightly
different ways, emphasizing different aspects of data analysis,
software engineering, and community management. Although this may
confuse some users, we believe that these differences enrich the
R package ecosystem, are beneficial to users, encourage diversity
and choice, and hopefully work as an encouragement for others to
continue trying out new ideas when using R for spatial data problems,
and to help carrying the R spatial flag.
