# R basics {#sec-rbasics}

This chapter provides some minimal R basics that may make it easier
to read this book. A more comprehensive book on R basics is given in
[@advr], chapter 2.

## Pipes {#sec-pipes}
\index{pipes}

The `|>` (pipe) symbols should be read as _then_: we read
```{r eval=FALSE}
a |> b() |> c() |> d(n = 10)
```
as _with `a` do `b`, then `c`, then `d` with `n` being 10_, and that is just alternative syntax for
```{r eval=FALSE}
d(c(b(a)), n = 10)
```
or 
```{r eval=FALSE}
tmp1 <- b(a)
tmp2 <- c(tmp1)
tmp3 <- d(tmp2, n = 10)
```
To many, the pipe-form is easier to read because execution order
follows reading order,  from left to right.  Like nested function
calls, it avoids the need to choose names for intermediate results,
as with nested function calls it is hard to debug intermediate
results that diverge from our expectations. Note that the
intermediate results do exist in memory, so neither form saves memory
allocation. The `|>` native pipe that appeared in R 4.1.0 as used
in this book, can be safely substituted by the `%>%` pipe of package
**magrittr**.

## Data structures {#sec-datastructures}
\index{data structures}

As pointed out by [@extending], _everything that exists in R is an
object_. This includes objects that make things happen, such as
language objects or functions, but also the more basic "things",
such as data objects. Some basic R data structures will now be
discussed.

### Homogeneous vectors
\index{vectors}

Data objects contain data, and possibly metadata. Data is always
in the form of a vector, which can have different type. We can
find the type by `typeof`, and vector length by `length`. Vectors
are created by `c`, which combines individual elements:
```{r}
typeof(1:10)
length(1:10)
typeof(1.0)
length(1.0)
typeof(c("foo", "bar"))
length(c("foo", "bar"))
typeof(c(TRUE, FALSE))
```
Vectors of this kind can only have a single type.

Note that vectors can have zero length, e.g. in, 
```{r}
i <- integer(0)
typeof(i)
i
length(i)
```

We can retrieve (or in assignments: replace) elements in a vector
using `[` or `[[`:
```{r}
a <- c(1,2,3)
a[2]
a[[2]]
a[2:3]
a[2:3] <- c(5,6)
a
a[[3]] <- 10
a
```
where the difference is that `[` can operate on an index _range_
(or multiple indexes), and `[[` operates on a single vector value.

### Heterogeneous vectors: `list`
\index{list}

An additional vector type is the `list`, which can combine any types in
its elements:
```{r}
l <- list(3, TRUE, "foo")
typeof(l)
length(l)
```

For lists, there is a further distinction between `[` and `[[`: the single
`[` returns always a list, and `[[` returns the _contents_ of a list element:
```{r}
l[1]
l[[1]]
```
For replacement, one case use `[` when providing a list, and `[[` when providing
a new value:
```{r}
l[1:2] <- list(4, FALSE)
l
l[[3]] <- "bar"
l
```
In case list elements are _named_, as in 
```{r}
l <- list(first = 3, second = TRUE, third = "foo")
l
```
we can use names as in `l[["second"]]` and this can be
abbreviated to
```{r}
l$second
l$second <- FALSE
l
```
This is convenient, but also requires name look-up in the names
attribute (see below).

### NULL and removing list elements
\index{NULL}

`NULL` is the null value in R; it is special in the sense that it doesn't work
in simple comparisons:
```{r}
3 == NULL # not FALSE!
NULL == NULL # not even TRUE!
```
but has to be treated specially, using `is.null`:
```{r}
is.null(NULL)
```

When we want to remove one or more list elements, we can do so by creating
a new list that does not contain the elements that needed removal, as in
```{r}
l <- l[c(1,3)] # remove second, implicitly
l
```
but we can also assign `NULL` to the element we want to eliminate:
```{r}
l$second <- NULL
l
```


### Attributes
\index{attributes!of R objects}

We can glue arbitrary metadata objects to data objects, as in
```{r}
a <- 1:3
attr(a, "some_meta_data") = "foo"
a
```
and this can be retrieved, or replaced by
```{r}
attr(a, "some_meta_data")
attr(a, "some_meta_data") <- "bar"
attr(a, "some_meta_data")
```
In essence, the attribute of an object is a named list, and we can
get or set the complete list by
```{r}
attributes(a)
attributes(a) = list(some_meta_data = "foo")
attributes(a)
```

A number of attributes are treated specially by R, see e.g. `?attributes`.

#### Object class and class attribute
\index{class}
\index[function]{class}

Every object in R "has a class", meaning that `class(obj)` returns
a character vector with the class of `obj`. Some objects have
an _implicit_ class, e.g. vectors

```{r}
class(1:3)
class(c(TRUE, FALSE))
class(c("TRUE", "FALSE"))
```
but we can also set the class explicit, either by using `attr` or by
using `class` in the left-hand side of an expression:
```{r}
a <- 1:3
class(a) <- "foo"
a
class(a)
attributes(a)
```
in which case the newly set class overrides the earlier implicit class. This way, 
we can add methods for class `foo`, e.g. by
```{r}
print.foo <- function(x, ...) { 
	print(paste("an object of class foo with length", length(x)))
}
print(a)
```
Providing such methods are generally intended to create more usable
software, but at the same time they may make the objects more opaque. It is
sometimes useful to see what an object "is made of" by printing it after the
class attribute is removed, as in
```{r}
unclass(a)
```
As a more elaborate example, consider the case where a polygon is made using
package **sf**:
```{r}
library(sf) |> suppressPackageStartupMessages()
p <- st_polygon(list(rbind(c(0,0), c(1,0), c(1,1), c(0,0))))
p
```
which prints the well-known-text form; to understand what the data structure is
like, we can use
```{r}
unclass(p)
```

#### The `dim` attribute
\index{dim}
\index[function]{dim}

The `dim` attribute sets the matrix or array dimensions:
```{r}
a <- 1:8
class(a)
attr(a, "dim") <- c(2,4) # or: dim(a) = c(2,4)
class(a)
a
attr(a, "dim") <- c(2,2,2) # or: dim(a) = c(2,2,2)
class(a)
a
```

### The names attributes

Named vectors carry their names in a `names` attribute. We saw examples
for lists above, an example for a numeric vector is:
```{r}
a <- c(first = 3, second = 4, last = 5)
a["second"]
attributes(a)
```
More name attributes are e.g. `dimnames` of matrices or arrays,
which not only names dimensions, but also the labels associated
with each of the dimensions:
```{r}
a <- matrix(1:4, 2, 2)
dimnames(a) <- list(rows = c("row1", "row2"),
				    cols = c("col1", "col2"))
a
attributes(a)
```
Data.frame objects have rows and columns, and each have names:
```{r}
df <- data.frame(a = 1:3, b = c(TRUE, FALSE, TRUE))
attributes(df)
```

### Using `structure`
\index[function]{structure}

When programming, the pattern of adding or modifying attributes before returning
an object is extremely common, an example being:
```{r, eval=FALSE}
f <- function(x) {
   a <- create_obj(x) # call some other function
   attributes(a) <- list(class = "foo", meta = 33)
   a
}
```
The last two statements can be contracted in
```{r, eval=FALSE}
f <- function(x) {
   a <- create_obj(x) # call some other function
   structure(a, class = "foo", meta = 33)
}
```
where function `structure` adds, replaces, or (in case of value `NULL`) removes
attributes from the object in its first argument.

## dissecting a `MULTIPOLYGON`

We can use the above examples to dissect an `sf` object with
`MULTIPOLYGON`s into pieces. Suppose we use the `nc` dataset,
```{r}
system.file("gpkg/nc.gpkg", package = "sf") %>%
	read_sf() -> nc
```
we can see from the attributes of `nc`,
```{r}
attributes(nc)
```
that the geometry column is named `geom`. When we take out this column,
```{r}
nc$geom
```
we see an object that has the following attributes
```{r}
attributes(nc$geom)
```
When we take the _contents_ of the fourth list element, we obtain
```{r}
nc$geom[[4]] |> format(width = 60, digits = 5)
```
which is a (classed) list,
```{r}
typeof(nc$geom[[4]])
```
with attributes
```{r}
attributes(nc$geom[[4]])
```
and length
```{r}
length(nc$geom[[4]])
```
The length indicates the number of outer rings: a multi-polygon
can consist of more than one polygon. We see that most counties
only have a single polygon:
```{r}
lengths(nc$geom)
```
A multi-polygon is a list with polygons,
```{r}
typeof(nc$geom[[4]])
```
and the _first_ polygon of the fourth multi-polygon is again a list,
because polygons have an outer ring _possibly_ followed by multiple inner 
rings (holes)
```{r}
typeof(nc$geom[[4]][[1]])
```
we see that it contains only one ring, the exterior ring:
```{r}
length(nc$geom[[4]][[1]])
```
and we can print type, the dimension and the first set of coordinates by
```{r}
typeof(nc$geom[[4]][[1]][[1]])
dim(nc$geom[[4]][[1]][[1]])
head(nc$geom[[4]][[1]][[1]])
```
and we can now for instance change the latitude of the third coordinate by
```{r}
nc$geom[[4]][[1]][[1]][3,2] <- 36.5
```

{{< include ga.qmd >}}
